### 2019-01-10 Trees_making.Rmd
Add files via upload
First lot of files for the exercise on making phylogenetic trees for meta-analyses.

---
title: "Trees"
author: "ML"
date: "10 January 2019"
output: html_document
---

```{r setup, cache = F, echo=FALSE}
knitr::opts_chunk$set(error = TRUE) #allow some execution errors for demonstration purposes
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, warning = FALSE)
```

# EXERCISES - PHYLOGENETIC TREES FOR META-ANALYSIS

Prepared by Malgorzata (Losia) Lagisz, BEES, UNSW, AU, losialagisz@yahoo.com.  
Last updated: 1 June 2016 by Losia Lagisz

Introduction
-------

Ecological and evolutionary meta-analyses usually aggregate data acros multiple species (there are single-species or non-species meta analyses too, but these are  minority).
Since species are related due to shared evolutionary history, effect sizes of closely related species can be more similar to these of less-related specis. 
This kind of non-independence can be represented as a correlation matrix of phylogenetic relatedness and added as a random factor to multilevel meta-analytic models. 
To create a relatedness matrix among species we need a phylogenetic tree for the list of species from our data set.

Currently, the easiest way to build a custom phylogenetic tree usable for meta-analysis is by using an R package called *rotl*.

A few things to note:  
1. We can use a tree with or without branch lengths (divergence measures).  
2. The tree has to be binary (strictly bifurcating = no polytomies, i.e. only two branches are coming from each node).  
3. The tree needs to be rooted (has a node representing the most recent common ancestor of all the species in the tree).   
4. The tree tip labels (representing species) must match **exactly** with the species names in the data set (otherwise MA models will not run with them).  
5. Trees can be also made and used for higher taxonomic levels than species, e.g. genus - level (not covered in this tutorial, but general principles are the same).

Start from installing (as required) and loading R packages.

```{r prepare, message = FALSE}
#install.packages("devtools","fulltxt","ape","metafor")
#update.packages("fulltxt")
#devtools::install_github("ropensci/rotl", dependencies = TRUE, build_vignette=TRUE)
#install.packages("rotl", "treebase", "diagram", "dplyr")

library(readxl)
library(ape)
library(fulltext)
library(metafor)
library(rotl)
library(treebase)
library(diagram)
library(dplyr)
```

********************************************************************************

### EXAMPLE 1 - using *rotl* package

We will try to create a simple tree using 10 commonly used taxa (lab model species) from across the Tree of Life.  
Rotl is an R package (https://peerj.com/preprints/1471/) allowing us to access synthetic phylogenetic tree available at Open Tree of Life database (https://opentree.wikispaces.com/).

**Important**: use Latin names, not common names.   
**Notes**: *rotl* crashes on some species lists. Also, Open Tree of Life database gets constantly updated with new info, so you can get different outputs at different timepoints.

We use a function *tnrs_match_names*, which returns a data frame that lists the Open Tree identifiers potentially matching our species.  
The function also returns other information, including alternative species names, to help users ensure that the taxa matched are the correct ones.

```{r rotl 10species list}
myspecies <- c("Escherichia colli", 
               "Chlamydomonas reinhardtii", 
               "Drosophila melanogaster",
               "Arabidopsis thaliana",
               "Rattus norvegicus",
               "Mus musculus",
               "Cavia porcellus",
               "Xenopus laevis",
               "Saccharomyces cervisae",
               "Danio rerio")
taxa <- tnrs_match_names(names = myspecies)
taxa
```

Note that two of the species names in our list have typos, and are labelled TRUE in a column "approximate_match". Still, they were matched to correct species names.  
If typos are detected, it is worth correcting the names in the original dataset and re-running tree search.

```{r rotl 10species corrected}
myspecies_fixed <- c("Escherichia coli", 
               "Chlamydomonas reinhardtii", 
               "Drosophila melanogaster",
               "Arabidopsis thaliana",
               "Rattus norvegicus",
               "Mus musculus",
               "Cavia porcellus",
               "Xenopus laevis",
               "Saccharomyces cerevisiae",
               "Danio rerio")
taxa <- tnrs_match_names(names = myspecies_fixed)
taxa
```

When we are happy with how the taxon names are matched to the Open Tree identifiers, we can pass them to the
function tol_induced_subtree to retrieve the relationships among these taxa (a trimmed sub-tree from the synthetic phylogenetic tree).  
The tree is returned as an ape::phylo object and it can be manipulated, printed and saved easily using functions from the ape package.

```{r rotl 10species plot}
tree <- tol_induced_subtree(ott_ids = taxa[["ott_id"]], label_format = "name")
plot(tree, cex=.8, label.offset =.1, no.margin = TRUE)
```

Note that the species names have the Open Tree identifiers appended to them and *E.coli* was replaced with "Emrcaott616ott617", which is non-taxon node within Bacteria. Given we dont have any ither bacteria in the dataset, and the position on the tree is correct, thats fine, we just need to rename the node back.

```{r rotl 10species names}
tree$tip.label #check tip labels
ecoli <- tol_node_info(ott_id=474506, include_lineage=TRUE)
ecoli
tax_lineage(ecoli)
tnrs_match_names(names = ("Escherichia")) #we can see what info is available for this genus - not much

#just fix the tip label
tree$tip.label <- gsub("mrcaott616ott617", "Escherichia coli", tree$tip.label) #replace the tip label back with our species name
tree$tip.labe
tree$tip.label <- gsub("_"," ", tree$tip.label) #get rid of the underscores
```

It is time to plot and tidy up the tree (note no branch lengths are included, they can be created later via simulations).

```{r rotl 10species tree}
#see the tree info
tree
#tree$node.label <- NULL #you can delete internal node labels
plot(tree) #plot the tree
```

Do a final check whether the tri is binary and if the tip labels of the tree match exactly our (fixed for typos) list of species.

```{r rotl 10species check}
#check if the tree is really binary 
is.binary.tree(tree) #TRUE (i.e. there are no polytomies)

#check the overlap with our species list
###check the overlap between species name from the data fila nad the tree file
intersect(as.character(tree$tip.label), myspecies_fixed) #10 matching names
setdiff(myspecies_fixed, as.character(tree$tip.label)) #0 mismatches
setdiff(as.character(tree$tip.label), myspecies_fixed) #0 mismatches
```

********************************************************************************

### EXAMPLE 2 - using fixed supertree from a file

Create a phylogenetic tree for 65 bird species used in Rutkowska, J., Dubiec, A. & Nakagawa, S. (2014) All eggs are made equal: meta-analysis of egg sexual size dimorphism in birds. Journal of Evolutionary Biology. 27: 153-160 http://onlinelibrary.wiley.com/doi/10.1111/jeb.12282/abstract

There are a few main supertrees for birds, so they can be used to create a subtree containing only the species from the list.  
Here we will use on of these supertrees "Ericson backbone" downloadable from http://birdtree.org/ (Jetz, W., G. H. Thomas, J. B. Joy, K. Hartmann, and A. O. Mooers. 2012. The global diversity of birds in space and time. Nature 491:444-448.). The website http://birdtree.org/ has an online tool for subsetting their supertrees generating distribution of trees rather than one tree. For simplicity and demonstration puproses we will use R to subset (prune) a single bird supertree down to a smaller list of species.

We have our custom list of bird species stored in a file *bird_list.csv* as a simple string factor and we have downloaded Ercison tree stored in a file *Ericson.tre*, in Newick format with branch lengths included. Load these files and check the species overlap.

```{r list birds load} 
birds <- unique(read.csv("bird_list.csv")$species_name) #load our list of bird species and make the names unique
#str(birds)
birds_stree <- read.tree("Ericson.tre") #load bird supertree
birds_stree #9993 tips = species
str(birds_stree) #has edge (branch) lengths
birds_stree <- collapse.singles(birds_stree) #clean up singularities, just in case

###check the overlap between species name from our list  and the bird supertree
intersect(as.character(birds_stree$tip.label), birds) #51 - all our species matching - they are contained in the tree
setdiff(birds, as.character(birds_stree$tip.label)) #0 mismatches - just making sure we have all our species included in the supertree
```

Prune supertree tree to a list of taxa from our list.

```{r list birds prune} 
pruned_birds_stree <- drop.tip(birds_stree,birds_stree$tip.label[-match(birds, birds_stree$tip.label)]) #pruning
#check the pruned tree
is.binary.tree(pruned_birds_stree) #TRUE
is.ultrametric(pruned_birds_stree) #TRUE

intersect(as.character(pruned_birds_stree$tip.label), birds) #51 - all our species matching - they are contained in the tree
setdiff(birds, as.character(pruned_birds_stree$tip.label)) #0 mismatches - just making sure we have all our species included in the tree

#write.tree(pruned_birds_stree, file = "birds_51sp_tree.tre", append = FALSE, digits = 10, tree.names = FALSE) #save, if needed
```

Plot the tree.

```{r plot tree birds, fig.width=10, fig.height=10, message=FALSE}
plot(pruned_birds_stree, cex=0.8) #plots with branch lengths when available
```

**Notes**: We were lucky that all our species had matches on the supertree. Quite often there are no matches for some of the species names. In such cases, first check the names for typos or unnecessary characters, then try to find synonymous names and check if these are included in the supertree. Finally, a sister species can be used when a species is truly missing. Remember to match the names on the final tree back to the original species list from the dataset.

********************************************************************************

### EXAMPLE 3 - using *rotl* package again

We will now handle a larger set of more closely related taxa, from a real data set.  
The data set we use the dataset *dat.curtis1998* (tudies on the Effects of Elevated CO2 Levels on Woody Plant Mass) included in the *metafor* package.

```{r load Curtis1998 data}
dat <- get(data(dat.curtis1998))
str(dat) # 102 rows, many columns including various data
```

Extract and clean the list of species included. Note that genus and species name are stored in separate columns.

```{r create Curtis1998 species list}
species_Curtis1998 <- paste(dat$genus, dat$species, sep=" ") 
str(species_Curtis1998)
species_Curtis1998 <- unique(tolower(species_Curtis1998)) #make species name unique
length(species_Curtis1998) #37
species_Curtis1998 #note: "populusx euramericana" should be same as "populus euramericana"
species_Curtis1998 <- gsub("populusx euramericana", "populus euramericana", species_Curtis1998) #merge 2 partially overlapping names (typo?)
species_Curtis1998 <- unique(tolower(species_Curtis1998))
length(species_Curtis1998) #36
#capitalise first letters
species_Curtis1998 <- paste(toupper(substr(species_Curtis1998, 1, 1)), substr(species_Curtis1998, 2, nchar(species_Curtis1998)), sep="")
species_Curtis1998
```

We have 36 unique species_Curtis1998 names. We try to run rotl searc with them.

```{r match Curtis1998 species_Curtis1998 list to otl taxa list}
taxa <- tnrs_match_names(names = species_Curtis1998)
dim(taxa) #36
taxa
```

Note that some alternative names were found. We asssume these are correct ones for now.  
Now that the taxon names are matched to the Open Tree identifiers, we can pass them to the function tol_induced_subtree to retrieve the relationships among these taxa.

```{r get Curtis1998 tree}
tree <- tol_induced_subtree(ott_ids= taxa[["ott_id"]], label_format = "name") 
#The following OTT ids were not found: [3915043].
taxa[taxa$ott_id==3915043,] #populus euramericana Populus x canadensis             FALSE 3915043       TRUE HYBRID
#note: Populus × canadensis, known as Canadian poplar or Carolina poplar, is a naturally occurring hybrid of Populus nigra and Populus deltoides
```

We run into a problem with Populus - need to fix this by using an alternative name.

```{r fix Curtis1998 species_Curtis1998 list}
populus <- tnrs_match_names(names = ("populus deltoides"))
populus
#replace populus euramericana with populus deltoides (note 2 versions of the name)
species_Curtis1998 <- gsub("Populus euramericana", "Populus deltoides", species_Curtis1998)
```

```{r get Curtis1998 tree again}
taxa <- tnrs_match_names(names = species_Curtis1998)
head(taxa)
dim(taxa) #36
tree <- tol_induced_subtree(ott_ids= taxa[["ott_id"]], label_format = "name") 
str(tree)
```

Plot the tree:

```{r plot tree Curtis1998, fig.width=10, fig.height=8, echo=TRUE, message=FALSE}
plot(tree, cex=.8, label.offset =.1, no.margin = TRUE)
```

```{r check tree Curtis1998}
tree$tip.label
tree$tip.label <- gsub("_"," ", tree$tip.label) #get rid of underscores
#tree$node.label <- NULL #delete node labels
```

Replace back "populus euramericana"" with "populus deltoides"" and save the tree in the Newick format.

```{r replace back populus hybrid }
tree$tip.label <- gsub( "Populus deltoides", "Populus euramericana", tree$tip.label)

#check the tree match
intersect(as.character(tree$tip.label), species_Curtis1998) #matches - they are contained in the tree and our species list
setdiff(species_Curtis1998, as.character(tree$tip.label)) #2 mismatches - these are only in species_Curtis1998
setdiff(as.character(tree$tip.label), species_Curtis1998) #2 mismatches - these are only in tree$tip.label

#amend the tip labels on the tree to match the species names from Curtis1998
tree$tip.label <- gsub("Fuscospora fusca", "Nothofagus fusca", tree$tip.label) #replace the tip label back with our species name
tree$tip.label <- gsub("Populus euramericana", "Populus deltoides", tree$tip.label) #replace the tip label back with our species name

#check the tree match
intersect(as.character(tree$tip.label), species_Curtis1998) #matches - they are contained in the tree and our species list
setdiff(species_Curtis1998, as.character(tree$tip.label)) #0 mismatches 
setdiff(as.character(tree$tip.label), species_Curtis1998) #0 mismatches

#write.tree(tree, file="tree_curtis1998.tre") #save the tree, if needed
#read.tree(file="tree_curtis1998.tre") #if you need to read it
```

Check if our tree is binary (you can see in the plot that there are a few polytomies).

```{r check Curtis1998 tree binary}
is.binary.tree(tree) #FALSE (i.e. there are polytomies)
#is.ultrametric(tree) #the tree has no branch lengths, so cant be checked
```

Its not binbary -  polytomies are present within genera: Acer, Betula, Quercus and Fraxinus.
We will resolve them at random.

```{r resolve Curtis1998 polytomies at random}
set.seed(111) #making it replicable
tree_random <- multi2di(tree,random=TRUE)
is.binary.tree(tree_random) #TRUE
write.tree(tree_random, file="tree_curtis1998_random.tre") #save the tree
#read.tree(tree_random="tree_curtis1998_random.tre") #if you need to read it
```

Plot the tree now - with randomly resolved polytomies.

```{r plot Curtis1998 tree_random, fig.width=10, fig.height=8, echo=TRUE, message=FALSE}
plot(tree_random)
```

Other, more involved approaches for resolving polytomies:

1. Resolving one or more multifurcations in all possible ways: http://blog.phytools.org/2016/08/resolving-one-or-more-multifurcations.html

2. Package for resolving by known discrete characters of the species_Curtis1998: https://www.rdocumentation.org/packages/paleotree/versions/3.1.3/topics/resolveTreeChar

3. Resolving manually by useing information from known more specialised published phylogeny without polytomies (see next exercise for how to find these).

********************************************************************************

### EXAMPLE 4 - using *treebase* package

Finding trees using Treebase database and *treebase* package (Treebase package tutorial: https://cran.r-project.org/web/packages/treebase/vignettes/vignette.html and https://www.carlboettiger.info/2011/05/16/treebase-in-r-a-first-tutorial.html).  
We will try to find more resolved phylogeny for the four species from genus Acer that were included in Curtis1998 tree.
You can search Treebase directly at www.treebase.org, but we will do this from R (note both approaches can be quite slow).

We need to creat a custom search string with our four Acer species names from Curtis1998 dataset.

```{r make treebase searchstring}
#Acer <- search_treebase("Acer_saccharinumc", by="taxon", max_trees=1) #example of using with a single species name - note different commas than in the tutorial and the underscores!
Acer4sp <- species[grep("Acer ", species)]
Acer4sp
Acer4sp <- gsub(" ","_", Acer4sp) #get underscores back
Acer4sp_string <- paste(Acer4sp, sep = "", collapse = ' and ') 
```

Run the treebase search - get amx 50 trees that contain exactly these species names.

```{r search treebase}
Acer_treebase <- search_treebase(Acer4sp_string, by=c("taxon","taxon","taxon","taxon"), max_trees=50, exact_match = TRUE, only_metadata = TRUE) #note different commas than in the tutorial!
Acer_treebase
str(Acer_treebase)
#for easier viewing put these inn a data frame (note that multiple tree IDs can come from the same study).
t(data.frame(matrix(unlist(Acer_treebase), nrow=6, byrow=F), row.names = c("S.id", "Tr.id", "type", "kind", "quality", "ntax"))) #note the study IDs that you are using - you can find these on Treebase
```
There are multiple trees available from several studies.  
Unfortunately the functions in treebase that are supposed to fetch metadata for the trees dont seem to work (show_metadata, metadata - if you get them workinng please let me know).

From oooking at Treebase website I know that this one will be good:   
tree Tr418 Study S342 - Ackerly D., Donoghue M.J.	1998	Leaf size, sapling allometry, and Corner's rules: phylogeny and correlated evolution in maples (Acer).	American Naturalist

```{r dwonload single tree from treebase}
Acers_tree <- search_treebase("418", by="id.tree", exact_match = TRUE) #returns a single tree
Acers_tree <- Acers_tree[[1]]
```

Plot the tree dwonloaded from Treebase - looks good.

```{r plot Acers_tree, fig.width=10, fig.height=8, echo=TRUE, message=FALSE}
plot(Acers_tree)
```

Prune this tree to our four species.

```{r prune Acers_tree}
#Prune Acers_tree to a list of four taxa from our list
Acers4sp_tree <- drop.tip(Acers_tree, Acers_tree$tip.label[-match(Acer4sp, Acers_tree$tip.label)])
```

Plot the pruned tree.

```{r plot pruned Acers_tree, fig.width=10, fig.height=4, echo=TRUE, message=FALSE}
plot(Acers4sp_tree)
```

Final checks.

```{r check Acers_tree}
###check the overlap between species name from the data fila nad the tree file
intersect(as.character(Acers4sp_tree$tip.label), Acer4sp) # all matching
setdiff(Acer4sp, as.character(Acers4sp_tree$tip.label)) #no mismatches
is.binary.tree(tree_random) #TRUE
```

********************************************************************************

### EXAMPLE 5 - try it yourself

Use rotl to create a tree for the bunch of reptiles.   
*Remember*: Make sure that species names match exactly tree tip labels

```{r rotl exercise, echo = TRUE, eval = TRUE}
myspecies2 <- c("Gekko japonicus", "Geochelone sulcata", "Gopherus agassizii", "Gopherus polyphemus", "Graptemys ouachitensis", "Graptemys pseudogeographica", "Hemidactylus bowringii", "Heterodon platirhinos", "Heteronotia binoei", "Iberolacerta cyeni", "Lacerta schreiberi", "Lampropholis delicata", "Lampropholis guichenoti", "Liasis fuscus")
```

```{r rotl solution, echo = TRUE, eval = TRUE}
taxa2 <- tnrs_match_names(names = myspecies2)
taxa2 #visual check

tree2 <- tol_induced_subtree(ott_ids= taxa2[["ott_id"]])
plot(tree2, cex=.8, label.offset =.1, no.margin = TRUE)

#change name on the tip lebel to match the original species list
tree2$tip.label <- gsub("\\_ott.*","", tree2$tip.label)
tree2$tip.label <- gsub("_"," ", tree2$tip.label)
tree2$tip.label
#tree$node.label <- NULL #delete node labels

#check if the tree2 is really binary 
is.binary.tree(tree2) #TRUE (i.e. there are no polytomies)

#check the overlap between species name from the list nad the tree2 
intersect(as.character(tree2$tip.label), myspecies2) #12 matching names
setdiff(myspecies2, as.character(tree2$tip.label)) #2 mismatches - these are only in myspecies2
setdiff(as.character(tree2$tip.label), myspecies2) #2 mismatches - these are only in the tree2$tip.label

#rename the tip labels on the tree to match the species names from our list
tree2$tip.label <- gsub("Centrochelys sulcata", "Geochelone sulcata", tree2$tip.label) 
tree2$tip.label <- gsub("Iberolacerta cyreni", "Iberolacerta cyeni", tree2$tip.label) 
#Note: you should actually check which name from each pair is "more correct" and accordingly check the names either on the tree or in your species list / data set

#re-check the overlap between species name from the list nad the tree2 
intersect(as.character(tree2$tip.label), myspecies2) #14 matching names
setdiff(myspecies2, as.character(tree2$tip.label)) #0 mismatches - these are only in myspecies2
setdiff(as.character(tree2$tip.label), myspecies2) #0 mismatches - these are only in the tree2$tip.label
```

********************************************************************************

### DISCUSSION

The general flow of worki is shown in the diagram below. 
You will usually start with species list from your data set (use binomial latin names). Check if there is any phylogenetic tree available that already contains all your species (e.g. a supertree). If such tree exist, just trim it down to your species. If no existing phylogeny is usable, there are a few non-exclusive options depending on the range of species in the data set:
- Use rotl - an R package (https://peerj.com/preprints/1471/) to access synthetic phylogenetic tree available at Open Tree of Life database (https://opentree.wikispaces.com/). Note: crashes on some species lists.  
- If the species are very diverged (e.g. mixture of vertabrates and invertebrates), use NCBI Taxonomy Common Tree tool (http://www.ncbi.nlm.nih.gov/Taxonomy/CommonTree/wwwcmt.cgi). At the webpage, use an input box at the top of the page to search each species name or upload lst of the species names from a file. This allows to generate a scaffold for a tree. Then resolve polytomies and add missing species (e.g. by grafting) at smaller scales, if needed. [iTOL  tool at http://itol.embl.de/other_trees.shtml is no longer available].  
- If few major clades are present, assemble together sub-trees created from bigger trees, if these exist for these clades.   
- Use phyloGenerator (http://willpearse.github.io/phyloGenerator/) or other phylogenetic software to create phylogeny based on DNA sequences, if a shared sequence set exists for a given set of species (check if the resulting phylogeny makes sense).

Be aware of species names problems: typos, scientific name synonyms, subspecies - have to be re-checked / corrected / substituted if generate errors at any stage.

Resolving polytomies: using existing phylogenetic information from other sources (e.g. published phylogenies) or at random.

```{r diagram, fig.width=10, fig.height=10, echo=FALSE, message=FALSE}
par(mar=c(1,1,1,1))
openplotmat()
elpos <- coordinates (pos = c(1, 1, 4, 1, 1, 2, 1))

fromto <- matrix(ncol = 2, byrow = TRUE, data = c(1,2, 2,3, 2,4, 2,5, 2,6, 3,7, 4,7, 5,7, 6,7, 7,8, 8,9, 8,10, 9,11, 10,11, 8,11, 4,3 ))
nr <- nrow(fromto)
arrpos <- matrix(ncol = 2, nrow = nr)
for (i in 1:nr) (arrpos[i, ] <- straightarrow (to = elpos[fromto[i, 2], ],from = elpos[fromto[i, 1], ],lwd = 2, arr.pos = 0.6, arr.length = 0.5))

textrect (mid = elpos[1,], radx = 0.1, rady = 0.05, lwd = 2, lab = "species list", box.col = "white", shadow.col = "grey", shadow.size = 0.005, cex = 1.5)
textellipse (mid = elpos[2,], radx = 0.1, rady = 0.05, lwd = 1, lab = "supertree or bigger tree exists?", box.col = "white", shadow.col = "grey", shadow.size = 0.005, cex = 1)
textrect (mid = elpos[3,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("prune","supertree"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 1.2)
textrect (mid = elpos[4,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("merge","supertrees"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 1)
textrect (mid = elpos[5,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("generate","taxonomic"," tree using NCBI"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 0.8)
textrect (mid = elpos[6,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("generate","phylogeny","from sequences"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 0.8)
textrect (mid = elpos[7,], radx = 0.1, rady = 0.05, lwd = 1, lab = "preliminary tree", box.col = "white", shadow.col = "grey", shadow.size = 0.005, cex = 1.5)
textellipse (mid = elpos[8,], radx = 0.1, rady = 0.05, lwd = 1, lab = "resolve polytomies?", box.col = "white", shadow.col = "grey", shadow.size = 0.005, cex = 1)
textrect (mid = elpos[9,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("using existing","information"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 1.2)
textrect (mid = elpos[10,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("at random"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 1.2)
textrect (mid = elpos[11,], radx = 0.1, rady = 0.05, lwd = 2, lab = "final tree", box.col = "white", shadow.col = "grey", shadow.size = 0.005, cex = 1.5)

text(arrpos[2, 1] + 0.00, arrpos[2, 2] + 0.02, "yes")
text(arrpos[3, 1] - 0.01, arrpos[3, 2] + 0.02, "no")
text(arrpos[4, 1] + 0.01, arrpos[4, 2] + 0.02, "no")
text(arrpos[5, 1] + 0.01, arrpos[5, 2] + 0.02, "no")

text(arrpos[11, 1] + 0.00, arrpos[11, 2] + 0.02, "yes")
text(arrpos[12, 1] - 0.01, arrpos[12, 2] + 0.02, "yes")
text(arrpos[15, 1] + 0.015, arrpos[15, 2] + 0.1, "no") 
```

OTHER RESOURCES
---------------

Search for help: discussion list R-sig-phylo archive at http://www.mail-archive.com/r-sig-phylo@r-project.org/

Tree manipulation in R: http://www.r-phylo.org/wiki/HowTo/DataTreeManipulation

R-based GUI tree viewer: https://klash.shinyapps.io/shinyTreeViewer

### 2019-01-10 Data_extractions.Rmd
add html file and starting file for the next exercise

--- 2019-01-10 Trees_making.Rmd
@@
 ---
-title: "Trees"
+title: "Data_extractions"
 author: "ML"
@@
 Prepared by Malgorzata (Losia) Lagisz, BEES, UNSW, AU, losialagisz@yahoo.com.
-Last updated: 1 June 2016 by Losia Lagisz
 
@@
 
-Ecological and evolutionary meta-analyses usually aggregate data acros multiple species (there are single-species or non-species meta analyses too, but these are  minority).
-Since species are related due to shared evolutionary history, effect sizes of closely related species can be more similar to these of less-related specis. 
-This kind of non-independence can be represented as a correlation matrix of phylogenetic relatedness and added as a random factor to multilevel meta-analytic models. 
-To create a relatedness matrix among species we need a phylogenetic tree for the list of species from our data set.
+**TO DO**:
 
-Currently, the easiest way to build a custom phylogenetic tree usable for meta-analysis is by using an R package called *rotl*.
+Prepare a set of 3 DAH papers to be extracted using *metaDigitise* package
+See: https://github.com/daniel1noble/metaDigitise (link to the paper?)
 
-A few things to note:  
-1. We can use a tree with or without branch lengths (divergence measures).  
-2. The tree has to be binary (strictly bifurcating = no polytomies, i.e. only two branches are coming from each node).  
-3. The tree needs to be rooted (has a node representing the most recent common ancestor of all the species in the tree).   
-4. The tree tip labels (representing species) must match **exactly** with the species names in the data set (otherwise MA models will not run with them).  
-5. Trees can be also made and used for higher taxonomic levels than species, e.g. genus - level (not covered in this tutorial, but general principles are the same).
+USE ES_calculator to calculate ES?
 
@@
 #install.packages("devtools","fulltxt","ape","metafor")
-#update.packages("fulltxt")
-#devtools::install_github("ropensci/rotl", dependencies = TRUE, build_vignette=TRUE)
-#install.packages("rotl", "treebase", "diagram", "dplyr")
 
-library(readxl)
-library(ape)
-library(fulltext)
-library(metafor)
-library(rotl)
-library(treebase)
-library(diagram)
-library(dplyr)
 ```
-
-********************************************************************************
-
-### EXAMPLE 1 - using *rotl* package
-
-We will try to create a simple tree using 10 commonly used taxa (lab model species) from across the Tree of Life.  
-Rotl is an R package (https://peerj.com/preprints/1471/) allowing us to access synthetic phylogenetic tree available at Open Tree of Life database (https://opentree.wikispaces.com/).
-
-**Important**: use Latin names, not common names.   
-**Notes**: *rotl* crashes on some species lists. Also, Open Tree of Life database gets constantly updated with new info, so you can get different outputs at different timepoints.
-
-We use a function *tnrs_match_names*, which returns a data frame that lists the Open Tree identifiers potentially matching our species.  
-The function also returns other information, including alternative species names, to help users ensure that the taxa matched are the correct ones.
-
-```{r rotl 10species list}
-myspecies <- c("Escherichia colli", 
-               "Chlamydomonas reinhardtii", 
-               "Drosophila melanogaster",
-               "Arabidopsis thaliana",
-               "Rattus norvegicus",
-               "Mus musculus",
-               "Cavia porcellus",
-               "Xenopus laevis",
-               "Saccharomyces cervisae",
-               "Danio rerio")
-taxa <- tnrs_match_names(names = myspecies)
-taxa
-```
-
-Note that two of the species names in our list have typos, and are labelled TRUE in a column "approximate_match". Still, they were matched to correct species names.  
-If typos are detected, it is worth correcting the names in the original dataset and re-running tree search.
-
-```{r rotl 10species corrected}
-myspecies_fixed <- c("Escherichia coli", 
-               "Chlamydomonas reinhardtii", 
-               "Drosophila melanogaster",
-               "Arabidopsis thaliana",
-               "Rattus norvegicus",
-               "Mus musculus",
-               "Cavia porcellus",
-               "Xenopus laevis",
-               "Saccharomyces cerevisiae",
-               "Danio rerio")
-taxa <- tnrs_match_names(names = myspecies_fixed)
-taxa
-```
-
-When we are happy with how the taxon names are matched to the Open Tree identifiers, we can pass them to the
-function tol_induced_subtree to retrieve the relationships among these taxa (a trimmed sub-tree from the synthetic phylogenetic tree).  
-The tree is returned as an ape::phylo object and it can be manipulated, printed and saved easily using functions from the ape package.
-
-```{r rotl 10species plot}
-tree <- tol_induced_subtree(ott_ids = taxa[["ott_id"]], label_format = "name")
-plot(tree, cex=.8, label.offset =.1, no.margin = TRUE)
-```
-
-Note that the species names have the Open Tree identifiers appended to them and *E.coli* was replaced with "Emrcaott616ott617", which is non-taxon node within Bacteria.
-Given we dont have any ither bacteria in the dataset, and the position on the tree is correct, thats fine, we just need to rename the node back.
-
-```{r rotl 10species names}
-tree$tip.label #check tip labels
-ecoli <- tol_node_info(ott_id=474506, include_lineage=TRUE)
-ecoli
-tax_lineage(ecoli)
-tnrs_match_names(names = ("Escherichia")) #we can see what info is available for this genus - not much
-
-#just fix the tip label
-tree$tip.label <- gsub("mrcaott616ott617", "Escherichia coli", tree$tip.label) #replace the tip label back with our species name
-tree$tip.labe
-tree$tip.label <- gsub("_"," ", tree$tip.label) #get rid of the underscores
-```
-
-It is time to plot and tidy up the tree (note no branch lengths are included, they can be created later via simulations).
-
-```{r rotl 10species tree}
-#see the tree info
-tree
-#tree$node.label <- NULL #you can delete internal node labels
-plot(tree) #plot the tree
-```
-
-Do a final check whether the tri is binary and if the tip labels of the tree match exactly our (fixed for typos) list of species.
-
-```{r rotl 10species check}
-#check if the tree is really binary 
-is.binary.tree(tree) #TRUE (i.e. there are no polytomies)
-
-#check the overlap with our species list
-###check the overlap between species name from the data fila nad the tree file
-intersect(as.character(tree$tip.label), myspecies_fixed) #10 matching names
-setdiff(myspecies_fixed, as.character(tree$tip.label)) #0 mismatches
-setdiff(as.character(tree$tip.label), myspecies_fixed) #0 mismatches
-```
-
-********************************************************************************
-
-### EXAMPLE 2 - using fixed supertree from a file
-
-Create a phylogenetic tree for 65 bird species used in Rutkowska, J., Dubiec, A.
-& Nakagawa, S.
-(2014) All eggs are made equal: meta-analysis of egg sexual size dimorphism in birds.
-Journal of Evolutionary Biology.
-27: 153-160 http://onlinelibrary.wiley.com/doi/10.1111/jeb.12282/abstract
-
-There are a few main supertrees for birds, so they can be used to create a subtree containing only the species from the list.  
-Here we will use on of these supertrees "Ericson backbone" downloadable from http://birdtree.org/ (Jetz, W., G.
-H.
-Thomas, J.
-B.
-Joy, K.
-Hartmann, and A.
-O.
-Mooers.
-2012.
-The global diversity of birds in space and time.
-Nature 491:444-448.).
-The website http://birdtree.org/ has an online tool for subsetting their supertrees generating distribution of trees rather than one tree.
-For simplicity and demonstration puproses we will use R to subset (prune) a single bird supertree down to a smaller list of species.
-
-We have our custom list of bird species stored in a file *bird_list.csv* as a simple string factor and we have downloaded Ercison tree stored in a file *Ericson.tre*, in Newick format with branch lengths included.
-Load these files and check the species overlap.
-
-```{r list birds load} 
-birds <- unique(read.csv("bird_list.csv")$species_name) #load our list of bird species and make the names unique
-#str(birds)
-birds_stree <- read.tree("Ericson.tre") #load bird supertree
-birds_stree #9993 tips = species
-str(birds_stree) #has edge (branch) lengths
-birds_stree <- collapse.singles(birds_stree) #clean up singularities, just in case
-
-###check the overlap between species name from our list  and the bird supertree
-intersect(as.character(birds_stree$tip.label), birds) #51 - all our species matching - they are contained in the tree
-setdiff(birds, as.character(birds_stree$tip.label)) #0 mismatches - just making sure we have all our species included in the supertree
-```
-
-Prune supertree tree to a list of taxa from our list.
-
-```{r list birds prune} 
-pruned_birds_stree <- drop.tip(birds_stree,birds_stree$tip.label[-match(birds, birds_stree$tip.label)]) #pruning
-#check the pruned tree
-is.binary.tree(pruned_birds_stree) #TRUE
-is.ultrametric(pruned_birds_stree) #TRUE
-
-intersect(as.character(pruned_birds_stree$tip.label), birds) #51 - all our species matching - they are contained in the tree
-setdiff(birds, as.character(pruned_birds_stree$tip.label)) #0 mismatches - just making sure we have all our species included in the tree
-
-#write.tree(pruned_birds_stree, file = "birds_51sp_tree.tre", append = FALSE, digits = 10, tree.names = FALSE) #save, if needed
-```
-
-Plot the tree.
-
-```{r plot tree birds, fig.width=10, fig.height=10, message=FALSE}
-plot(pruned_birds_stree, cex=0.8) #plots with branch lengths when available
-```
-
-**Notes**: We were lucky that all our species had matches on the supertree.
-Quite often there are no matches for some of the species names.
-In such cases, first check the names for typos or unnecessary characters, then try to find synonymous names and check if these are included in the supertree.
-Finally, a sister species can be used when a species is truly missing.
-Remember to match the names on the final tree back to the original species list from the dataset.
-
-********************************************************************************
-
-### EXAMPLE 3 - using *rotl* package again
-
-We will now handle a larger set of more closely related taxa, from a real data set.  
-The data set we use the dataset *dat.curtis1998* (tudies on the Effects of Elevated CO2 Levels on Woody Plant Mass) included in the *metafor* package.
-
-```{r load Curtis1998 data}
-dat <- get(data(dat.curtis1998))
-str(dat) # 102 rows, many columns including various data
-```
-
-Extract and clean the list of species included. Note that genus and species name are stored in separate columns.
-
-```{r create Curtis1998 species list}
-species_Curtis1998 <- paste(dat$genus, dat$species, sep=" ") 
-str(species_Curtis1998)
-species_Curtis1998 <- unique(tolower(species_Curtis1998)) #make species name unique
-length(species_Curtis1998) #37
-species_Curtis1998 #note: "populusx euramericana" should be same as "populus euramericana"
-species_Curtis1998 <- gsub("populusx euramericana", "populus euramericana", species_Curtis1998) #merge 2 partially overlapping names (typo?)
-species_Curtis1998 <- unique(tolower(species_Curtis1998))
-length(species_Curtis1998) #36
-#capitalise first letters
-species_Curtis1998 <- paste(toupper(substr(species_Curtis1998, 1, 1)), substr(species_Curtis1998, 2, nchar(species_Curtis1998)), sep="")
-species_Curtis1998
-```
-
-We have 36 unique species_Curtis1998 names. We try to run rotl searc with them.
-
-```{r match Curtis1998 species_Curtis1998 list to otl taxa list}
-taxa <- tnrs_match_names(names = species_Curtis1998)
-dim(taxa) #36
-taxa
-```
-
-Note that some alternative names were found. We asssume these are correct ones for now.  
-Now that the taxon names are matched to the Open Tree identifiers, we can pass them to the function tol_induced_subtree to retrieve the relationships among these taxa.
-
-```{r get Curtis1998 tree}
-tree <- tol_induced_subtree(ott_ids= taxa[["ott_id"]], label_format = "name") 
-#The following OTT ids were not found: [3915043].
-taxa[taxa$ott_id==3915043,] #populus euramericana Populus x canadensis             FALSE 3915043       TRUE HYBRID
-#note: Populus × canadensis, known as Canadian poplar or Carolina poplar, is a naturally occurring hybrid of Populus nigra and Populus deltoides
-```
-
-We run into a problem with Populus - need to fix this by using an alternative name.
-
-```{r fix Curtis1998 species_Curtis1998 list}
-populus <- tnrs_match_names(names = ("populus deltoides"))
-populus
-#replace populus euramericana with populus deltoides (note 2 versions of the name)
-species_Curtis1998 <- gsub("Populus euramericana", "Populus deltoides", species_Curtis1998)
-```
-
-```{r get Curtis1998 tree again}
-taxa <- tnrs_match_names(names = species_Curtis1998)
-head(taxa)
-dim(taxa) #36
-tree <- tol_induced_subtree(ott_ids= taxa[["ott_id"]], label_format = "name") 
-str(tree)
-```
-
-Plot the tree:
-
-```{r plot tree Curtis1998, fig.width=10, fig.height=8, echo=TRUE, message=FALSE}
-plot(tree, cex=.8, label.offset =.1, no.margin = TRUE)
-```
-
-```{r check tree Curtis1998}
-tree$tip.label
-tree$tip.label <- gsub("_"," ", tree$tip.label) #get rid of underscores
-#tree$node.label <- NULL #delete node labels
-```
-
-Replace back "populus euramericana"" with "populus deltoides"" and save the tree in the Newick format.
-
-```{r replace back populus hybrid }
-tree$tip.label <- gsub( "Populus deltoides", "Populus euramericana", tree$tip.label)
-
-#check the tree match
-intersect(as.character(tree$tip.label), species_Curtis1998) #matches - they are contained in the tree and our species list
-setdiff(species_Curtis1998, as.character(tree$tip.label)) #2 mismatches - these are only in species_Curtis1998
-setdiff(as.character(tree$tip.label), species_Curtis1998) #2 mismatches - these are only in tree$tip.label
-
-#amend the tip labels on the tree to match the species names from Curtis1998
-tree$tip.label <- gsub("Fuscospora fusca", "Nothofagus fusca", tree$tip.label) #replace the tip label back with our species name
-tree$tip.label <- gsub("Populus euramericana", "Populus deltoides", tree$tip.label) #replace the tip label back with our species name
-
-#check the tree match
-intersect(as.character(tree$tip.label), species_Curtis1998) #matches - they are contained in the tree and our species list
-setdiff(species_Curtis1998, as.character(tree$tip.label)) #0 mismatches 
-setdiff(as.character(tree$tip.label), species_Curtis1998) #0 mismatches
-
-#write.tree(tree, file="tree_curtis1998.tre") #save the tree, if needed
-#read.tree(file="tree_curtis1998.tre") #if you need to read it
-```
-
-Check if our tree is binary (you can see in the plot that there are a few polytomies).
-
-```{r check Curtis1998 tree binary}
-is.binary.tree(tree) #FALSE (i.e. there are polytomies)
-#is.ultrametric(tree) #the tree has no branch lengths, so cant be checked
-```
-
-Its not binbary -  polytomies are present within genera: Acer, Betula, Quercus and Fraxinus.
-We will resolve them at random.
-
-```{r resolve Curtis1998 polytomies at random}
-set.seed(111) #making it replicable
-tree_random <- multi2di(tree,random=TRUE)
-is.binary.tree(tree_random) #TRUE
-write.tree(tree_random, file="tree_curtis1998_random.tre") #save the tree
-#read.tree(tree_random="tree_curtis1998_random.tre") #if you need to read it
-```
-
-Plot the tree now - with randomly resolved polytomies.
-
-```{r plot Curtis1998 tree_random, fig.width=10, fig.height=8, echo=TRUE, message=FALSE}
-plot(tree_random)
-```
-
-Other, more involved approaches for resolving polytomies:
-
-1. Resolving one or more multifurcations in all possible ways: http://blog.phytools.org/2016/08/resolving-one-or-more-multifurcations.html
-
-2. Package for resolving by known discrete characters of the species_Curtis1998: https://www.rdocumentation.org/packages/paleotree/versions/3.1.3/topics/resolveTreeChar
-
-3. Resolving manually by useing information from known more specialised published phylogeny without polytomies (see next exercise for how to find these).
-
-********************************************************************************
-
-### EXAMPLE 4 - using *treebase* package
-
-Finding trees using Treebase database and *treebase* package (Treebase package tutorial: https://cran.r-project.org/web/packages/treebase/vignettes/vignette.html and https://www.carlboettiger.info/2011/05/16/treebase-in-r-a-first-tutorial.html).
-We will try to find more resolved phylogeny for the four species from genus Acer that were included in Curtis1998 tree.
-You can search Treebase directly at www.treebase.org, but we will do this from R (note both approaches can be quite slow).
-
-We need to creat a custom search string with our four Acer species names from Curtis1998 dataset.
-
-```{r make treebase searchstring}
-#Acer <- search_treebase("Acer_saccharinumc", by="taxon", max_trees=1) #example of using with a single species name - note different commas than in the tutorial and the underscores!
-Acer4sp <- species[grep("Acer ", species)]
-Acer4sp
-Acer4sp <- gsub(" ","_", Acer4sp) #get underscores back
-Acer4sp_string <- paste(Acer4sp, sep = "", collapse = ' and ') 
-```
-
-Run the treebase search - get amx 50 trees that contain exactly these species names.
-
-```{r search treebase}
-Acer_treebase <- search_treebase(Acer4sp_string, by=c("taxon","taxon","taxon","taxon"), max_trees=50, exact_match = TRUE, only_metadata = TRUE) #note different commas than in the tutorial!
-Acer_treebase
-str(Acer_treebase)
-#for easier viewing put these inn a data frame (note that multiple tree IDs can come from the same study).
-t(data.frame(matrix(unlist(Acer_treebase), nrow=6, byrow=F), row.names = c("S.id", "Tr.id", "type", "kind", "quality", "ntax"))) #note the study IDs that you are using - you can find these on Treebase
-```
-There are multiple trees available from several studies.  
-Unfortunately the functions in treebase that are supposed to fetch metadata for the trees dont seem to work (show_metadata, metadata - if you get them workinng please let me know).
-
-From oooking at Treebase website I know that this one will be good:   
-tree Tr418 Study S342 - Ackerly D., Donoghue M.J.	1998	Leaf size, sapling allometry, and Corner's rules: phylogeny and correlated evolution in maples (Acer).	American Naturalist
-
-```{r dwonload single tree from treebase}
-Acers_tree <- search_treebase("418", by="id.tree", exact_match = TRUE) #returns a single tree
-Acers_tree <- Acers_tree[[1]]
-```
-
-Plot the tree dwonloaded from Treebase - looks good.
-
-```{r plot Acers_tree, fig.width=10, fig.height=8, echo=TRUE, message=FALSE}
-plot(Acers_tree)
-```
-
-Prune this tree to our four species.
-
-```{r prune Acers_tree}
-#Prune Acers_tree to a list of four taxa from our list
-Acers4sp_tree <- drop.tip(Acers_tree, Acers_tree$tip.label[-match(Acer4sp, Acers_tree$tip.label)])
-```
-
-Plot the pruned tree.
-
-```{r plot pruned Acers_tree, fig.width=10, fig.height=4, echo=TRUE, message=FALSE}
-plot(Acers4sp_tree)
-```
-
-Final checks.
-
-```{r check Acers_tree}
-###check the overlap between species name from the data fila nad the tree file
-intersect(as.character(Acers4sp_tree$tip.label), Acer4sp) # all matching
-setdiff(Acer4sp, as.character(Acers4sp_tree$tip.label)) #no mismatches
-is.binary.tree(tree_random) #TRUE
-```
-
-********************************************************************************
-
-### EXAMPLE 5 - try it yourself
-
-Use rotl to create a tree for the bunch of reptiles.   
-*Remember*: Make sure that species names match exactly tree tip labels
-
-```{r rotl exercise, echo = TRUE, eval = TRUE}
-myspecies2 <- c("Gekko japonicus", "Geochelone sulcata", "Gopherus agassizii", "Gopherus polyphemus", "Graptemys ouachitensis", "Graptemys pseudogeographica", "Hemidactylus bowringii", "Heterodon platirhinos", "Heteronotia binoei", "Iberolacerta cyeni", "Lacerta schreiberi", "Lampropholis delicata", "Lampropholis guichenoti", "Liasis fuscus")
-```
-
-```{r rotl solution, echo = TRUE, eval = TRUE}
-taxa2 <- tnrs_match_names(names = myspecies2)
-taxa2 #visual check
-
-tree2 <- tol_induced_subtree(ott_ids= taxa2[["ott_id"]])
-plot(tree2, cex=.8, label.offset =.1, no.margin = TRUE)
-
-#change name on the tip lebel to match the original species list
-tree2$tip.label <- gsub("\\_ott.*","", tree2$tip.label)
-tree2$tip.label <- gsub("_"," ", tree2$tip.label)
-tree2$tip.label
-#tree$node.label <- NULL #delete node labels
-
-#check if the tree2 is really binary 
-is.binary.tree(tree2) #TRUE (i.e. there are no polytomies)
-
-#check the overlap between species name from the list nad the tree2 
-intersect(as.character(tree2$tip.label), myspecies2) #12 matching names
-setdiff(myspecies2, as.character(tree2$tip.label)) #2 mismatches - these are only in myspecies2
-setdiff(as.character(tree2$tip.label), myspecies2) #2 mismatches - these are only in the tree2$tip.label
-
-#rename the tip labels on the tree to match the species names from our list
-tree2$tip.label <- gsub("Centrochelys sulcata", "Geochelone sulcata", tree2$tip.label) 
-tree2$tip.label <- gsub("Iberolacerta cyreni", "Iberolacerta cyeni", tree2$tip.label) 
-#Note: you should actually check which name from each pair is "more correct" and accordingly check the names either on the tree or in your species list / data set
-
-#re-check the overlap between species name from the list nad the tree2 
-intersect(as.character(tree2$tip.label), myspecies2) #14 matching names
-setdiff(myspecies2, as.character(tree2$tip.label)) #0 mismatches - these are only in myspecies2
-setdiff(as.character(tree2$tip.label), myspecies2) #0 mismatches - these are only in the tree2$tip.label
-```
-
-********************************************************************************
-
-### DISCUSSION
-
-The general flow of worki is shown in the diagram below. 
-You will usually start with species list from your data set (use binomial latin names).
-Check if there is any phylogenetic tree available that already contains all your species (e.g.
-a supertree).
-If such tree exist, just trim it down to your species.
-If no existing phylogeny is usable, there are a few non-exclusive options depending on the range of species in the data set:
-- Use rotl - an R package (https://peerj.com/preprints/1471/) to access synthetic phylogenetic tree available at Open Tree of Life database (https://opentree.wikispaces.com/).
-Note: crashes on some species lists.
-- If the species are very diverged (e.g.
-mixture of vertabrates and invertebrates), use NCBI Taxonomy Common Tree tool (http://www.ncbi.nlm.nih.gov/Taxonomy/CommonTree/wwwcmt.cgi).
-At the webpage, use an input box at the top of the page to search each species name or upload lst of the species names from a file.
-This allows to generate a scaffold for a tree.
-Then resolve polytomies and add missing species (e.g.
-by grafting) at smaller scales, if needed.
-[iTOL  tool at http://itol.embl.de/other_trees.shtml is no longer available].
-- If few major clades are present, assemble together sub-trees created from bigger trees, if these exist for these clades.   
-- Use phyloGenerator (http://willpearse.github.io/phyloGenerator/) or other phylogenetic software to create phylogeny based on DNA sequences, if a shared sequence set exists for a given set of species (check if the resulting phylogeny makes sense).
-
-Be aware of species names problems: typos, scientific name synonyms, subspecies - have to be re-checked / corrected / substituted if generate errors at any stage.
-
-Resolving polytomies: using existing phylogenetic information from other sources (e.g. published phylogenies) or at random.
-
-```{r diagram, fig.width=10, fig.height=10, echo=FALSE, message=FALSE}
-par(mar=c(1,1,1,1))
-openplotmat()
-elpos <- coordinates (pos = c(1, 1, 4, 1, 1, 2, 1))
-
-fromto <- matrix(ncol = 2, byrow = TRUE, data = c(1,2, 2,3, 2,4, 2,5, 2,6, 3,7, 4,7, 5,7, 6,7, 7,8, 8,9, 8,10, 9,11, 10,11, 8,11, 4,3 ))
-nr <- nrow(fromto)
-arrpos <- matrix(ncol = 2, nrow = nr)
-for (i in 1:nr) (arrpos[i, ] <- straightarrow (to = elpos[fromto[i, 2], ],from = elpos[fromto[i, 1], ],lwd = 2, arr.pos = 0.6, arr.length = 0.5))
-
-textrect (mid = elpos[1,], radx = 0.1, rady = 0.05, lwd = 2, lab = "species list", box.col = "white", shadow.col = "grey", shadow.size = 0.005, cex = 1.5)
-textellipse (mid = elpos[2,], radx = 0.1, rady = 0.05, lwd = 1, lab = "supertree or bigger tree exists?", box.col = "white", shadow.col = "grey", shadow.size = 0.005, cex = 1)
-textrect (mid = elpos[3,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("prune","supertree"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 1.2)
-textrect (mid = elpos[4,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("merge","supertrees"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 1)
-textrect (mid = elpos[5,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("generate","taxonomic"," tree using NCBI"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 0.8)
-textrect (mid = elpos[6,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("generate","phylogeny","from sequences"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 0.8)
-textrect (mid = elpos[7,], radx = 0.1, rady = 0.05, lwd = 1, lab = "preliminary tree", box.col = "white", shadow.col = "grey", shadow.size = 0.005, cex = 1.5)
-textellipse (mid = elpos[8,], radx = 0.1, rady = 0.05, lwd = 1, lab = "resolve polytomies?", box.col = "white", shadow.col = "grey", shadow.size = 0.005, cex = 1)
-textrect (mid = elpos[9,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("using existing","information"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 1.2)
-textrect (mid = elpos[10,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("at random"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 1.2)
-textrect (mid = elpos[11,], radx = 0.1, rady = 0.05, lwd = 2, lab = "final tree", box.col = "white", shadow.col = "grey", shadow.size = 0.005, cex = 1.5)
-
-text(arrpos[2, 1] + 0.00, arrpos[2, 2] + 0.02, "yes")
-text(arrpos[3, 1] - 0.01, arrpos[3, 2] + 0.02, "no")
-text(arrpos[4, 1] + 0.01, arrpos[4, 2] + 0.02, "no")
-text(arrpos[5, 1] + 0.01, arrpos[5, 2] + 0.02, "no")
-
-text(arrpos[11, 1] + 0.00, arrpos[11, 2] + 0.02, "yes")
-text(arrpos[12, 1] - 0.01, arrpos[12, 2] + 0.02, "yes")
-text(arrpos[15, 1] + 0.015, arrpos[15, 2] + 0.1, "no") 
-```
-
-OTHER RESOURCES
----------------
-
-Search for help: discussion list R-sig-phylo archive at http://www.mail-archive.com/r-sig-phylo@r-project.org/
-
-Tree manipulation in R: http://www.r-phylo.org/wiki/HowTo/DataTreeManipulation
-
-R-based GUI tree viewer: https://klash.shinyapps.io/shinyTreeViewer

### 2019-01-16 Trees_making.Rmd
removed update info
improved exercises after input from Alfredo and Nick

@@
 author: "ML"
-date: "10 January 2019"
+date: "16 January 2019"
 output: html_document
@@
 
-Prepared by Malgorzata (Losia) Lagisz, BEES, UNSW, AU, losialagisz@yahoo.com.  
-Last updated: 1 June 2016 by Losia Lagisz
+Prepared by Malgorzata (Losia) Lagisz, BEES, UNSW, AU, losialagisz@gmail.com
 
@@
 
-Ecological and evolutionary meta-analyses usually aggregate data acros multiple species (there are single-species or non-species meta analyses too, but these are  minority).
-Since species are related due to shared evolutionary history, effect sizes of closely related species can be more similar to these of less-related specis. 
+Ecological and evolutionary meta-analyses usually aggregate data across multiple species (there are single-species or non-species meta analyses too, but these are  minority).
+Since species are related due to shared evolutionary history, effect sizes of closely related species can be more similar to these of less-related species. 
 This kind of non-independence can be represented as a correlation matrix of phylogenetic relatedness and added as a random factor to multilevel meta-analytic models. 
 To create a relatedness matrix among species we need a phylogenetic tree for the list of species from our data set.
-
-Currently, the easiest way to build a custom phylogenetic tree usable for meta-analysis is by using an R package called *rotl*.
 
@@
 4. The tree tip labels (representing species) must match **exactly** with the species names in the data set (otherwise MA models will not run with them).  
-5. Trees can be also made and used for higher taxonomic levels than species, e.g. genus - level (not covered in this tutorial, but general principles are the same).
+5. Trees can be also made and used for higher taxonomic levels than species, e.g. genus-level (not covered in this tutorial, but general principles are the same).
 
-Start from installing (as required) and loading R packages.
+The general flow of tree-making work is shown in the diagram below.  
+You will usually start with species list from the mata-analytic data set (use binomial latin names for the species list and in your database).
+First, check if there is any phylogenetic tree available that already contains all your species (e.g.
+a supertree).
+Currently, the easiest way to build a custom phylogenetic tree usable for meta-analysis is by using an R package called *rotl*, which accesses a synthetic supertree from Open Tree of Life database (https://opentree.wikispaces.com/).
+We will use this approach in our examples below.
+Most of the time *rotl* will work well, especially with the common species.
+There are other super-trees available, e.g.
+for birds - we will use one of thes as our example too.
+Overall, if you can find a tree that contains all your specis, you just trim (subset) it down to your species.
+If a few species is missing, it is sometimes possible to substitute them with closely related species.
+If no existing phylogeny is readily usable, there are a few non-exclusive options depending on the range of species in the data set:
+ - If the species are very diverged (e.g.
+mixture of vertabrates and invertebrates), use NCBI Taxonomy Common Tree tool (http://www.ncbi.nlm.nih.gov/Taxonomy/CommonTree/wwwcmt.cgi).
+At the webpage, use an input box at the top of the page to search each species name or upload lst of the species names from a file.
+This allows to generate a scaffold for a tree.
+ - If few major clades are present, assemble together sub-trees created from bigger trees, if these exist for these clades.    
+ - Use phyloGenerator (http://willpearse.github.io/phyloGenerator/) or other phylogenetic software to create phylogeny based on DNA sequences, if a shared sequence set exists for a given set of species (check if the resulting phylogeny makes sense).
 
-```{r prepare, message = FALSE}
-#install.packages("devtools","fulltxt","ape","metafor")
-#update.packages("fulltxt")
-#devtools::install_github("ropensci/rotl", dependencies = TRUE, build_vignette=TRUE)
-#install.packages("rotl", "treebase", "diagram", "dplyr")
+If the resulting tree contains polytomies, these can be resolved at random, or based on information from more detailed trees (e.g.
+published phylogeneies of specific lower taxa).
+You can also graft missing species on your tree.
+Be aware of species names problems: typos, scientific name synonyms, subspecies - have to be re-checked / corrected / substituted if generate errors at any stage.
+Again, your final tree, must contain exactly the same list of species names as your meta-analytic data set.
 
-library(readxl)
-library(ape)
-library(fulltext)
-library(metafor)
+```{r prepare, message = FALSE, echo = TRUE, eval = FALSE, warning = FALSE}
+install.packages("diagram")
+install.packages("tidyverse")
+install.packages("ape")
+install.packages("curl")
+install.packages("fulltext")
+install.packages("treebase")
+install.packages("devtools")
+devtools::install_github("ropensci/rotl", dependencies = TRUE, build_vignette=TRUE)
+install.packages("rotl")
+```
+
+```{r load packages, message = FALSE, echo = TRUE, eval = TRUE, warning = FALSE}
+library(diagram, tidyverse)
+library(ape, curl)
+library(fulltext, metafor)
+library(treebase, devtools)
 library(rotl)
-library(treebase)
-library(diagram)
-library(dplyr)
 ```
+
+```{r diagram, fig.width=10, fig.height=10, echo=FALSE, message=FALSE}
+par(mar=c(1,1,1,1))
+openplotmat()
+elpos <- coordinates (pos = c(1, 1, 4, 1, 1, 2, 1))
+
+fromto <- matrix(ncol = 2, byrow = TRUE, data = c(1,2, 2,3, 2,4, 2,5, 2,6, 3,7, 4,7, 5,7, 6,7, 7,8, 8,9, 8,10, 9,11, 10,11, 8,11, 4,3 ))
+nr <- nrow(fromto)
+arrpos <- matrix(ncol = 2, nrow = nr)
+for (i in 1:nr) (arrpos[i, ] <- straightarrow (to = elpos[fromto[i, 2], ],from = elpos[fromto[i, 1], ],lwd = 2, arr.pos = 0.6, arr.length = 0.5))
+
+textrect (mid = elpos[1,], radx = 0.1, rady = 0.05, lwd = 2, lab = "species list", box.col = "white", shadow.col = "grey", shadow.size = 0.005, cex = 1.5)
+textellipse (mid = elpos[2,], radx = 0.1, rady = 0.05, lwd = 1, lab = "supertree or bigger tree exists?", box.col = "white", shadow.col = "grey", shadow.size = 0.005, cex = 1)
+textrect (mid = elpos[3,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("prune","supertree"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 1.2)
+textrect (mid = elpos[4,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("merge","supertrees"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 1)
+textrect (mid = elpos[5,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("generate","taxonomic"," tree using NCBI"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 0.8)
+textrect (mid = elpos[6,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("generate","phylogeny","from sequences"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 0.8)
+textrect (mid = elpos[7,], radx = 0.1, rady = 0.05, lwd = 1, lab = "preliminary tree", box.col = "white", shadow.col = "grey", shadow.size = 0.005, cex = 1.5)
+textellipse (mid = elpos[8,], radx = 0.1, rady = 0.05, lwd = 1, lab = "resolve polytomies?", box.col = "white", shadow.col = "grey", shadow.size = 0.005, cex = 1)
+textrect (mid = elpos[9,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("using existing","information"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 1.2)
+textrect (mid = elpos[10,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("at random"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 1.2)
+textrect (mid = elpos[11,], radx = 0.1, rady = 0.05, lwd = 2, lab = "final tree", box.col = "white", shadow.col = "grey", shadow.size = 0.005, cex = 1.5)
+
+text(arrpos[2, 1] + 0.00, arrpos[2, 2] + 0.02, "yes")
+text(arrpos[3, 1] - 0.01, arrpos[3, 2] + 0.02, "no")
+text(arrpos[4, 1] + 0.01, arrpos[4, 2] + 0.02, "no")
+text(arrpos[5, 1] + 0.01, arrpos[5, 2] + 0.02, "no")
+
+text(arrpos[11, 1] + 0.00, arrpos[11, 2] + 0.02, "yes")
+text(arrpos[12, 1] - 0.01, arrpos[12, 2] + 0.02, "yes")
+text(arrpos[15, 1] + 0.015, arrpos[15, 2] + 0.1, "no") 
+```
+
+Examples and exercises
+-------
+
+We will give you some examples on how to use *rotl* package to download sub-trees from the Open Tree of Life supertree, resolve polytomies, trim and plot trees and deal with some of the most common issues.
+We will try to do most of theses tasks in R, but doing many of these outside R (using websites, text and tree ditors, etc.) is also possible.
+We hope you get the idea of the main principles of the process.
+
+Now, we start from installing (as required) and loading R packages we will use.
 
@@
 
-We will try to create a simple tree using 10 commonly used taxa (lab model species) from across the Tree of Life.  
+We will try to create a simple tree using 10 commonly used taxa (lab model species) from across the tree of life.  
 Rotl is an R package (https://peerj.com/preprints/1471/) allowing us to access synthetic phylogenetic tree available at Open Tree of Life database (https://opentree.wikispaces.com/).
@@
 
-When we are happy with how the taxon names are matched to the Open Tree identifiers, we can pass them to the
+When we are happy with the taxa found Open Tree identifiers, we can pass them to the
 function tol_induced_subtree to retrieve the relationships among these taxa (a trimmed sub-tree from the synthetic phylogenetic tree).
+
+```{r rotl 10species tree, warning = FALSE}
+tree <- tol_induced_subtree(ott_ids = taxa[["ott_id"]], label_format = "name") #we can get a Warning meassage here about collapsing single nodes
+```
+
 The tree is returned as an ape::phylo object and it can be manipulated, printed and saved easily using functions from the ape package.
 
-```{r rotl 10species plot}
-tree <- tol_induced_subtree(ott_ids = taxa[["ott_id"]], label_format = "name")
+```{r rotl 10species plot, fig.width=10, fig.height=5, echo=TRUE, message=FALSE}
 plot(tree, cex=.8, label.offset =.1, no.margin = TRUE)
@@
 
-Note that the species names have the Open Tree identifiers appended to them and *E.coli* was replaced with "Emrcaott616ott617", which is non-taxon node within Bacteria.
-Given we dont have any ither bacteria in the dataset, and the position on the tree is correct, thats fine, we just need to rename the node back.
+Note that the species names have the Open Tree identifiers appended to them and *E.coli* was replaced with "mrcaott616ott617", which is non-taxon node within Bacteria.
+Given we dont have any other bacteria in the dataset, and the position on the tree is correct, thats fine, we just need to rename the tree tip back to "Escherichia coli".
 
 ```{r rotl 10species names}
-tree$tip.label #check tip labels
+tree$tip.label #see the current tree tip labels
 ecoli <- tol_node_info(ott_id=474506, include_lineage=TRUE)
-ecoli
-tax_lineage(ecoli)
-tnrs_match_names(names = ("Escherichia")) #we can see what info is available for this genus - not much
-
-#just fix the tip label
+tax_lineage(ecoli) #domain within Bacteria
+tnrs_match_names(names = ("Escherichia")) #we can see what info is available for this genus - not much, ooks like a messy bit of a tree flagged as "INCONSISTENT"
 tree$tip.label <- gsub("mrcaott616ott617", "Escherichia coli", tree$tip.label) #replace the tip label back with our species name
-tree$tip.labe
 tree$tip.label <- gsub("_"," ", tree$tip.label) #get rid of the underscores
@@
 
-```{r rotl 10species tree}
-#see the tree info
-tree
-#tree$node.label <- NULL #you can delete internal node labels
-plot(tree) #plot the tree
+```{r rotl 10species tree Ecoli, fig.width=10, fig.height=5, echo=TRUE, message=FALSE}
+# tree #see the tree info
+# tree$node.label <- NULL #you can delete internal node labels
+plot(tree, cex=.8, label.offset =.1, no.margin = TRUE) #plot the tree
 ```
 
-Do a final check whether the tri is binary and if the tip labels of the tree match exactly our (fixed for typos) list of species.
+Do a final check whether the tree is binary and if the tip labels of the tree match exactly our (fixed for typos) list of species.
 
@@
 
-#check the overlap with our species list
-###check the overlap between species name from the data fila nad the tree file
+#check the overlap between the species names from the data set (our species list) nad the final tree
 intersect(as.character(tree$tip.label), myspecies_fixed) #10 matching names
@@
 
+That was hopefully easy! We can move to a different set of species and method...
+
 ********************************************************************************
 
-### EXAMPLE 2 - using fixed supertree from a file
+### EXAMPLE 2 - using a supertree from a file
 
-Create a phylogenetic tree for 65 bird species used in Rutkowska, J., Dubiec, A.
+We will now create a phylogenetic tree for 65 bird species used in Rutkowska, J., Dubiec, A.
 & Nakagawa, S.
@@
 
-There are a few main supertrees for birds, so they can be used to create a subtree containing only the species from the list.  
-Here we will use on of these supertrees "Ericson backbone" downloadable from http://birdtree.org/ (Jetz, W., G.
+There are a few main existing and established supertrees for birds, so they can be used to create a subtree containing only the species from the list.  
+Here we will use on of these supertrees called "Ericson backbone", which is downloadable from http://birdtree.org/ (Jetz, W., G.
 H.
@@
 Nature 491:444-448.).
-The website http://birdtree.org/ has an online tool for subsetting their supertrees generating distribution of trees rather than one tree.
+The website http://birdtree.org/ also has an online tool for subsetting their supertrees generating distribution of trees rather than one tree.
 For simplicity and demonstration puproses we will use R to subset (prune) a single bird supertree down to a smaller list of species.
@@
 birds <- unique(read.csv("bird_list.csv")$species_name) #load our list of bird species and make the names unique
-#str(birds)
 birds_stree <- read.tree("Ericson.tre") #load bird supertree
 birds_stree #9993 tips = species
-str(birds_stree) #has edge (branch) lengths
-birds_stree <- collapse.singles(birds_stree) #clean up singularities, just in case
+# str(birds_stree) # ypu can see this tree has edge (branch) lengths
 
@@
 intersect(as.character(birds_stree$tip.label), birds) #51 - all our species matching - they are contained in the tree
-setdiff(birds, as.character(birds_stree$tip.label)) #0 mismatches - just making sure we have all our species included in the supertree
+setdiff(birds, as.character(birds_stree$tip.label)) #0 mismatches - just making sure we have all our species included in the supertree, if you get any mismatches here look for typos or name synonyms!
 ```
@@
 is.ultrametric(pruned_birds_stree) #TRUE
+has.singles(pruned_birds_stree) #check for singles - the single nodes (i.e., with a single descendant) in a tree
+# pruned_birds_stree <- collapse.singles(pruned_birds_stree) #if singles are preset, you can clean them up using collapse.singles function
 
-intersect(as.character(pruned_birds_stree$tip.label), birds) #51 - all our species matching - they are contained in the tree
+# intersect(as.character(pruned_birds_stree$tip.label), birds) #51 - all our species matching - they are contained in the tree
 setdiff(birds, as.character(pruned_birds_stree$tip.label)) #0 mismatches - just making sure we have all our species included in the tree
-
-#write.tree(pruned_birds_stree, file = "birds_51sp_tree.tre", append = FALSE, digits = 10, tree.names = FALSE) #save, if needed
+# write.tree(pruned_birds_stree, file = "birds_51sp_tree.tre", append = FALSE, digits = 10, tree.names = FALSE) #save, if needed
 ```
@@
 
-```{r plot tree birds, fig.width=10, fig.height=10, message=FALSE}
+```{r plot tree birds, fig.width=10, fig.height=8, message=FALSE}
 plot(pruned_birds_stree, cex=0.8) #plots with branch lengths when available
@@
 We will now handle a larger set of more closely related taxa, from a real data set.  
-The data set we use the dataset *dat.curtis1998* (tudies on the Effects of Elevated CO2 Levels on Woody Plant Mass) included in the *metafor* package.
+The data set we use the dataset *dat.curtis1998* (contains studies on the Effects of Elevated CO2 Levels on Woody Plant Mass) included in the *metafor* package.
 
 ```{r load Curtis1998 data}
-dat <- get(data(dat.curtis1998))
-str(dat) # 102 rows, many columns including various data
+dat <- metafor::dat.curtis1998
+str(dat) #102 rows, many columns including various data
 ```
@@
 species_Curtis1998 <- paste(dat$genus, dat$species, sep=" ") 
-str(species_Curtis1998)
+# str(species_Curtis1998)
 species_Curtis1998 <- unique(tolower(species_Curtis1998)) #make species name unique
@@
 species_Curtis1998 #note: "populusx euramericana" should be same as "populus euramericana"
-species_Curtis1998 <- gsub("populusx euramericana", "populus euramericana", species_Curtis1998) #merge 2 partially overlapping names (typo?)
-species_Curtis1998 <- unique(tolower(species_Curtis1998))
+species_Curtis1998 <- gsub("populusx euramericana", "populus euramericana", species_Curtis1998) #merge 2 partially overlapping names (typo? or hybrid?) - remember to fix this in the main data set!
+species_Curtis1998 <- unique(tolower(species_Curtis1998)) #make the names unique again
 length(species_Curtis1998) #36
-#capitalise first letters
-species_Curtis1998 <- paste(toupper(substr(species_Curtis1998, 1, 1)), substr(species_Curtis1998, 2, nchar(species_Curtis1998)), sep="")
-species_Curtis1998
+species_Curtis1998 <- paste(toupper(substr(species_Curtis1998, 1, 1)), substr(species_Curtis1998, 2, nchar(species_Curtis1998)), sep="") #capitalise first letters
+species_Curtis1998 #looks better
 ```
 
-We have 36 unique species_Curtis1998 names. We try to run rotl searc with them.
+We now have 36 unique species_Curtis1998 names. We try to run *rotl* search with them.
 
@@
 taxa <- tnrs_match_names(names = species_Curtis1998)
-dim(taxa) #36
-taxa
+dim(taxa) #36 species
+# taxa
 ```
 
-Note that some alternative names were found. We asssume these are correct ones for now.  
+If you display the names (taxa), note that some alternative names were found. We asssume these are correct ones for now.  
 Now that the taxon names are matched to the Open Tree identifiers, we can pass them to the function tol_induced_subtree to retrieve the relationships among these taxa.
@@
 tree <- tol_induced_subtree(ott_ids= taxa[["ott_id"]], label_format = "name") 
-#The following OTT ids were not found: [3915043].
+#We get an error message that one OTT id was not found: [3915043].
 taxa[taxa$ott_id==3915043,] #populus euramericana Populus x canadensis             FALSE 3915043       TRUE HYBRID
@@
 
-We run into a problem with Populus - need to fix this by using an alternative name.
+So, we run into a problem with one of the Populus names - need to fix this by using an alternative name.
 
@@
 populus <- tnrs_match_names(names = ("populus deltoides"))
-populus
-#replace populus euramericana with populus deltoides (note 2 versions of the name)
-species_Curtis1998 <- gsub("Populus euramericana", "Populus deltoides", species_Curtis1998)
+populus #looks ok
+species_Curtis1998 <- gsub("Populus euramericana", "Populus deltoides", species_Curtis1998) #replace populus euramericana with populus deltoides
 ```
 
-```{r get Curtis1998 tree again}
+Re-run *rotl* tree search.
+
+```{r get Curtis1998 tree again, warning = FALSE}
 taxa <- tnrs_match_names(names = species_Curtis1998)
-head(taxa)
 dim(taxa) #36
-tree <- tol_induced_subtree(ott_ids= taxa[["ott_id"]], label_format = "name") 
+tree <- tol_induced_subtree(ott_ids= taxa[["ott_id"]], label_format = "name")  #we can get a Warning meassage here about collapsing single nodes
 str(tree)
@@
 
-```{r check tree Curtis1998}
-tree$tip.label
+Replace back "Populus euramericana" with "Populus deltoides" and save the tree in the Newick format.
+
+```{r check replace back populus hybrid }
 tree$tip.label <- gsub("_"," ", tree$tip.label) #get rid of underscores
-#tree$node.label <- NULL #delete node labels
-```
-
-Replace back "populus euramericana"" with "populus deltoides"" and save the tree in the Newick format.
-
-```{r replace back populus hybrid }
-tree$tip.label <- gsub( "Populus deltoides", "Populus euramericana", tree$tip.label)
+# tree$node.label <- NULL #delete node labels
 
 #check the tree match
-intersect(as.character(tree$tip.label), species_Curtis1998) #matches - they are contained in the tree and our species list
-setdiff(species_Curtis1998, as.character(tree$tip.label)) #2 mismatches - these are only in species_Curtis1998
-setdiff(as.character(tree$tip.label), species_Curtis1998) #2 mismatches - these are only in tree$tip.label
+# intersect(as.character(tree$tip.label), species_Curtis1998) #matches - they are contained in the tree and our species list
+setdiff(species_Curtis1998, as.character(tree$tip.label)) #1 mismatch - these are only in species_Curtis1998
+setdiff(as.character(tree$tip.label), species_Curtis1998) #1 mismatch - these are only in tree$tip.label
 
@@
 tree$tip.label <- gsub("Fuscospora fusca", "Nothofagus fusca", tree$tip.label) #replace the tip label back with our species name
-tree$tip.label <- gsub("Populus euramericana", "Populus deltoides", tree$tip.label) #replace the tip label back with our species name
 
-#check the tree match
-intersect(as.character(tree$tip.label), species_Curtis1998) #matches - they are contained in the tree and our species list
+#check the tree match again
+# intersect(as.character(tree$tip.label), species_Curtis1998) #matches - they are contained in the tree and our species list
 setdiff(species_Curtis1998, as.character(tree$tip.label)) #0 mismatches 
@@
 
-#write.tree(tree, file="tree_curtis1998.tre") #save the tree, if needed
-#read.tree(file="tree_curtis1998.tre") #if you need to read it
+# write.tree(tree, file="tree_curtis1998.tre") #save the tree in Newick format, if needed
+# read.tree(file="tree_curtis1998.tre") #if you need to read it
 ```
@@
 is.binary.tree(tree) #FALSE (i.e. there are polytomies)
-#is.ultrametric(tree) #the tree has no branch lengths, so cant be checked
+# is.ultrametric(tree) #the tree has no branch lengths, so cant be checked
 ```
@@
 write.tree(tree_random, file="tree_curtis1998_random.tre") #save the tree
-#read.tree(tree_random="tree_curtis1998_random.tre") #if you need to read it
+# read.tree(file="tree_curtis1998_random.tre") #if you need to read it
 ```
@@
 ```{r plot Curtis1998 tree_random, fig.width=10, fig.height=8, echo=TRUE, message=FALSE}
-plot(tree_random)
+plot(tree_random, cex=.8, label.offset =.1, no.margin = TRUE)
 ```
@@
 
-3. Resolving manually by useing information from known more specialised published phylogeny without polytomies (see next exercise for how to find these).
+3. Resolving manually by useing information from known more specialised published phylogeny without polytomies (see next exercise for how to try to find these).
 
@@
 
-Finding trees using Treebase database and *treebase* package (Treebase package tutorial: https://cran.r-project.org/web/packages/treebase/vignettes/vignette.html and https://www.carlboettiger.info/2011/05/16/treebase-in-r-a-first-tutorial.html).
-We will try to find more resolved phylogeny for the four species from genus Acer that were included in Curtis1998 tree.
-You can search Treebase directly at www.treebase.org, but we will do this from R (note both approaches can be quite slow).
+We can try to find some relevant trees using Treebase database and *treebase* package (Treebase package tutorial: https://cran.r-project.org/web/packages/treebase/vignettes/vignette.html and https://www.carlboettiger.info/2011/05/16/treebase-in-r-a-first-tutorial.html).
 
-We need to creat a custom search string with our four Acer species names from Curtis1998 dataset.
+As an exercise, we will try to find more resolved phylogeny for the four species from genus Acer that were included in Curtis1998 from R.
+However, Treebase sometimes may not connect or run the search string correctly. You can try search Treebase directly at www.treebase.org, but sometimes its quite slow or not working.
+So, just in case we also provide the full tree files downloaded from there, which can be used in this exercise.
+In R, we first try to find a tree via Treebase website search - go to https://treebase.org/treebase-web/search/studySearch.html and enter Acer in teh search box and select "Abstract" from the drop-down menu next to it, press "Search" button.
+There are 4 relevant studies which potentially contain relevant trees - we will use the S342 Ackerly D., Donoghue M.J.	1998, so check the box next to it and then "Discard unchecked items" button at the bottom.
+Then, go to the "Trees"" tab where you will se 2 trees from this study.
+We will use Tr418 - you can download a Nexus file by clicking the icons on the right.
+Not: you can start the search from Tree tab, by searching tree titles and then look at the studies from which these trees are, or even search for the most obscure species in the species tab - sometimes these approaches work better.
+
+Now we will try to do similar work from R. First, we create a custom search string with our four Acer species names from Curtis1998 dataset.
 
 ```{r make treebase searchstring}
-#Acer <- search_treebase("Acer_saccharinumc", by="taxon", max_trees=1) #example of using with a single species name - note different commas than in the tutorial and the underscores!
-Acer4sp <- species[grep("Acer ", species)]
-Acer4sp
-Acer4sp <- gsub(" ","_", Acer4sp) #get underscores back
+Acer4sp <- species_Curtis1998[grep("Acer ", species_Curtis1998)]
+Acer4sp #you should get "Acer rubrum"        "Acer saccharinum"   "Acer pensylvanicum" "Acer saccharum"  
+Acer4sp <- gsub(" ","_", Acer4sp) #put underscores back in the names
 Acer4sp_string <- paste(Acer4sp, sep = "", collapse = ' and ') 
+Acer4sp_string #you should get "Acer_rubrum and Acer_saccharinum and Acer_pensylvanicum and Acer_saccharum"
 ```
 
-Run the treebase search - get amx 50 trees that contain exactly these species names.
+Run the treebase search - get amx 50 trees that contain exactly these four species names.
 
 ```{r search treebase}
-Acer_treebase <- search_treebase(Acer4sp_string, by=c("taxon","taxon","taxon","taxon"), max_trees=50, exact_match = TRUE, only_metadata = TRUE) #note different commas than in the tutorial!
-Acer_treebase
-str(Acer_treebase)
-#for easier viewing put these inn a data frame (note that multiple tree IDs can come from the same study).
+#this is an example of how to search for a single species:
+# Acer <- search_treebase("Acer_saccharinumc", by="taxon", max_trees=10) #example of using with a single species name - note different commas than in the tutorial and the underscores!
+
+Acer_treebase <- search_treebase(Acer4sp_string, by=c("taxon","taxon","taxon","taxon"), max_trees=50, exact_match = TRUE, only_metadata = TRUE) #note different commas than in the treebase tutorial!
+# Acer_treebase #list of 49 meta-data results of our search 
+# str(Acer_treebase) #some more info can be seen here
+
+#Unfortunately the functions in treebase that are supposed to fetch metadata for the trees dont seem to work (show_metadata, metadata - if you get them workinng please let me know). 
+#For easier viewing we will put these inn a data frame (note that multiple tree IDs can come from the same study).
 t(data.frame(matrix(unlist(Acer_treebase), nrow=6, byrow=F), row.names = c("S.id", "Tr.id", "type", "kind", "quality", "ntax"))) #note the study IDs that you are using - you can find these on Treebase
 ```
-There are multiple trees available from several studies.  
-Unfortunately the functions in treebase that are supposed to fetch metadata for the trees dont seem to work (show_metadata, metadata - if you get them workinng please let me know).
 
-From oooking at Treebase website I know that this one will be good:   
-tree Tr418 Study S342 - Ackerly D., Donoghue M.J.	1998	Leaf size, sapling allometry, and Corner's rules: phylogeny and correlated evolution in maples (Acer).	American Naturalist
+There are multiple trees available from several studies. Note that the lost of found results includes the tree 418 from study 342 we already identified in our Treebase website search.  
+So, we will request specifically tree Tr418 Study S342 - Ackerly D., Donoghue M.J.	1998	Leaf size, sapling allometry, and Corner's rules: phylogeny and correlated evolution in maples (Acer).	American Naturalist.
+It may take a moment, depending on the size of the tree.
+You can try requesting different trees later.
 
 ```{r dwonload single tree from treebase}
-Acers_tree <- search_treebase("418", by="id.tree", exact_match = TRUE) #returns a single tree
-Acers_tree <- Acers_tree[[1]]
+Acers_tree <- search_treebase("418", by="id.tree", exact_match = TRUE) #returns a single tree in a list
+Acers_tree <- Acers_tree[[1]] #extract tree from the list
 ```
 
-Plot the tree dwonloaded from Treebase - looks good.
+If downloading from R treebase did not work, upload tree saved in a file:
+```{r dwonload single tree from file, eval = FALSE}
+Acers_tree <- read.nexus(file = "T418.nex") #read Nexus tree file
+```
+
+Plot the Acers tree from Treebase - looks good.
 
 ```{r plot Acers_tree, fig.width=10, fig.height=8, echo=TRUE, message=FALSE}
-plot(Acers_tree)
+plot(Acers_tree, cex=.8, label.offset =.1, no.margin = TRUE)
 ```
@@
 ```{r plot pruned Acers_tree, fig.width=10, fig.height=4, echo=TRUE, message=FALSE}
-plot(Acers4sp_tree)
+plot(Acers4sp_tree, cex=.8, label.offset =.1, no.margin = TRUE)
 ```
@@
 
-### EXAMPLE 5 - try it yourself
+### EXAMPLE 5 - try it yourself - rotl
 
-Use rotl to create a tree for the bunch of reptiles.   
-*Remember*: Make sure that species names match exactly tree tip labels
+Use *rotl* to create a tree for the bunch of reptiles.   
+**Remember**: Make sure that species names match exactly tree tip labels.
+
+Here is the species list:
 
@@
 
-```{r rotl solution, echo = TRUE, eval = TRUE}
-taxa2 <- tnrs_match_names(names = myspecies2)
-taxa2 #visual check
+********************************************************************************
 
-tree2 <- tol_induced_subtree(ott_ids= taxa2[["ott_id"]])
-plot(tree2, cex=.8, label.offset =.1, no.margin = TRUE)
+### EXAMPLE 6 - try it yourself- treebase
 
-#change name on the tip lebel to match the original species list
-tree2$tip.label <- gsub("\\_ott.*","", tree2$tip.label)
-tree2$tip.label <- gsub("_"," ", tree2$tip.label)
-tree2$tip.label
-#tree$node.label <- NULL #delete node labels
+Use *treebase* to create a tree for the three Quercus species from dat.Curtis1998.
 
-#check if the tree2 is really binary 
-is.binary.tree(tree2) #TRUE (i.e. there are no polytomies)
+**Hint**: Quercus prinus is also called Quercus montana (https://en.wikipedia.org/wiki/Quercus_montana).   
+Once you substitute, all three species can be found on Tree 113123 of Study 22057, and also in several other studies and trees.
 
-#check the overlap between species name from the list nad the tree2 
-intersect(as.character(tree2$tip.label), myspecies2) #12 matching names
-setdiff(myspecies2, as.character(tree2$tip.label)) #2 mismatches - these are only in myspecies2
-setdiff(as.character(tree2$tip.label), myspecies2) #2 mismatches - these are only in the tree2$tip.label
+Here is the species list:
 
-#rename the tip labels on the tree to match the species names from our list
-tree2$tip.label <- gsub("Centrochelys sulcata", "Geochelone sulcata", tree2$tip.label) 
-tree2$tip.label <- gsub("Iberolacerta cyreni", "Iberolacerta cyeni", tree2$tip.label) 
-#Note: you should actually check which name from each pair is "more correct" and accordingly check the names either on the tree or in your species list / data set
-
-#re-check the overlap between species name from the list nad the tree2 
-intersect(as.character(tree2$tip.label), myspecies2) #14 matching names
-setdiff(myspecies2, as.character(tree2$tip.label)) #0 mismatches - these are only in myspecies2
-setdiff(as.character(tree2$tip.label), myspecies2) #0 mismatches - these are only in the tree2$tip.label
+```{r treebase exercise, echo = TRUE, eval = TRUE, warning = FALSE}
+Quercus3sp <- species_Curtis1998[grep("Quercus ", species_Curtis1998)]
 ```
@@
 ### DISCUSSION
-
-The general flow of worki is shown in the diagram below. 
-You will usually start with species list from your data set (use binomial latin names).
-Check if there is any phylogenetic tree available that already contains all your species (e.g.
-a supertree).
-If such tree exist, just trim it down to your species.
-If no existing phylogeny is usable, there are a few non-exclusive options depending on the range of species in the data set:
-- Use rotl - an R package (https://peerj.com/preprints/1471/) to access synthetic phylogenetic tree available at Open Tree of Life database (https://opentree.wikispaces.com/).
-Note: crashes on some species lists.
-- If the species are very diverged (e.g.
-mixture of vertabrates and invertebrates), use NCBI Taxonomy Common Tree tool (http://www.ncbi.nlm.nih.gov/Taxonomy/CommonTree/wwwcmt.cgi).
-At the webpage, use an input box at the top of the page to search each species name or upload lst of the species names from a file.
-This allows to generate a scaffold for a tree.
-Then resolve polytomies and add missing species (e.g.
-by grafting) at smaller scales, if needed.
-[iTOL  tool at http://itol.embl.de/other_trees.shtml is no longer available].
-- If few major clades are present, assemble together sub-trees created from bigger trees, if these exist for these clades.   
-- Use phyloGenerator (http://willpearse.github.io/phyloGenerator/) or other phylogenetic software to create phylogeny based on DNA sequences, if a shared sequence set exists for a given set of species (check if the resulting phylogeny makes sense).
-
-Be aware of species names problems: typos, scientific name synonyms, subspecies - have to be re-checked / corrected / substituted if generate errors at any stage.
-
-Resolving polytomies: using existing phylogenetic information from other sources (e.g. published phylogenies) or at random.
-
-```{r diagram, fig.width=10, fig.height=10, echo=FALSE, message=FALSE}
-par(mar=c(1,1,1,1))
-openplotmat()
-elpos <- coordinates (pos = c(1, 1, 4, 1, 1, 2, 1))
-
-fromto <- matrix(ncol = 2, byrow = TRUE, data = c(1,2, 2,3, 2,4, 2,5, 2,6, 3,7, 4,7, 5,7, 6,7, 7,8, 8,9, 8,10, 9,11, 10,11, 8,11, 4,3 ))
-nr <- nrow(fromto)
-arrpos <- matrix(ncol = 2, nrow = nr)
-for (i in 1:nr) (arrpos[i, ] <- straightarrow (to = elpos[fromto[i, 2], ],from = elpos[fromto[i, 1], ],lwd = 2, arr.pos = 0.6, arr.length = 0.5))
-
-textrect (mid = elpos[1,], radx = 0.1, rady = 0.05, lwd = 2, lab = "species list", box.col = "white", shadow.col = "grey", shadow.size = 0.005, cex = 1.5)
-textellipse (mid = elpos[2,], radx = 0.1, rady = 0.05, lwd = 1, lab = "supertree or bigger tree exists?", box.col = "white", shadow.col = "grey", shadow.size = 0.005, cex = 1)
-textrect (mid = elpos[3,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("prune","supertree"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 1.2)
-textrect (mid = elpos[4,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("merge","supertrees"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 1)
-textrect (mid = elpos[5,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("generate","taxonomic"," tree using NCBI"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 0.8)
-textrect (mid = elpos[6,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("generate","phylogeny","from sequences"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 0.8)
-textrect (mid = elpos[7,], radx = 0.1, rady = 0.05, lwd = 1, lab = "preliminary tree", box.col = "white", shadow.col = "grey", shadow.size = 0.005, cex = 1.5)
-textellipse (mid = elpos[8,], radx = 0.1, rady = 0.05, lwd = 1, lab = "resolve polytomies?", box.col = "white", shadow.col = "grey", shadow.size = 0.005, cex = 1)
-textrect (mid = elpos[9,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("using existing","information"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 1.2)
-textrect (mid = elpos[10,], radx = 0.1, rady = 0.05, lwd = 1, lab = c("at random"), box.col = "white", shadow.col = "grey", shadow.size = 0.001, cex = 1.2)
-textrect (mid = elpos[11,], radx = 0.1, rady = 0.05, lwd = 2, lab = "final tree", box.col = "white", shadow.col = "grey", shadow.size = 0.005, cex = 1.5)
-
-text(arrpos[2, 1] + 0.00, arrpos[2, 2] + 0.02, "yes")
-text(arrpos[3, 1] - 0.01, arrpos[3, 2] + 0.02, "no")
-text(arrpos[4, 1] + 0.01, arrpos[4, 2] + 0.02, "no")
-text(arrpos[5, 1] + 0.01, arrpos[5, 2] + 0.02, "no")
-
-text(arrpos[11, 1] + 0.00, arrpos[11, 2] + 0.02, "yes")
-text(arrpos[12, 1] - 0.01, arrpos[12, 2] + 0.02, "yes")
-text(arrpos[15, 1] + 0.015, arrpos[15, 2] + 0.1, "no") 
-```
+W hope you learn a few tricks on how to make phylogenetic-like trees of species that can be used in meta-analytic models.
+There is much more to learn, especially dealing with missing or rare species names, grafting trees together, resolving polytomies, plotting cool trees etc.
+In some cases getting "perfect" tree may not be possible - but phylogenies are always just estimats of the species relationships, and as long as you get the key features right, the tree will be ok, and so your analyses (in most cases).
 

### 2019-01-16 Trees_making_solutions.Rmd
added treebase exercise and moved exercise solutions to a separate file

---
title: "Trees_making_solutions"
author: "ML"
date: "1/16/2019"
output: html_document
---

```{r prepare, message = FALSE, echo = FALSE, eval = TRUE, warning = FALSE}
knitr::opts_chunk$set(error = TRUE) #allow some execution errors for demonstration purposes
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, warning = FALSE) # install.packages("diagram")
# install.packages("tidyverse")
# install.packages("ape")
# install.packages("curl")
# install.packages("fulltext")
# install.packages("treebase")
# install.packages("devtools")
# devtools::install_github("ropensci/rotl", dependencies = TRUE, build_vignette=TRUE)
# install.packages("rotl")
library(diagram, tidyverse)
library(ape, curl)
library(fulltext, metafor)
library(treebase, devtools)
library(rotl)
```

********************************************************************************

### EXAMPLE 5 - try it yourself - rotl

Use *rotl* to create a tree for the bunch of reptiles.   
**Remember**: Make sure that species names match exactly tree tip labels

```{r rotl solution, echo = TRUE, eval = TRUE, warning = FALSE}
myspecies2 <- c("Gekko japonicus", "Geochelone sulcata", "Gopherus agassizii", "Gopherus polyphemus", "Graptemys ouachitensis", "Graptemys pseudogeographica", "Hemidactylus bowringii", "Heterodon platirhinos", "Heteronotia binoei", "Iberolacerta cyeni", "Lacerta schreiberi", "Lampropholis delicata", "Lampropholis guichenoti", "Liasis fuscus")

taxa2 <- tnrs_match_names(names = myspecies2)
taxa2 #visual check

tree2 <- tol_induced_subtree(ott_ids= taxa2[["ott_id"]], label_format = "name") #dont worry about the Warning message
plot(tree2, cex=.8, label.offset =.1, no.margin = TRUE)
tree2$tip.label <- gsub("_"," ", tree2$tip.label) #get rid of underscores
# tree2$tip.label
#tree$node.label <- NULL #delete node labels

#check if the tree2 is really binary 
is.binary.tree(tree2) #TRUE (i.e. there are no polytomies)

#check the overlap between species name from the list nad the tree2 
intersect(as.character(tree2$tip.label), myspecies2) #12 matching names
setdiff(myspecies2, as.character(tree2$tip.label)) #2 mismatches - these are only in myspecies2
setdiff(as.character(tree2$tip.label), myspecies2) #2 mismatches - these are only in the tree2$tip.label

#rename the tip labels on the tree to match the species names from our list
tree2$tip.label <- gsub("Centrochelys sulcata", "Geochelone sulcata", tree2$tip.label) 
tree2$tip.label <- gsub("Iberolacerta cyreni", "Iberolacerta cyeni", tree2$tip.label) 
#Note: you should actually check which name from each pair is "more correct" and accordingly check the names either on the tree or in your species list / data set

#re-check the overlap between species name from the list nad the tree2 
intersect(as.character(tree2$tip.label), myspecies2) #14 matching names
setdiff(myspecies2, as.character(tree2$tip.label)) #0 mismatches
setdiff(as.character(tree2$tip.label), myspecies2) #0 mismatches
```
Done!

********************************************************************************

### EXAMPLE 6 - try it yourself- treebase

Use *treebase* to create a tree for the three Quercus species from dat.Curtis1998.

**Hint**: Quercus prinus is also called Quercus montana (https://en.wikipedia.org/wiki/Quercus_montana).   
Once you substitute, all three species can be found on Tree 113123 of Study 22057, and also in several other studies and trees

```{r treebase solution, echo = TRUE, eval = TRUE, warning = FALSE}
#Code you need intially to get a list of species from Curtis 1998, in case you did not run Example 3:
dat <- metafor::dat.curtis1998
species_Curtis1998 <- paste(dat$genus, dat$species, sep=" ") 
species_Curtis1998 <- unique(tolower(species_Curtis1998)) #make species name unique
species_Curtis1998 <- gsub("populusx euramericana", "populus euramericana", species_Curtis1998) #merge 2 partially overlapping names (typo? or hybrid?) - remember to fix this in the main data set!
species_Curtis1998 <- unique(tolower(species_Curtis1998)) #make the names unique again
species_Curtis1998 <- paste(toupper(substr(species_Curtis1998, 1, 1)), substr(species_Curtis1998, 2, nchar(species_Curtis1998)), sep="") #capitalise first letters

#extract three Quercus species names:
Quercus3sp <- species_Curtis1998[grep("Quercus ", species_Curtis1998)]
Quercus3sp #3 species
Quercus3sp <- gsub(" ","_", Quercus3sp) #put underscores back in the names
Quercus3sp <- gsub("prinus","montana", Quercus3sp) #change the name to its currentky used synonym

Quercus3sp_string <- paste(Quercus3sp, sep = "", collapse = ' and ') 
Quercus3sp_string #you should get "Quercus_rubrum and Quercus_saccharinum and Quercus_pensylvanicum and Quercus_saccharum"

Quercus_treebase <- search_treebase(Quercus3sp_string, by=c("taxon","taxon","taxon"), max_trees=10, exact_match = TRUE, only_metadata = TRUE) #note different commas than in the treebase tutorial!
t(data.frame(matrix(unlist(Quercus_treebase), nrow=6, byrow=F), row.names = c("S.id", "Tr.id", "type", "kind", "quality", "ntax"))) #note the study IDs that you are using - you can find these on Treebase
Quercuss_tree <- search_treebase("113123", by="id.tree", exact_match = TRUE) #returns a single tree in a list 
Quercuss_tree <- Quercuss_tree[[1]] #extract tree from the list
plot(Quercuss_tree, cex=.8, label.offset =.1, no.margin = TRUE) #big tree of all sorts of trees - needs good trimming!
Quercus3sp_tree <- drop.tip(Quercuss_tree, Quercuss_tree$tip.label[-match(Quercus3sp, Quercuss_tree$tip.label)])
plot(Quercus3sp_tree, cex=.8, label.offset =.1, no.margin = TRUE)

###check the overlap between species name from the data fila nad the tree file
intersect(as.character(Quercus3sp_tree$tip.label), Quercus3sp) # all matching, but we changed the species name in teh list, so it will no longer match the original dataset
Quercus3sp_tree$tip.label <- gsub("montana", "prinus", Quercus3sp_tree$tip.label) #replace the tip label back with our species name
Quercus3sp_tree$tip.label <- gsub("_"," ", Quercus3sp_tree$tip.label) #get rid of the underscores
plot(Quercus3sp_tree, cex=.8, label.offset =.1, no.margin = TRUE)
```

Done!

### 2019-01-16 Trees_making_solutions.Rmd
added code to upload Quercuss tree from a local file

@@
 Quercuss_tree <- Quercuss_tree[[1]] #extract tree from the list
-plot(Quercuss_tree, cex=.8, label.offset =.1, no.margin = TRUE) #big tree of all sorts of trees - needs good trimming!
+#if Treebase does not work use: Quercuss_tree <- read.nexus(file = "T113123.nex") #read Nexus tree file
+
+# plot(Quercuss_tree, cex=.8, label.offset =.1, no.margin = TRUE) #a big tree of all sorts of trees - needs good trimming!
 Quercus3sp_tree <- drop.tip(Quercuss_tree, Quercuss_tree$tip.label[-match(Quercus3sp, Quercuss_tree$tip.label)])
@@
 
-###check the overlap between species name from the data fila nad the tree file
+#check the overlap between species name from the data fila nad the tree file
 intersect(as.character(Quercus3sp_tree$tip.label), Quercus3sp) # all matching, but we changed the species name in teh list, so it will no longer match the original dataset

### 2019-01-19 Trees_making.Rmd
added sessionInfo() line

@@
 knitr::opts_chunk$set(eval = TRUE, echo = TRUE, warning = FALSE)
+sessionInfo()
 ```

### 2019-01-19 Data_extractions.Rmd
added intro and story

@@
 
-Prepared by Malgorzata (Losia) Lagisz, BEES, UNSW, AU, losialagisz@yahoo.com.
+Prepared by Malgorzata (Losia) Lagisz, BEES, UNSW, AU, losialagisz@gmail.com.
 
-Introduction  
+### Purpose of the exercise:   
+In this exercise you get some practice and tips on:   
+ * Organising dataset  
+ * Extracting and coding
+ * Converting different types of information   
+ * Tidying up
+
+## Introduction  
 -------
 
+Imagine, you decided to carry on your Terminal Investment Meta-analysis and you actually fnished searching and screening literature for relevant papers.
+You have a collection of around 30 include full-text papers stored neatly as pdf files and supplementary information in whatever format available.
+
+********************************************************************************
+###  Organising dataset
+
+********************************************************************************
+###  Extracting and coding
+
 **TO DO**:  
-
 Prepare a set of 3 DAH papers to be extracted using *metaDigitise* package
 See: https://github.com/daniel1noble/metaDigitise (link to the paper?)
+Tools: metaDigitise
 
-USE ES_calculator to calculate ES?
+********************************************************************************
+###  Converting different types of information
+
+**TO DO**:  
+ES_calculator to convert ES?
+
+********************************************************************************
+###  Tidying up
 
@@
 ```{r prepare, message = FALSE}
-#install.packages("devtools","fulltxt","ape","metafor")
-
+install.packages("devtools")
+devtools::install_github("daniel1noble/metaDigitise")
+#install.packages("metaDigitise")
+library(metaDigitise)
 ```

### 2019-01-23 Papers_search.Rmd
adding first draft of searching and screening exercise

---
title: "Papers_search"
author: "ML"
date: "22 January 2019"
output: html_document
---

```{r setup, cache = F, echo=FALSE, results = FALSE}
knitr::opts_chunk$set(error = TRUE) #allow some execution errors for demonstration purposes
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, warning = FALSE, collapse = TRUE, comment = "#>")
sessionInfo()
```

```{r prepare, message = FALSE, echo = FALSE, eval = FALSE, warning = FALSE}
install.packages("diagram")
```

```{r load packages, message = FALSE, echo = FALSE, eval = TRUE, warning = FALSE}
library(diagram)
```

# EXERCISES - SEARCHING AND SCREENING PAPERS FOR META-ANALYSIS

Prepared by Malgorzata (Losia) Lagisz, BEES, UNSW, AU, losialagisz@gmail.com.

### Purpose of the exercise:   
In this exercise you get some practice and tips on:

 * Formulating the question for a meta-analysis   
 * Deciding on inclusion criteria   
 * Making decision trees   
 * Performing searches for relevant literature   
 * Screening the literature  
 * PRISMA and other documentation  
<br>

********************************************************************************  
### Introduction

Imagine, you woke up in the morning feeling rather sick. You decided to work from home to get some rest a get better more quickly. You dod not feel energetic enough to crunch numbers or write, so you thought you may catch up on scientific reading instead. You pick up the book sitting on top of the "to-read"" pile and open it at a random place. It happens to be **Ecoimmunology** (edited by Gregory Demas, Randy Nelson, Randy Joe Nelson, 2012) and it opens on the following fragment:

```{r TI_book_fragment, echo=FALSE, fig.cap=" ", out.width='50%'}
knitr::include_graphics("./images/TI_book_fragment.png")
```

You find it intriguing and start thinking this might be something (a topic) you could do a meta-analysis on.  *(Ok, this is a hypothetical example of coming up with the topic for a meta-analysis, but it sometimes happens - you read or hear some interesting questions and it sparks desire to put some numbers on some pattern...)*

Let's use this hypothetical topicfor the purpos of the exercise. We hope the topic is quite appealing and quiet easy to understand (we all get sick and most of us reproduce, or at least try to).

In this exercise, we will use a few basic tools to illustrate how to perform the first few steps of meta-analysis: refinig the question, searching and screenig literature (these are the same steps as in Systematic Reviews, so the tools are the same and can be useually described as systematic literature/evidence reviewing tools).  
Note that some R packeages (and many other online/software tools) are available (and more always coming) that can help with some of the tasks during this exercise, but there is no single "perfect" tool for the whole process usable for all disciplines and types of questions (althoght some attempts have been made). We simply do not have enough time to introduce more tools and the main purpose of this exercise is to familiarise you with the basic principles/issues that apply when you use any tools. You can always try alternative tools in your own time.

**Note:** There will be some **QUESTIONS** to answer and **TASKS** to be performed, as well as **RESOURCES**      
There is no *R* code to be run in this exercise.   
<br>

********************************************************************************
###  Formulating the question for a meta-analysis

Topic is usually broad and vague area of interest, for a meta-analysis we need a much more concrete question.
Generally, the more focused the question, the easier and quicker it is to perform the meta-analysis, but the answer becomes less general and there might be less evidence available.

**QUESTION 0:**  
What is your initial feeling about the meta-analytical question that can be derived from the text fragment cited above?

**ANSWER 0:**  
The most general question would be "What is the evidence for terminal investment happening?"    
But that is not a good question, because it is way too general.

**QUESTION 1:**  
What are at least 2 ways of improving this initial question to make it more amenable for a meta-analysis?

**ANSWER 1:**  
 1. Note that the listed examples involve challenge to immune system - this is becouse experimental manipulation can give clearest answer to whether termnal investment can be happening. So, the more precise  question would be "whether immune challenge can result in terminal investment"
 2. The examples given involve animals (birds), but plants and organisms from other kingdoms can also be used to answer such questions. However, they have quite dfferent immune systems and life histories, so it might be a good idea to restrict the question to animals, or even to a specific animal taxon (birds or insects). Note that in plant research different terminology can be used, e.g. overcompensation (there is a meta-analysis just published in Ecology on overcompensation in response to herbivory: https://www.ncbi.nlm.nih.gov/pubmed/30554427 , and a few older ones).

These question refinements are leading us gently into deciding which studies to use in a meta-analysis. However, before we move onto inclusion criteria, it is wise to gather more intelligence (Unless you are already an expert on the topic).

**QUESTION 2:**  
What should we do now?    
 a) Check if there already is a good recent meta-analysis on the topic.  
 b) Find and carefully read key reviews on the topic.  
 c) Find and carefully read a few representative empirical papers.  
 d) Further refine the question.  
 e) All of the above.

**ANSWER 2:**  
e) All of the above.   
Doing some scoping searches and some more reading will help to get to know theoretical background, make predictions and get a feel of what sort of empirical papers are available, and order of magnitude of the number of available papers. These searches can reveal mexisting meta-analyses on the topic, key narrative reviews and empirical papers, all of these will help refining the question and deciding on teh inclusion criteria for meta-analysis. We dont have time to do this all during this exercise, so we just do a quick scoping search for existing meta-analyses to familiarise you with GoogleScholar:

**TASK 1:**  
Are there any meta-analyses (or smilar) on terminal investment in animals?   (you can skip the following few tasks if short of time during the exercise)   
In GoogleScholar (https://scholar.google.com) run:  
*"meta-analytic" OR "meta-analysis" "terminal investment"*   
(you can paste this into search box or use Advanced search; see instructions and other GoogleScholar tips here: https://guides.library.ucsc.edu/c.php?g=745384&p=5361954 and https://scholar.google.com/intl/en/scholar/help.html).

You should get around 300 results. By skimming first few pages you can see a few meta-analyses on stress, diet, senescence, parasitism etc., but none probably specifically on terminal investment. There are many empirical papers coming up in this search because they mention meta-analysis in their body text or reference list - note that GoogleScholar searches full text by default. Also by default, the returned results are arranged in order of relevance to the search (whatever that means - its a complex algorithm that decides) and a few first pages contain some potentially relevant empirical papers.

**TASK 2:**   
You can restrict the search to titles only using:  
*allintitle: "meta analytic" OR "meta analysis" "terminal investment"*   
Now there are no matching hits; but there is no option for searching just in titles and abstracts.

Other things you should know about GoogleScholar - its search interface is rather simplistic and restrictive, it shows only up to 1,000 results for any particular search query, ranking algorithms are obscured, its hard to export the references, the literature coverage may fluctuate (its based on web scraping) and the searches may not be reproducible. Becouse of this it is not really recommended for the main searches in systematic reviews. On the upside, it can be helpful for additional searches and finding grey literature (not published in peer-review journals).

**RESOURCES:**   
Other software useful for exploratory (scoping) ad additional searches (check them up later, also note that they may have similar disadvantages to these of GoogleScholar):
 * SemanticScholar: https://www.semanticscholar.org/
 * Dimensions: https://www.dimensions.ai/
 * CiteHero: https://citehero.com/

**TASK 3:**   
Run another Google Scholar quick search:  
*"terminal investment" reproduction*   
This time you should get around 1900 records. Note that Goofgle Scholar does so called Automated Query Expansion for terms that are not set to be exact phrases (in quotes) - it uses alternative forms of these terms.

**TASK 4:**   
Try Google Scholar quick search:   
*"terminal investment" "immune challenge"*  
This time you should get around 400 records. Skim throug the first three pages of the results to get a feel of the diversity of the studies found (e.g. what taxa are present and what they might measure as reproductive investment) and note the wording of their titles.

**QUESTION 3:**  
In medical sciences, most meta-analyses (and systematic reviews) fit into the PICO (or related) question framework.

PICO stands for:   
  * P = Population   
  * I = Intervention  
  * C = Comparison/Control group   
  * O = Outcome

Formulate meta-analytic question "whether immune challenge can result in terminal investment" using PICO terms (actually this fits really well becouse we can use experimental data).

**ANSWER 3:**   
  * P = Population                = animal species   
  * I = Intervention              = immune challenge   
  * C = Comparison/Control group  = unchallenged group of animals, otherwiese in the same state   
  * O = Outcome                   = reproduction-related traits   
(further restrictions for all of these can ba added here or later at the criteria stage - we will add them later)

**RESOURCES:**   
An extensive list of alternative systematic review formulation question framweworks here:  https://www.networks.nhs.uk/nhs-networks/nwas-library-and-information-service/documents/alternative-question-structures-for-different-types-of-systematic-review    )

**Note:** you should always be able to justify your question as important, relevant and timely. 
<br>

********************************************************************************
###  Deciding on inclusion criteria

The basic PICO components are usually not sufficient to perform the effective screening of the papers for inclusion in meta-analysis.
There are additional both technical and biological refinements to be considered for inclusion criteria in our project.
Most importantly, you always need to be able to justify your inclusion criteria!

**QUESTION 4:**  
Timespan: should we include studies from any year?

**ANSWER 4:**   
Probably yes, unless you can come up with a reason for not doing so.

**QUESTION 5:**  
Language: should we include studies published in any language?

**ANSWER 5:**   
Most meta-analyses only include studies published in English by default. However, if possible it is recommend to include studies also in other languages (usually not many are found, unles you also perform your searches in other languages). 
**Note:** publication statsu can be also used as a criterium - are you happy to include some types of grey literature (e.g. conference proceedings, theses)?

**QUESTION 6:**  
Study type: should we include any type of study?

**ANSWER 6:**   
Since we investigate a specific intervention, we should restrict study type to experimental (we should explicitly exclude simulation or experimental studies kind of matching our question).

**QUESTION 7:**  
Taxa: should we include any type of taxa?   
**ANSWER 7:**   
Thats a difficult one. Animal kingdom is vast and diverse, and some species are especially odd (e.g. humans). Does it make sense to include single-cell animals? Does it make sense to include humans? Probably not.
How about multicellular animals, except humans? If many studies ascross multiple taxa exist, we could focus on the most frequently studied taxa or biologically most interesting. Alternatively just vertebrates, except humans? Another thing worth being specific about is that the animals should be wild-type (non-mutant, not disease models) and ideally not subject to any other significant manipulations that could affect their response (e.g. drug treatment; note there is a thin fuzzy line what these manipulations are). There might also be reasons to exclude standard laboratory species (mice, rat, yeast, fruitlfy) or maybe focus just on them? But, for now, we will stick to any multicellular animals, except humans.

**QUESTION 8:**   
Manipulation type: should we include any immune challenge?   
**ANSWER 8:**   
There are several established types of experimental immune challenge (blood cells (e.g. SRBC), protein (e.g. PHA), dead or live pathogens or parasites), so its worth including these and similar ones. Non-experimental immune challenge (e.g. natural infestation levels)  should not be included. If something odd comes up later it can come under "exclusion criteria", with justification.

**QUESTION 9:**   
Control gorup: should we accept any type of control group?  
**ANSWER 9:**   
Control group (unmanipulated) should consist of smimilar individuals, as far as reasonable and not be subject to any other maipulations.

**QUESTION 10:**   
Outcome type: should we accept any reproduction-related measurements and what that means?   
**ANSWER 10:**    
What is reproduction-related trait will depend on the species studied (may require additional reading or relying on the assertions of the authors). Different taxa will have different suites of reproduction related-traits, but these can be grouped into broader categories (e.g. mate attraction includes vocal signals, visual ornaments, pheromone production etc.; reproductive fitness will include number and quality of offspring) and different specific measurments will be available for different taxa and in different studies - this can be quite messy. If we do not have many studies on the topic and very specific question (e.g. "is the number of offspring affected?"") it is ok to be inclusive and later on try to figure out whether some traits are affected more than others.

**QUESTION 11:**  
Data type: should we accept any type of data for meta-analysis?   
**ANSWER 11:**   
It depends on the effect size we are planning to use. In our project it can be Cohens d (Hedges g) or lnRR, which are typically calcualted from mean, SD or SE and N (sample sizes). However, thay can be also inferred from some of the test statistics, or even proportions, if means and other descriptive statistics are not available. Other considerations include what to do if some data is missing or unclear.   
<br>

********************************************************************************
### Making decision trees

Good decision trees make screening the literature faster and more reliable.  
Briefly, if the initial (most general) questions, representing screening criteria, are not fulfilled, the study gets excluded and there is no need to evaluate the remaining criteria, so we save some time.

**QUESTION 12:**  
How many stages there are during literature screening process?   
**ANSWER 12:**   
The screening of literature is usually done in two stages:   
 1. abstract +  title + keywords  
 2. full text + supplementary information

**QUESTION 13:**  
How many decision trees do we need, and why, for the literature screening process?   
**ANSWER 13:**    
We need two decision trees! (one for each stage).  
This is because abstracts, titles and keywords do not contein enough information to conclusively answer all of the inclusion criteria questions - you should be excluding every paper for which you cannot answer "yes"" to all the questions. Thus this first stage of screening hes to be more tentative, and we would use a more permissive and simpler decision tree. A good tree starts from the questions that are easiest to answer (if we had any year or language restrictions we would put these on the top). Note thet during abstract screening we exclude the records where definete "no" can be inferred from available information, but we allow for some uncertainity for the "yes" answers by allowing "proabably" as an acceptable answer for inclusion, so we can cghheck for more info in the full text.

Example decision tree for abstract screening:

```{r diagram decision tree abstracts, fig.width=10, fig.height=6, echo=FALSE, message=FALSE}
par(mar=c(1,1,1,1))
openplotmat() #create an empyty plot
elpos <- coordinates (pos = c(2, 2, 2, 2, 2))
fromto <- matrix(ncol = 2, byrow = TRUE, data = c(1,2, 1,3, 3,4, 3,5, 5,6, 5,7, 7,8, 7,9))
nr <- nrow(fromto)
arrpos <- matrix(ncol = 2, nrow = nr)
for (i in 1:nr) (arrpos[i, ] <- straightarrow (to = elpos[fromto[i, 2], ], from = elpos[fromto[i, 1], ], lwd = 2, arr.pos = 0.7, arr.length = 0.4, arr.type = "simple"))

textrect (mid = elpos[1,], radx = 0.15, rady = 0.05, lwd = 1, lab = c("Study on non-human","multicellular animal?"), box.col = "white", shadow.col = "grey", shadow.size = 0, cex = 0.6)

textellipse (mid = elpos[2,], radx = 0.15, rady = 0.05, lwd = 1, lab = "EXCLUDE", box.col = "coral1", shadow.col = "grey", shadow.size = 0, cex = 0.8)

textrect (mid = elpos[3,], radx = 0.15, rady = 0.05, lwd = 1, lab = c("Experimental study?"), box.col = "white", shadow.col = "grey", shadow.size = 0, cex = 0.6)

textellipse (mid = elpos[4,], radx = 0.15, rady = 0.05, lwd = 1, lab = "EXCLUDE", box.col = "coral1", shadow.col = "grey", shadow.size = 0.001, cex = 0.8)

textrect (mid = elpos[5,], radx = 0.15, rady = 0.05, lwd = 1, lab = c("Experimental","immune challenge?"), box.col = "white", shadow.col = "grey", shadow.size = 0, cex = 0.6)

textellipse (mid = elpos[6,], radx = 0.15, rady = 0.05, lwd = 1, lab = "EXCLUDE", box.col = "coral1", shadow.col = "grey", shadow.size = 0, cex = 0.8)

textrect (mid = elpos[7,], radx = 0.15, rady = 0.05, lwd = 1, lab = c("Reproduction-related","traits measured?"), box.col = "white", shadow.col = "grey", shadow.size = 0, cex = 0.6)

textellipse (mid = elpos[8,], radx = 0.15, rady = 0.05, lwd = 1, lab = "EXCLUDE", box.col = "coral1", shadow.col = "grey", shadow.size = 0, cex = 0.8)

textellipse (mid = elpos[9,], radx = 0.15, rady = 0.05, lwd = 1, lab = c("INCLUDE", "for full-text screening"), box.col = "light green", shadow.col = "grey", shadow.size = 0, cex = 0.8)

text(arrpos[1, 1] - 0.1, arrpos[1, 2] + 0.02, "no")
text(arrpos[2, 1] + 0.07, arrpos[2, 2] + 0.03, "yes/probably")
text(arrpos[3, 1] - 0.1, arrpos[3, 2] + 0.02, "no")
text(arrpos[4, 1] + 0.07, arrpos[4, 2] + 0.03, "yes/probably")
text(arrpos[5, 1] - 0.1, arrpos[5, 2] + 0.02, "no")
text(arrpos[6, 1] + 0.07, arrpos[6, 2] + 0.03, "yes/probably")
text(arrpos[7, 1] - 0.1, arrpos[7, 2] + 0.02, "no")
text(arrpos[8, 1] + 0.07, arrpos[8, 2] + 0.03, "yes/probably")
```

**QUESTION 14:**  
Why in the abstract screening decision tree there is no question about the type of data used to express measurements reported in the study?

**ANSWER 14:**  
Most likely this will not be reported in the abstract, or only the measurments that were found to be significant will be mentioned This question needs to be answered during the full-text screening phase.

Full-text screening is more strict and we want to have answers to all our inclusion criteraia. In principle we only have "yes" and "no" answers to the screening questions. However, in practice, if a study looks very relevant but some information is ambiguous or missing, you may decide to contact the authors to get some extra information or data. in such case, you can provisionary include such papers and make final decision once you get, or not, the replies.

Example decision tree for full-text screening:

```{r diagram decision tree fulltexts, fig.width=10, fig.height=6, echo=FALSE, message=FALSE}
par(mar=c(1,1,1,1))
openplotmat() #create an empyty plot
elpos <- coordinates (pos = c(2, 2, 2, 2, 2, 2))
fromto <- matrix(ncol = 2, byrow = TRUE, data = c(1,2, 1,3, 3,4, 3,5, 5,6, 5,7, 7,8, 7,9, 9,10, 9,11 ))
nr <- nrow(fromto)
arrpos <- matrix(ncol = 2, nrow = nr)
for (i in 1:nr) (arrpos[i, ] <- straightarrow (to = elpos[fromto[i, 2], ], from = elpos[fromto[i, 1], ], lwd = 2, arr.pos = 0.7, arr.length = 0.4, arr.type = "simple"))

textrect (mid = elpos[1,], radx = 0.15, rady = 0.05, lwd = 1, lab = c("Study on non-human","multicellular animal?"), box.col = "white", shadow.col = "grey", shadow.size = 0, cex = 0.6)

textellipse (mid = elpos[2,], radx = 0.15, rady = 0.05, lwd = 1, lab = "EXCLUDE", box.col = "coral1", shadow.col = "grey", shadow.size = 0, cex = 0.8)

textrect (mid = elpos[3,], radx = 0.15, rady = 0.05, lwd = 1, lab = c("Experimental study?"), box.col = "white", shadow.col = "grey", shadow.size = 0, cex = 0.6)

textellipse (mid = elpos[4,], radx = 0.15, rady = 0.05, lwd = 1, lab = "EXCLUDE", box.col = "coral1", shadow.col = "grey", shadow.size = 0.001, cex = 0.8)

textrect (mid = elpos[5,], radx = 0.15, rady = 0.05, lwd = 1, lab = c("Experimental","immune challenge?"), box.col = "white", shadow.col = "grey", shadow.size = 0, cex = 0.6)

textellipse (mid = elpos[6,], radx = 0.15, rady = 0.05, lwd = 1, lab = "EXCLUDE", box.col = "coral1", shadow.col = "grey", shadow.size = 0, cex = 0.8)

textrect (mid = elpos[7,], radx = 0.15, rady = 0.05, lwd = 1, lab = c("Reproduction-related","traits measured?"), box.col = "white", shadow.col = "grey", shadow.size = 0, cex = 0.6)

textellipse (mid = elpos[8,], radx = 0.15, rady = 0.05, lwd = 1, lab = "EXCLUDE", box.col = "coral1", shadow.col = "grey", shadow.size = 0, cex = 0.8)

textrect (mid = elpos[9,], radx = 0.15, rady = 0.05, lwd = 1, lab = c("Reported means with SD (or SE)", "and N (sample sizes) for the treatment", "and control groups or test statistics", "from which effect sizes for the difference","between two groups can be calculated?"), box.col = "white", shadow.col = "grey", shadow.size = 0, cex = 0.6)

textellipse (mid = elpos[10,], radx = 0.15, rady = 0.05, lwd = 1, lab = "EXCLUDE", box.col = "coral1", shadow.col = "grey", shadow.size = 0, cex = 0.8)

textellipse (mid = elpos[11,], radx = 0.15, rady = 0.05, lwd = 1, lab = c("INCLUDE", "for meta-analysis"), box.col = "light green", shadow.col = "grey", shadow.size = 0, cex = 0.8)

text(arrpos[1, 1] - 0.1, arrpos[1, 2] + 0.02, "no")
text(arrpos[2, 1] + 0.05, arrpos[2, 2] + 0.03, "yes")
text(arrpos[3, 1] - 0.1, arrpos[3, 2] + 0.02, "no")
text(arrpos[4, 1] + 0.05, arrpos[4, 2] + 0.03, "yes")
text(arrpos[5, 1] - 0.1, arrpos[5, 2] + 0.02, "no")
text(arrpos[6, 1] + 0.05, arrpos[6, 2] + 0.03, "yes")
text(arrpos[7, 1] - 0.1, arrpos[7, 2] + 0.02, "no")
text(arrpos[8, 1] + 0.05, arrpos[8, 2] + 0.03, "yes")
text(arrpos[9, 1] - 0.1, arrpos[9, 2] + 0.02, "no")
text(arrpos[10, 1] + 0.05, arrpos[10, 2] + 0.03, "yes")
```

**QUESTION 15:**  
What question is missing on the top of the full-text decision tree?   
(Tip: its not about the Language or publication year)

**ANSWER 15:**    
The question is "Is the full text of this study available for our assessment?"   
It is often not explicitly stated, but without the full text it is impossible to extract all the necessary information about the study and the data. 
It is a good practice to keep this in mind and use as one of your potential inclusion/exclusion critera.   
<br>

********************************************************************************
###  Performing searches for relevant literature

Normally, we attempt to do a fairly comprehensive literature search (a full systematic review). The reason is that we want to have as many relevant data points as possible to increase the power of statistical analyses. The other reason is that we make effort to capture obscure studies that potentially are more likely to report non-significant results, and this way we actively try to reducte publication bias in our data set. Search strategy should include using at least two wide-coverage/interdisciplinary databases (we recommend Scopus and Web of Science), and a few additional sources of literature such as forward and backward reference searching (looking at cited and citing papers), publication lists of the key authors or labs, searching for unpublished studies and datasets (we will not cover these here).

In this exercise we only do a bit of the core search using 2 cdatabases:
Scopus and Web of Science.

**RESOURCES:** 
Scopus: https://www.scopus.com/   
Web of Science (WoS): http://apps.webofknowledge.com   
**Note:** WoS coverage depends of institutional subscription package purchased, so results of searches can vary among institutions.

**TASK 5:**   
Using Scopus **Advanced Search** run this search strings:  
 1. *TITLE-ABS-KEY ( "terminal investment"  AND  "immune challenge" )*    
(approx. 14 hits - search terms too restrictive)   
 2. *TITLE-ABS-KEY ( ( "terminal investment"  OR  "reproductive effort" )  AND  ( "immune challeng\*"  OR  "immunochalleng\*" ) )  *
(approx. 25 hits - a bit better, but still too restrictive)   
 3. *TITLE-ABS-KEY ( ( "terminal investment"  OR  "reproductive effort" )  AND  ( "immune challeng\*"  OR  "immunochalleng\*"  OR  "infect\*" ) )  *
(approx. 180 hits - much better number, but we probably now have included many non-experimental studies while still missing relevant ones)    
 4. *( TITLE-ABS-KEY ( ( "terminal investment"  OR  "reproductive effort" )  AND  ( "immune challeng\*"  OR  "immunochalleng\*"  OR  "infect\*" ) )  AND NOT  TITLE-ABS-KEY ( load  OR  human  OR  people ) )* 
(approx. 144 hits - by adding a group of exclusion tems (AND NOT) we probably got rid of some irrelevant ones).

Ideally, we shoult think and test some more terms for inclusion, after doing some scoping searches, reading reviews and sentinel relevant empirical papers. We stop for now and assume that the string above is our final searxh string. Note that Scopus Advanced Search stores all your searches ans search strings in a neat table at the bottom of the search window - its easy to copy this out and include in the documentation on how you developed your search strategy (same applies to WoS database). The \* symbol stands for wildcart and indicates "any symbols" since these databases do not do Automatic Query Expansion.

**TASK 6:**    
We can easily translate our final search string from Scopus into Web of Science (WoS) **Advanced Search**. Run this search string:    
*TS = (( ( "terminal investment" OR "reproductive effort" ) AND ( "immune challeng\*"  OR  "immunochalleng\*"  OR  "infect\*"  ) ) NOT ( load OR human OR people ) )*  
(approx. 249 hits - more than from Scopus).   WoS stores your search history and also wich sub-data bases were included in the search: Indexes=SCI-EXPANDED, SSCI, A&HCI, CPCI-S, CPCI-SSH, BKCI-S, BKCI-SSH, ESCI, CCR-EXPANDED, IC Timespan=All years" (you can restrit these in Advanced Search)

Both Scopus and WoS allow exporting the found records (with limitations on the max. number per export, **remember to tick the boxes for exporting abstracts!**) - there are multiple file formats you can use, but .ris and .bib are most commonly used ones  - you will need to figure out which ones work for your reference manager software. The next step would be downloading all found records from your "final" search into reference manager of your choice. Combine the references from two databasesinto a single project/folder and removing duplicates (most reference managers have such function, but it never works perfectly - do a visual check after sorting by titles!) before proceeding to screening stage.

**Note:** It is also a good time to write up your meta-analysis protocol and preregistering your meta-analysis (e.g. on OSF)!
<br>

********************************************************************************
###  Screening the literature

Theoretically you could do screening in teh database browser or your reference manager, but since it is recommended that 2 people do the screening independently (or at least part of the screening), it is good to use a dedicated tool to manage this process.

**RESOURCES:** 
Rayyan: https://rayyan.qcri.org/welcome (free, you can use it in the web browser or as a phone app; for more details see: http://libraryguides.mcgill.ca/rayyan/home)  
Abstrackr:  http://abstrackr.cebm.brown.edu/account/login (another free tool)

**Note:* it is possible to deduplication algorythms in these are not very good, so its advisable to deduplicate and check your records in the reference manager beforeuploading to the screening software.

**TASK 7:**  
We prepared a file (*Scopus10.ris*) with 10 references sampled from the set found using search string (these are sets of references from *( ( "terminal investment"  OR  "reproductive effort" )  AND  ( "immune challeng\*"  OR  "immunochalleng\*" ) )  *) in Scopus (same references came up in the equivalent search in WoS).

We will use these ten referneces to practice title and abstract screening in Rayyan. Screening can be also done in Abstracker, but we found Rayyan bit more relaible and more user-friendly.   
If you are warking in pairs or small groups, one person should create a new project in Rayyan (see below) and invite another person/people to the project using their email address.

 1. Open Rayyan (we assume you already created a free account) and log in.    
 2. My Reviews -> New REVIEW...   
 3. Enter project title, e.g. "Terminal Investment"" and description, e.g. "test"."   
 4. Select files -> upload *Scopus10.ris* and press Continue.     
It may take several minutes for larger files and slower connections, but eventually the uploaded references should appear in the left panel unde "Search methods" (you can upload more files with references via "Add new" link). Click on the "Uploaded References [scopus10.ris]" link in this panel to see the references in the right panel. Generally clicking on the links within panels on the left will filter the records in the right panels to include on higlithed sets (filtering, to get rid of the filter click on it again to remove hihlighting).  
In the left panels there are several sub-panels.  
 5. In the "Keywords for include"" you can add (via "Add new" link) folowing terms (or at least a few of these): challenge, challenged, injected, injection, infected, infection, PHA, SRBC, vaccine (these terms will be higlighted green in the titles and abstracts to help them find more easily; Rayyan tries to populate this list by itself - but its algorythm is medical-centered so you should adjust this list for your projects).   
 6. "Keywords for exclude" - " set to: review, reviewed, model, human, people, women, men (these terms will be higlighted red in the titles and abstracts to help them find more easily; Rayyan tries to populate this list by itself - but its algorythm is medical-centered so you should adjust this list for your projects).    
 7. Click on "All reviews" button in the top right corner to go to the project lists, then open your=project again by chosing it from the list. A view with a pie chart (your screening progress) and a few buttons will appear. If you need to invite other prople to do screening in parallel with you, do that by clicking on the "Invite" button. Stick with "Collaborator" selection and enter the email and message (e.g. "help me!"). Opne your project again - you should see your collaborators added with some info on their work so far. Press on "Show" button to see the references view.  
 8. Click on the first review from the top - an abstract and other record info will appear in the bottom panel.  
 9. Using pour abstract screening decision tree decide whether include this paper for the next stage of screening and press "Include" or "Exclude" button accordingly. And do the assessment for the remaining records. As you go you can create and assign new labels to individual records, e.g.: "unsure/check". You could also enter reasons for exclusion, e.g. "human" (we normally dont do this as there are often hundreds or thousands esxcluded records per project at this stage).  
10. Go to "All reviews" and check if othe rpeople finished their screening, if applicable. After that you can lift the "blind" mode, export all the records with decisions into a .csv files. For collaborative screens rayyan will now provide filters to disply all consensus included, consensus excluded and records with conflicting decisions. If you have any "conflicts" have a look again and discuss why you couldnt agree on a given record , come to agreement and make final decision. Export the records with final consensus decisions, for documentation.

**QUESTION 15:** 
How many papers you excluded? How many conflicts you got if you collaboratively screened (and if so what was cousing these conflicts, can the decision tree be improved?)

**ANSWER 15:**   
We think 4 should have been excluded: Simmons2009 and Clancy 2013 as human studies, Gershman2010 and Gonzales-Tokman 2013 as not likely to have any reprodaction-related measurements.

**TASK 8:**  
Its time to do the full-text screening for the papers that passed totle and abstract screening. There are some full-text papers that passed the intial screening in the directory - assess Nielsen2012 and Ardia2005 using our full-text screening decision tree.

**QUESTION 16:**  
What was your decision about Nielsen2012 and Ardia2005 papers?

**ANSWER 16:**    
Nielsen2012 - include, Ardia2005 - exclude due to the lack of good control group.  
<br>

********************************************************************************
###  PRISMA and other documentation

It is very important to keep detailed record of the whole searching, screening and data extraction process. Searching and screening are usually represented as some variant of PRISMA diagram.
With our 144 records from Scopus, 249 from Wos (these two databases are our "main search"), and imaginary around 400 records from other searches we might have done if we had more time, we would proabably end up with something like this one:

```{r diagram decision tree PRISMA, fig.width=10, fig.height=6, echo=FALSE, message=FALSE}
par(mar=c(1,1,1,1))
openplotmat() #create an empyty plot
elpos <- coordinates (pos = c(3, 3, 4, 3, 4, 3, 4, 3, 4, 4))
#text(elpos) #temporary label element positions with numbers

elpos[19,] <- elpos[19,] + c(0.02,0) #shift positions for the "EXCLUDED" element
elpos[26, ] <- elpos[26,] + c(0.02,0) #shift positions for the "EXCLUDED" element

#draw arrows
fromto1 <- matrix(ncol = 2, byrow = TRUE, data = c(1,8, 2,8, 8,15, 15,22, 22,29, 29,33, 3,13, 13,20, 20,27, 27,33, 15,19, 22,26)) #first set of arrows
arrpos1 <- matrix(ncol = 2, nrow = nrow(fromto1)) #set arrow psitions
for (i in 1:nrow(fromto1)) (arrpos1[i, ] <- straightarrow (to = elpos[fromto1[i, 2], ], from = elpos[fromto1[i, 1], ], lwd = 2, arr.pos = 0.8, arr.length = 0.3, arr.type = "simple")) #draw arrows

fromto2 <- matrix(ncol = 2, byrow = TRUE, data = c(13,19, 27,26)) #second set of arrows
arrpos2 <- matrix(ncol = 2, nrow = nrow(fromto2)) #set arrow psitions
for (i in 1:nrow(fromto2)) (arrpos2[i, ] <- straightarrow (to = elpos[fromto2[i, 2], ], from = elpos[fromto2[i, 1], ], lwd = 2, arr.pos = 0.65, arr.length = 0.3, arr.type = "simple")) #draw arrows

#draw rectangles and ellipses
textrect (mid = elpos[1,], radx = 0.15, rady = 0.04, lwd = 1, lab = c("SCOPUS","N = 144"), box.col = "white", shadow.col = "grey", shadow.size = 0, cex = 0.8)

textrect (mid = elpos[2,], radx = 0.15, rady = 0.04, lwd = 1, lab = c("Web of Science", "N = 249"), box.col = "white", shadow.col = "grey", shadow.size = 0, cex = 0.8)

textrect (mid = elpos[3,], radx = 0.15, rady = 0.04, lwd = 1, lab = c("other sources", "N = approx. 400"), box.col = "white", shadow.col = "grey", shadow.size = 0, cex = 0.8)

textrect (mid = elpos[8,], radx = 0.15, rady = 0.04, lwd = 1, lab = c("After duplicates removed", "N = 310"), box.col = "white", shadow.col = "grey", shadow.size = 0, cex = 0.8)

textrect (mid = elpos[15,], radx = 0.15, rady = 0.04, lwd = 1, lab = c("Deduplicated records screened", "N = 310"), box.col = "white", shadow.col = "grey", shadow.size = 0, cex = 0.8)

textrect (mid = elpos[22,], radx = 0.15, rady = 0.04, lwd = 1, lab = c("Full-text screened", "N = 50"), box.col = "white", shadow.col = "grey", shadow.size = 0, cex = 0.8)

textrect (mid = elpos[13,], radx = 0.15, rady = 0.04, lwd = 1, lab = c("Records screened", "N = approx. 400"), box.col = "white", shadow.col = "grey", shadow.size = 0, cex = 0.8)

textrect (mid = elpos[20,], radx = 0.15, rady = 0.04, lwd = 1, lab = c("Dedupicated against main search", "N = 10"), box.col = "white", shadow.col = "grey", shadow.size = 0, cex = 0.8)

textrect (mid = elpos[27,], radx = 0.15, rady = 0.04, lwd = 1, lab = c("Full-text screened", "N = 10"), box.col = "white", shadow.col = "grey", shadow.size = 0, cex = 0.8)

textellipse (mid = elpos[33,], radx = 0.15, rady = 0.05, lwd = 1, lab = c("INCLUDED", "in meta-analysis", "N = 39"), box.col = "light green", shadow.col = "grey", shadow.size = 0, cex = 0.8)

textellipse (mid = elpos[19,] + c(0.05,0), radx = 0.06, rady = 0.02, lwd = 1, lab = "EXCLUDED", box.col = "coral1", shadow.col = "grey", shadow.size = 0, cex = 0.8)

textellipse (mid = elpos[26,]+ c(0.05,0), radx = 0.06, rady = 0.02, lwd = 1, lab = "EXCLUDED", box.col = "coral1", shadow.col = "grey", shadow.size = 0, cex = 0.8)

#put numbers on exclusion arrows
text(elpos[19, 1] - 0.07, elpos[19, 2] + 0.02, "260")
text(elpos[26, 1] - 0.07, elpos[26, 2] + 0.02, "15")
text(elpos[19, 1] + 0.12, elpos[19, 2] + 0.12, "approx.380")
text(elpos[26, 1] + 0.14, elpos[26, 2] + 0.02, "6")
text(elpos[29, 1] + 0.02, elpos[29, 2] + 0.06, "35")
text(elpos[30, 1] + 0, elpos[30, 2] - 0.02, "4")
#text(arrpos[3, 1] - 0.1, arrpos[3, 2] + 0.02, "no")
```
<br>

**QUESTION 17:**   
How many papers in total were excluded during the full-text screening?

**ANSWER 17:**    
15+6 = 21

**Note** When you exclude records in the full-text screen keep recore for each indicvidual paper on the main reason for exclusion. Later present this information as a supplementary table for your meta-analyses. (Table of included studies should go in the main text, if possible).

Can you think of any other good questions for this section?

### 2019-01-24 Papers_search.Rmd
after spellcheck

@@
 You decided to work from home to get some rest a get better more quickly.
-You dod not feel energetic enough to crunch numbers or write, so you thought you may catch up on scientific reading instead.
+You don't not feel energetic enough to crunch numbers or write, so you thought you may catch up on scientific reading instead.
 You pick up the book sitting on top of the "to-read"" pile and open it at a random place.
@@
 
-Let's use this hypothetical topicfor the purpos of the exercise. We hope the topic is quite appealing and quiet easy to understand (we all get sick and most of us reproduce, or at least try to).
+Let's use this hypothetical topic for the purpose of the exercise. We hope the topic is quite appealing and quiet easy to understand (we all get sick and most of us reproduce, or at least try to).
 
-In this exercise, we will use a few basic tools to illustrate how to perform the first few steps of meta-analysis: refinig the question, searching and screenig literature (these are the same steps as in Systematic Reviews, so the tools are the same and can be useually described as systematic literature/evidence reviewing tools).
-Note that some R packeages (and many other online/software tools) are available (and more always coming) that can help with some of the tasks during this exercise, but there is no single "perfect" tool for the whole process usable for all disciplines and types of questions (althoght some attempts have been made).
-We simply do not have enough time to introduce more tools and the main purpose of this exercise is to familiarise you with the basic principles/issues that apply when you use any tools.
+In this exercise, we will use a few basic tools to illustrate how to perform the first few steps of meta-analysis: refining the question, searching and screening literature (these are the same steps as in Systematic Reviews, so the tools are the same and can be usually described as systematic literature/evidence reviewing tools).
+Note that some R packages (and many other online/software tools) are available (and more always coming) that can help with some of the tasks during this exercise, but there is no single "perfect" tool for the whole process usable for all disciplines and types of questions (although some attempts have been made).
+We simply do not have enough time to introduce more tools and the main purpose of this exercise is to familiarize you with the basic principles/issues that apply when you use any tools.
 You can always try alternative tools in your own time.
@@
  1.
-Note that the listed examples involve challenge to immune system - this is becouse experimental manipulation can give clearest answer to whether termnal investment can be happening.
+Note that the listed examples involve challenge to immune system - this is because experimental manipulation can give clearest answer to whether terminal investment can be happening.
 So, the more precise  question would be "whether immune challenge can result in terminal investment"
@@
 The examples given involve animals (birds), but plants and organisms from other kingdoms can also be used to answer such questions.
-However, they have quite dfferent immune systems and life histories, so it might be a good idea to restrict the question to animals, or even to a specific animal taxon (birds or insects).
+However, they have quite different immune systems and life histories, so it might be a good idea to restrict the question to animals, or even to a specific animal taxon (birds or insects).
 Note that in plant research different terminology can be used, e.g.
@@
 Doing some scoping searches and some more reading will help to get to know theoretical background, make predictions and get a feel of what sort of empirical papers are available, and order of magnitude of the number of available papers.
-These searches can reveal mexisting meta-analyses on the topic, key narrative reviews and empirical papers, all of these will help refining the question and deciding on teh inclusion criteria for meta-analysis.
-We dont have time to do this all during this exercise, so we just do a quick scoping search for existing meta-analyses to familiarise you with GoogleScholar:
+These searches can reveal existing meta-analyses on the topic, key narrative reviews and empirical papers, all of these will help refining the question and deciding on the inclusion criteria for meta-analysis.
+We don't have time to do this all during this exercise, so we just do a quick scoping search for existing meta-analyses to familiarize you with Google Scholar:
 
 **TASK 1:**  
-Are there any meta-analyses (or smilar) on terminal investment in animals?   (you can skip the following few tasks if short of time during the exercise)   
-In GoogleScholar (https://scholar.google.com) run:  
+Are there any meta-analyses (or similar) on terminal investment in animals?   (you can skip the following few tasks if short of time during the exercise)   
+In Google Scholar (https://scholar.google.com) run:  
 *"meta-analytic" OR "meta-analysis" "terminal investment"*   
-(you can paste this into search box or use Advanced search; see instructions and other GoogleScholar tips here: https://guides.library.ucsc.edu/c.php?g=745384&p=5361954 and https://scholar.google.com/intl/en/scholar/help.html).
+(you can paste this into search box or use Advanced search; see instructions and other Google Scholar tips here: https://guides.library.ucsc.edu/c.php?g=745384&p=5361954 and https://scholar.google.com/intl/en/scholar/help.html).
 
@@
 By skimming first few pages you can see a few meta-analyses on stress, diet, senescence, parasitism etc., but none probably specifically on terminal investment.
-There are many empirical papers coming up in this search because they mention meta-analysis in their body text or reference list - note that GoogleScholar searches full text by default.
+There are many empirical papers coming up in this search because they mention meta-analysis in their body text or reference list - note that Google Scholar searches full text by default.
 Also by default, the returned results are arranged in order of relevance to the search (whatever that means - its a complex algorithm that decides) and a few first pages contain some potentially relevant empirical papers.
@@
 
-Other things you should know about GoogleScholar - its search interface is rather simplistic and restrictive, it shows only up to 1,000 results for any particular search query, ranking algorithms are obscured, its hard to export the references, the literature coverage may fluctuate (its based on web scraping) and the searches may not be reproducible.
-Becouse of this it is not really recommended for the main searches in systematic reviews.
+Other things you should know about Google Scholar - its search interface is rather simplistic and restrictive, it shows only up to 1,000 results for any particular search query, ranking algorithms are obscured, its hard to export the references, the literature coverage may fluctuate (its based on web scraping) and the searches may not be reproducible.
+Because of this it is not really recommended for the main searches in systematic reviews.
 On the upside, it can be helpful for additional searches and finding grey literature (not published in peer-review journals).
@@
 **RESOURCES:**   
-Other software useful for exploratory (scoping) ad additional searches (check them up later, also note that they may have similar disadvantages to these of GoogleScholar):
- * SemanticScholar: https://www.semanticscholar.org/
+Other software useful for exploratory (scoping) ad additional searches (check them up later, also note that they may have similar disadvantages to these of Google Scholar):
+ * Semantic Scholar: https://www.semanticscholar.org/
  * Dimensions: https://www.dimensions.ai/
@@
 This time you should get around 1900 records.
-Note that Goofgle Scholar does so called Automated Query Expansion for terms that are not set to be exact phrases (in quotes) - it uses alternative forms of these terms.
+Note that Google Scholar does so called Automated Query Expansion for terms that are not set to be exact phrases (in quotes) - it uses alternative forms of these terms.
 
@@
 This time you should get around 400 records.
-Skim throug the first three pages of the results to get a feel of the diversity of the studies found (e.g.
+Skim through the first three pages of the results to get a feel of the diversity of the studies found (e.g.
 what taxa are present and what they might measure as reproductive investment) and note the wording of their titles.
@@
 
-Formulate meta-analytic question "whether immune challenge can result in terminal investment" using PICO terms (actually this fits really well becouse we can use experimental data).
+Formulate meta-analytic question "whether immune challenge can result in terminal investment" using PICO terms (actually this fits really well because we can use experimental data).
 
@@
   * I = Intervention              = immune challenge   
-  * C = Comparison/Control group  = unchallenged group of animals, otherwiese in the same state   
+  * C = Comparison/Control group  = unchallenged group of animals, otherwise in the same state   
   * O = Outcome                   = reproduction-related traits   
-(further restrictions for all of these can ba added here or later at the criteria stage - we will add them later)
+(further restrictions for all of these can be added here or later at the criteria stage - we will add them later)
 
 **RESOURCES:**   
-An extensive list of alternative systematic review formulation question framweworks here:  https://www.networks.nhs.uk/nhs-networks/nwas-library-and-information-service/documents/alternative-question-structures-for-different-types-of-systematic-review    )
+An extensive list of alternative systematic review formulation question frameworks here:  https://www.networks.nhs.uk/nhs-networks/nwas-library-and-information-service/documents/alternative-question-structures-for-different-types-of-systematic-review    )
 
@@
 **QUESTION 4:**  
-Timespan: should we include studies from any year?
+Time-span: should we include studies from any year?
 
@@
 Most meta-analyses only include studies published in English by default.
-However, if possible it is recommend to include studies also in other languages (usually not many are found, unles you also perform your searches in other languages).
-**Note:** publication statsu can be also used as a criterium - are you happy to include some types of grey literature (e.g. conference proceedings, theses)?
+However, if possible it is recommend to include studies also in other languages (usually not many are found, unless you also perform your searches in other languages).
+**Note:** publication status can be also used as a criterium - are you happy to include some types of grey literature (e.g. conference proceedings, theses)?
 
@@
 **ANSWER 7:**   
-Thats a difficult one.
+That is a difficult one.
 Animal kingdom is vast and diverse, and some species are especially odd (e.g.
@@
 How about multicellular animals, except humans?
-If many studies ascross multiple taxa exist, we could focus on the most frequently studied taxa or biologically most interesting.
+If many studies across multiple taxa exist, we could focus on the most frequently studied taxa or biologically most interesting.
 Alternatively just vertebrates, except humans?
@@
 drug treatment; note there is a thin fuzzy line what these manipulations are).
-There might also be reasons to exclude standard laboratory species (mice, rat, yeast, fruitlfy) or maybe focus just on them?
+There might also be reasons to exclude standard laboratory species (mice, rat, yeast, fruit fly) or maybe focus just on them?
 But, for now, we will stick to any multicellular animals, except humans.
@@
 **QUESTION 9:**   
-Control gorup: should we accept any type of control group?  
+Control group: should we accept any type of control group?  
 **ANSWER 9:**   
-Control group (unmanipulated) should consist of smimilar individuals, as far as reasonable and not be subject to any other maipulations.
+Control group (unmanipulated) should consist of similar individuals, as far as reasonable and not be subject to any other manipulations.
 
@@
 Different taxa will have different suites of reproduction related-traits, but these can be grouped into broader categories (e.g.
-mate attraction includes vocal signals, visual ornaments, pheromone production etc.; reproductive fitness will include number and quality of offspring) and different specific measurments will be available for different taxa and in different studies - this can be quite messy.
+mate attraction includes vocal signals, visual ornaments, pheromone production etc.; reproductive fitness will include number and quality of offspring) and different specific measurements will be available for different taxa and in different studies - this can be quite messy.
 If we do not have many studies on the topic and very specific question (e.g.
-"is the number of offspring affected?"") it is ok to be inclusive and later on try to figure out whether some traits are affected more than others.
+"is the number of offspring affected?"") it is OK to be inclusive and later on try to figure out whether some traits are affected more than others.
 
@@
 It depends on the effect size we are planning to use.
-In our project it can be Cohens d (Hedges g) or lnRR, which are typically calcualted from mean, SD or SE and N (sample sizes).
-However, thay can be also inferred from some of the test statistics, or even proportions, if means and other descriptive statistics are not available.
+In our project it can be Cohen's d (Hedges g) or lnRR, which are typically calculated from mean, SD or SE and N (sample sizes).
+However, they can be also inferred from some of the test statistics, or even proportions, if means and other descriptive statistics are not available.
 Other considerations include what to do if some data is missing or unclear.
@@
 We need two decision trees! (one for each stage).  
-This is because abstracts, titles and keywords do not contein enough information to conclusively answer all of the inclusion criteria questions - you should be excluding every paper for which you cannot answer "yes"" to all the questions.
+This is because abstracts, titles and keywords do not contain enough information to conclusively answer all of the inclusion criteria questions - you should be excluding every paper for which you cannot answer "yes"" to all the questions.
 Thus this first stage of screening hes to be more tentative, and we would use a more permissive and simpler decision tree.
 A good tree starts from the questions that are easiest to answer (if we had any year or language restrictions we would put these on the top).
-Note thet during abstract screening we exclude the records where definete "no" can be inferred from available information, but we allow for some uncertainity for the "yes" answers by allowing "proabably" as an acceptable answer for inclusion, so we can cghheck for more info in the full text.
+Note that during abstract screening we exclude the records where definite "no" can be inferred from available information, but we allow for some uncertainty for the "yes" answers by allowing "probably" as an acceptable answer for inclusion, so we can check for more info in the full text.
 
@@
 **ANSWER 14:**  
-Most likely this will not be reported in the abstract, or only the measurments that were found to be significant will be mentioned This question needs to be answered during the full-text screening phase.
+Most likely this will not be reported in the abstract, or only the measurements that were found to be significant will be mentioned This question needs to be answered during the full-text screening phase.
 
-Full-text screening is more strict and we want to have answers to all our inclusion criteraia.
+Full-text screening is more strict and we want to have answers to all our inclusion criteria.
 In principle we only have "yes" and "no" answers to the screening questions.
@@
 The reason is that we want to have as many relevant data points as possible to increase the power of statistical analyses.
-The other reason is that we make effort to capture obscure studies that potentially are more likely to report non-significant results, and this way we actively try to reducte publication bias in our data set.
-Search strategy should include using at least two wide-coverage/interdisciplinary databases (we recommend Scopus and Web of Science), and a few additional sources of literature such as forward and backward reference searching (looking at cited and citing papers), publication lists of the key authors or labs, searching for unpublished studies and datasets (we will not cover these here).
+The other reason is that we make effort to capture obscure studies that potentially are more likely to report non-significant results, and this way we actively try to reduce publication bias in our data set.
+Search strategy should include using at least two wide-coverage/interdisciplinary databases (we recommend Scopus and Web of Science), and a few additional sources of literature such as forward and backward reference searching (looking at cited and citing papers), publication lists of the key authors or labs, searching for unpublished studies and data sets (we will not cover these here).
 
-In this exercise we only do a bit of the core search using 2 cdatabases:
+In this exercise we only do a bit of the core search using 2 databases:
 Scopus and Web of Science.
@@
 *( TITLE-ABS-KEY ( ( "terminal investment"  OR  "reproductive effort" )  AND  ( "immune challeng\*"  OR  "immunochalleng\*"  OR  "infect\*" ) )  AND NOT  TITLE-ABS-KEY ( load  OR  human  OR  people ) )*
-(approx. 144 hits - by adding a group of exclusion tems (AND NOT) we probably got rid of some irrelevant ones).
+(approx. 144 hits - by adding a group of exclusion terms (AND NOT) we probably got rid of some irrelevant ones).
 
-Ideally, we shoult think and test some more terms for inclusion, after doing some scoping searches, reading reviews and sentinel relevant empirical papers.
-We stop for now and assume that the string above is our final searxh string.
+Ideally, we should think and test some more terms for inclusion, after doing some scoping searches, reading reviews and sentinel relevant empirical papers.
+We stop for now and assume that the string above is our final search string.
 Note that Scopus Advanced Search stores all your searches ans search strings in a neat table at the bottom of the search window - its easy to copy this out and include in the documentation on how you developed your search strategy (same applies to WoS database).
-The \* symbol stands for wildcart and indicates "any symbols" since these databases do not do Automatic Query Expansion.
+The \* symbol stands for wildcard and indicates "any symbols" since these databases do not do Automatic Query Expansion.
 
@@
 249 hits - more than from Scopus).
-WoS stores your search history and also wich sub-data bases were included in the search: Indexes=SCI-EXPANDED, SSCI, A&HCI, CPCI-S, CPCI-SSH, BKCI-S, BKCI-SSH, ESCI, CCR-EXPANDED, IC Timespan=All years" (you can restrit these in Advanced Search)
+WoS stores your search history and also which sub-data bases were included in the search: Indexes=SCI-EXPANDED, SSCI, A&HCI, CPCI-S, CPCI-SSH, BKCI-S, BKCI-SSH, ESCI, CCR-EXPANDED, IC Timespan=All years" (you can restrict these in Advanced Search)
 
@@
 The next step would be downloading all found records from your "final" search into reference manager of your choice.
-Combine the references from two databasesinto a single project/folder and removing duplicates (most reference managers have such function, but it never works perfectly - do a visual check after sorting by titles!) before proceeding to screening stage.
+Combine the references from two databases into a single project/folder and removing duplicates (most reference managers have such function, but it never works perfectly - do a visual check after sorting by titles!) before proceeding to screening stage.
 
@@
 
-Theoretically you could do screening in teh database browser or your reference manager, but since it is recommended that 2 people do the screening independently (or at least part of the screening), it is good to use a dedicated tool to manage this process.
+Theoretically you could do screening in the database browser or your reference manager, but since it is recommended that 2 people do the screening independently (or at least part of the screening), it is good to use a dedicated tool to manage this process.
 
@@
 
-**Note:* it is possible to deduplication algorythms in these are not very good, so its advisable to deduplicate and check your records in the reference manager beforeuploading to the screening software.
+**Note:* it is possible to deduplication algorithms in these are not very good, so its advisable to deduplicate and check your records in the reference manager before uploading to the screening software.
 
@@
 
-We will use these ten referneces to practice title and abstract screening in Rayyan. Screening can be also done in Abstracker, but we found Rayyan bit more relaible and more user-friendly.   
-If you are warking in pairs or small groups, one person should create a new project in Rayyan (see below) and invite another person/people to the project using their email address.
+We will use these ten references to practice title and abstract screening in Rayyan. Screening can be also done in Abstracker, but we found Rayyan bit more reliable and more user-friendly.   
+If you are working in pairs or small groups, one person should create a new project in Rayyan (see below) and invite another person/people to the project using their email address.
 
@@
  4. Select files -> upload *Scopus10.ris* and press Continue.     
-It may take several minutes for larger files and slower connections, but eventually the uploaded references should appear in the left panel unde "Search methods" (you can upload more files with references via "Add new" link).
+It may take several minutes for larger files and slower connections, but eventually the uploaded references should appear in the left panel under "Search methods" (you can upload more files with references via "Add new" link).
 Click on the "Uploaded References [scopus10.ris]" link in this panel to see the references in the right panel.
-Generally clicking on the links within panels on the left will filter the records in the right panels to include on higlithed sets (filtering, to get rid of the filter click on it again to remove hihlighting).
+Generally clicking on the links within panels on the left will filter the records in the right panels to include on highlighted sets (filtering, to get rid of the filter click on it again to remove highlighting).
 In the left panels there are several sub-panels.  
  5.
-In the "Keywords for include"" you can add (via "Add new" link) folowing terms (or at least a few of these): challenge, challenged, injected, injection, infected, infection, PHA, SRBC, vaccine (these terms will be higlighted green in the titles and abstracts to help them find more easily; Rayyan tries to populate this list by itself - but its algorythm is medical-centered so you should adjust this list for your projects).
+In the "Keywords for include"" you can add (via "Add new" link) following terms (or at least a few of these): challenge, challenged, injected, injection, infected, infection, PHA, SRBC, vaccine (these terms will be highlighted green in the titles and abstracts to help them find more easily; Rayyan tries to populate this list by itself - but its algorithm is medical-centered so you should adjust this list for your projects).
  6.
-"Keywords for exclude" - " set to: review, reviewed, model, human, people, women, men (these terms will be higlighted red in the titles and abstracts to help them find more easily; Rayyan tries to populate this list by itself - but its algorythm is medical-centered so you should adjust this list for your projects).
+"Keywords for exclude" - " set to: review, reviewed, model, human, people, women, men (these terms will be highlighted red in the titles and abstracts to help them find more easily; Rayyan tries to populate this list by itself - but its algorithm is medical-centered so you should adjust this list for your projects).
  7.
-Click on "All reviews" button in the top right corner to go to the project lists, then open your=project again by chosing it from the list.
+Click on "All reviews" button in the top right corner to go to the project lists, then open your=project again by choosing it from the list.
 A view with a pie chart (your screening progress) and a few buttons will appear.
-If you need to invite other prople to do screening in parallel with you, do that by clicking on the "Invite" button.
+If you need to invite other people to do screening in parallel with you, do that by clicking on the "Invite" button.
 Stick with "Collaborator" selection and enter the email and message (e.g.
 "help me!").
-Opne your project again - you should see your collaborators added with some info on their work so far.
+Open your project again - you should see your collaborators added with some info on their work so far.
 Press on "Show" button to see the references view.
@@
 You could also enter reasons for exclusion, e.g.
-"human" (we normally dont do this as there are often hundreds or thousands esxcluded records per project at this stage).
+"human" (we normally don't do this as there are often hundreds or thousands excluded records per project at this stage).
 10.
-Go to "All reviews" and check if othe rpeople finished their screening, if applicable.
+Go to "All reviews" and check if other people finished their screening, if applicable.
 After that you can lift the "blind" mode, export all the records with decisions into a .csv files.
-For collaborative screens rayyan will now provide filters to disply all consensus included, consensus excluded and records with conflicting decisions.
-If you have any "conflicts" have a look again and discuss why you couldnt agree on a given record , come to agreement and make final decision.
+For collaborative screens Rayyan will now provide filters to display all consensus included, consensus excluded and records with conflicting decisions.
+If you have any "conflicts" have a look again and discuss why you couldn't agree on a given record , come to agreement and make final decision.
 Export the records with final consensus decisions, for documentation.
@@
 **QUESTION 15:** 
-How many papers you excluded? How many conflicts you got if you collaboratively screened (and if so what was cousing these conflicts, can the decision tree be improved?)
+How many papers you excluded? How many conflicts you got if you collaboratively screened (and if so what was causing these conflicts, can the decision tree be improved?)
 
 **ANSWER 15:**   
-We think 4 should have been excluded: Simmons2009 and Clancy 2013 as human studies, Gershman2010 and Gonzales-Tokman 2013 as not likely to have any reprodaction-related measurements.
+We think 4 should have been excluded: Simmons2009 and Clancy 2013 as human studies, Gershman2010 and Gonzales-Tokman 2013 as not likely to have any reproduction-related measurements.
 
 **TASK 8:**  
-Its time to do the full-text screening for the papers that passed totle and abstract screening.
-There are some full-text papers that passed the intial screening in the directory - assess Nielsen2012 and Ardia2005 using our full-text screening decision tree.
+Its time to do the full-text screening for the papers that passed title and abstract screening.
+There are some full-text papers that passed the initial screening in the directory - assess Nielsen2012 and Ardia2005 using our full-text screening decision tree.
 
@@
 It is very important to keep detailed record of the whole searching, screening and data extraction process. Searching and screening are usually represented as some variant of PRISMA diagram.
-With our 144 records from Scopus, 249 from Wos (these two databases are our "main search"), and imaginary around 400 records from other searches we might have done if we had more time, we would proabably end up with something like this one:
+With our 144 records from Scopus, 249 from WoS (these two databases are our "main search"), and imaginary around 400 records from other searches we might have done if we had more time, we would probably end up with something like this one:
 
@@
 
-**Note** When you exclude records in the full-text screen keep recore for each indicvidual paper on the main reason for exclusion.
+**Note:** When you exclude records in the full-text screen keep record for each individual paper on the main reason for exclusion.
 Later present this information as a supplementary table for your meta-analyses.
@@
 
-Can you think of any other good questions for this section?
+*Can you think of any other good questions for this section?   *

### 2019-01-24 Trees_making.Rmd
after spellcheck and minor corretions

--- 2019-01-16 Trees_making.Rmd
@@
 
-```{r setup, cache = F, echo=FALSE}
+```{r setup, cache = F, echo=FALSE, results = FALSE}
 knitr::opts_chunk$set(error = TRUE) #allow some execution errors for demonstration purposes
 knitr::opts_chunk$set(eval = TRUE, echo = TRUE, warning = FALSE)
+sessionInfo()
 ```
@@
 The general flow of tree-making work is shown in the diagram below.  
-You will usually start with species list from the mata-analytic data set (use binomial latin names for the species list and in your database).
+You will usually start with species list from the meta-analytic data set (use binomial Latin names for the species list and in your database).
 First, check if there is any phylogenetic tree available that already contains all your species (e.g.
-a supertree).
-Currently, the easiest way to build a custom phylogenetic tree usable for meta-analysis is by using an R package called *rotl*, which accesses a synthetic supertree from Open Tree of Life database (https://opentree.wikispaces.com/).
+a super-tree).
+Currently, the easiest way to build a custom phylogenetic tree usable for meta-analysis is by using an R package called *rotl*, which accesses a synthetic super-tree from Open Tree of Life database (https://opentree.wikispaces.com/).
 We will use this approach in our examples below.
@@
 There are other super-trees available, e.g.
-for birds - we will use one of thes as our example too.
-Overall, if you can find a tree that contains all your specis, you just trim (subset) it down to your species.
+for birds - we will use one of these as our example too.
+Overall, if you can find a tree that contains all your species, you just trim (subset) it down to your species.
 If a few species is missing, it is sometimes possible to substitute them with closely related species.
@@
  - If the species are very diverged (e.g.
-mixture of vertabrates and invertebrates), use NCBI Taxonomy Common Tree tool (http://www.ncbi.nlm.nih.gov/Taxonomy/CommonTree/wwwcmt.cgi).
-At the webpage, use an input box at the top of the page to search each species name or upload lst of the species names from a file.
+mixture of vertebrates and invertebrates), use NCBI Taxonomy Common Tree tool (http://www.ncbi.nlm.nih.gov/Taxonomy/CommonTree/wwwcmt.cgi).
+At the webpage, use an input box at the top of the page to search each species name or upload list of the species names from a file.
 This allows to generate a scaffold for a tree.
@@
 If the resulting tree contains polytomies, these can be resolved at random, or based on information from more detailed trees (e.g.
-published phylogeneies of specific lower taxa).
+published phylogenies of specific lower taxa).
 You can also graft missing species on your tree.
@@
 ```{r prepare, message = FALSE, echo = TRUE, eval = FALSE, warning = FALSE}
+install.packages("metafor")
 install.packages("diagram")
@@
 
-We will give you some examples on how to use *rotl* package to download sub-trees from the Open Tree of Life supertree, resolve polytomies, trim and plot trees and deal with some of the most common issues.
-We will try to do most of theses tasks in R, but doing many of these outside R (using websites, text and tree ditors, etc.) is also possible.
+We will give you some examples on how to use *rotl* package to download sub-trees from the Open Tree of Life super-tree, resolve polytomies, trim and plot trees and deal with some of the most common issues.
+We will try to do most of theses tasks in R, but doing many of these outside R (using websites, text and tree editors, etc.) is also possible.
 We hope you get the idea of the main principles of the process.
@@
 **Important**: use Latin names, not common names.   
-**Notes**: *rotl* crashes on some species lists. Also, Open Tree of Life database gets constantly updated with new info, so you can get different outputs at different timepoints.
+**Notes**: *rotl* crashes on some species lists. Also, Open Tree of Life database gets constantly updated with new info, so you can get different outputs at different time-points.
 
@@
 Note that two of the species names in our list have typos, and are labelled TRUE in a column "approximate_match". Still, they were matched to correct species names.  
-If typos are detected, it is worth correcting the names in the original dataset and re-running tree search.
+If typos are detected, it is worth correcting the names in the original data set and re-running tree search.
 
@@
 
-```{r rotl 10species tree, warning = FALSE}
+```{r rotl 10species tree, warning = FALSE, results=FALSE}
 tree <- tol_induced_subtree(ott_ids = taxa[["ott_id"]], label_format = "name") #we can get a Warning meassage here about collapsing single nodes
@@
 Note that the species names have the Open Tree identifiers appended to them and *E.coli* was replaced with "mrcaott616ott617", which is non-taxon node within Bacteria.
-Given we dont have any other bacteria in the dataset, and the position on the tree is correct, thats fine, we just need to rename the tree tip back to "Escherichia coli".
+Given we dont have any other bacteria in the data set, and the position on the tree is correct, that's fine, we just need to rename the tree tip back to "Escherichia coli".
 
@@
 
-There are a few main existing and established supertrees for birds, so they can be used to create a subtree containing only the species from the list.  
-Here we will use on of these supertrees called "Ericson backbone", which is downloadable from http://birdtree.org/ (Jetz, W., G.
+There are a few main existing and established super-trees for birds, so they can be used to create a sub-tree containing only the species from the list.  
+Here we will use on of these super-trees called "Ericson backbone", which is downloadable from http://birdtree.org/ (Jetz, W., G.
 H.
@@
 Nature 491:444-448.).
-The website http://birdtree.org/ also has an online tool for subsetting their supertrees generating distribution of trees rather than one tree.
-For simplicity and demonstration puproses we will use R to subset (prune) a single bird supertree down to a smaller list of species.
+The website http://birdtree.org/ also has an online tool for subsetting their super-trees generating distribution of trees rather than one tree.
+For simplicity and demonstration purposes we will use R to subset (prune) a single bird super-tree down to a smaller list of species.
 
@@
 
-Prune supertree tree to a list of taxa from our list.
+Prune super-tree tree to a list of taxa from our list.
 
@@
 
-**Notes**: We were lucky that all our species had matches on the supertree.
+**Notes**: We were lucky that all our species had matches on the super-tree.
 Quite often there are no matches for some of the species names.
-In such cases, first check the names for typos or unnecessary characters, then try to find synonymous names and check if these are included in the supertree.
+In such cases, first check the names for typos or unnecessary characters, then try to find synonymous names and check if these are included in the super-tree.
 Finally, a sister species can be used when a species is truly missing.
-Remember to match the names on the final tree back to the original species list from the dataset.
+Remember to match the names on the final tree back to the original species list from the data set.
 
@@
 We will now handle a larger set of more closely related taxa, from a real data set.  
-The data set we use the dataset *dat.curtis1998* (contains studies on the Effects of Elevated CO2 Levels on Woody Plant Mass) included in the *metafor* package.
+The data set we use the data set *dat.curtis1998* (contains studies on the Effects of Elevated CO2 Levels on Woody Plant Mass) included in the *metafor* package.
 
@@
 
-If you display the names (taxa), note that some alternative names were found. We asssume these are correct ones for now.  
+If you display the names (taxa), note that some alternative names were found. We assume these are correct ones for now.  
 Now that the taxon names are matched to the Open Tree identifiers, we can pass them to the function tol_induced_subtree to retrieve the relationships among these taxa.
@@
 
-Its not binbary -  polytomies are present within genera: Acer, Betula, Quercus and Fraxinus.
+Its not binary -  polytomies are present within genera: Acer, Betula, Quercus and Fraxinus.
 We will resolve them at random.
@@
 
-1. Resolving one or more multifurcations in all possible ways: http://blog.phytools.org/2016/08/resolving-one-or-more-multifurcations.html
+1. Resolving one or more multifurcations (i.e. polytomies) in all possible ways: http://blog.phytools.org/2016/08/resolving-one-or-more-multifurcations.html
 
@@
 
-3. Resolving manually by useing information from known more specialised published phylogeny without polytomies (see next exercise for how to try to find these).
+3. Resolving manually by using information from known more specialized published phylogeny without polytomies (see next exercise for how to try to find these).
 
@@
 So, just in case we also provide the full tree files downloaded from there, which can be used in this exercise.
-In R, we first try to find a tree via Treebase website search - go to https://treebase.org/treebase-web/search/studySearch.html and enter Acer in teh search box and select "Abstract" from the drop-down menu next to it, press "Search" button.
-There are 4 relevant studies which potentially contain relevant trees - we will use the S342 Ackerly D., Donoghue M.J.	1998, so check the box next to it and then "Discard unchecked items" button at the bottom.
-Then, go to the "Trees"" tab where you will se 2 trees from this study.
-We will use Tr418 - you can download a Nexus file by clicking the icons on the right.
-Not: you can start the search from Tree tab, by searching tree titles and then look at the studies from which these trees are, or even search for the most obscure species in the species tab - sometimes these approaches work better.
+To see more easily what is in the Treebase database, we first try to find a tree via Treebase website search - you can go to https://treebase.org/treebase-web/search/studySearch.html and enter "Acer" in the search box, then select "Title" from the drop-down menu next to it, press "Search" button.
 
-Now we will try to do similar work from R. First, we create a custom search string with our four Acer species names from Curtis1998 dataset.
+There are 57 studies, but most of them were selected because they contain word "spaACER" (we are a bit unlucky with the taxon name).
+There are a few study titles that mention Acer which potentially may contain relevant trees and species: S2155, S2157 and S342.
+You can tick boxes next to them and then press button "Discard unchecked items" at the bottom of the window.
+Now, if you select just one of the remaining 3 studies and then go to the Taxa tab, you will see the list of taxa associated with a given study.
+
+In this exercise we will use study S342 Ackerly D., Donoghue M.J.	1998, so check the box next to it and then "Discard unchecked items" button at the bottom.
+Then, go to the "Trees"" tab where you will see 2 trees from this study.
+We pick Tr418 (consensus, i.e.
+made from collapsing multiple possible similar single trees).
+You can download it as a Nexus file by clicking the icons on the right.
+Note: you can also start the search from Tree tab, by searching tree titles and then look at the studies from which these trees are, or even start the search for the most obscure species in the Species tab - sometimes these approaches may work better.
+
+Now we will try to do similar work from R. First, we create a custom search string with our four Acer species names from Curtis1998 data set.
 
@@
 
-Run the treebase search - get amx 50 trees that contain exactly these four species names.
+Run the treebase search - get max 50 trees that contain exactly these four species names.
 
@@
 There are multiple trees available from several studies. Note that the lost of found results includes the tree 418 from study 342 we already identified in our Treebase website search.  
-So, we will request specifically tree Tr418 Study S342 - Ackerly D., Donoghue M.J.	1998	Leaf size, sapling allometry, and Corner's rules: phylogeny and correlated evolution in maples (Acer).	American Naturalist.
+So, we will request specifically tree Tr418 from Study S342 (the one we already found via Treebase website search; Ackerly D., Donoghue M.J.	1998	Leaf size, sapling allometry, and Corner's rules: phylogeny and correlated evolution in maples (Acer).	American Naturalist).
 It may take a moment, depending on the size of the tree.
-You can try requesting different trees later.
+You can try requesting different trees later and comparing whether you get different sub-trees after pruning.
 
-```{r dwonload single tree from treebase}
+```{r dwonload one tree from treebase, results=FALSE}
 Acers_tree <- search_treebase("418", by="id.tree", exact_match = TRUE) #returns a single tree in a list
@@
 If downloading from R treebase did not work, upload tree saved in a file:
-```{r dwonload single tree from file, eval = FALSE}
+```{r dwonload one tree from file, eval = FALSE}
 Acers_tree <- read.nexus(file = "T418.nex") #read Nexus tree file
@@
 There is much more to learn, especially dealing with missing or rare species names, grafting trees together, resolving polytomies, plotting cool trees etc.
-In some cases getting "perfect" tree may not be possible - but phylogenies are always just estimats of the species relationships, and as long as you get the key features right, the tree will be ok, and so your analyses (in most cases).
+In some cases getting "perfect" tree may not be possible - but phylogenies are always just estimates of the species relationships, and as long as you get the key features right, the tree will be OK, and so your analyses (in most cases).
 

### 2019-01-24 Papers_search.Rmd
added Q18 and some comments and links

@@
 **Note:** WoS coverage depends of institutional subscription package purchased, so results of searches can vary among institutions.
+Also searching just core collection vs.
+all available databases can make a difference.
 
 **TASK 5:**   
-Using Scopus **Advanced Search** run this search strings:  
+Using Scopus **Advanced Search** (you can read more about the syntax here: https://dev.elsevier.com/tips/ScopusSearchTips.htm) run this search strings:  
  1. *TITLE-ABS-KEY ( "terminal investment"  AND  "immune challenge" )*    
@@
 We stop for now and assume that the string above is our final search string.
-Note that Scopus Advanced Search stores all your searches ans search strings in a neat table at the bottom of the search window - its easy to copy this out and include in the documentation on how you developed your search strategy (same applies to WoS database).
+Note that Scopus Advanced Search stores all your searches ans search strings in a neat table at the bottom of the search window - its easy to copy this out and include in the documentation on how you developed your search strategy (same applies to WoS database; you need to register in Scopus and WoS to save searches and alerts across multiple search sessions).
 The \* symbol stands for wildcard and indicates "any symbols" since these databases do not do Automatic Query Expansion.
@@
 **TASK 6:**    
-We can easily translate our final search string from Scopus into Web of Science (WoS) **Advanced Search**. Run this search string:    
+We can easily translate our final search string from Scopus into Web of Science (WoS) **Advanced Search** (you can read more about the syntax here: https://images.webofknowledge.com/images/help/WOS/hs_search_operators.html).
+Run this search string:
 *TS = (( ( "terminal investment" OR "reproductive effort" ) AND ( "immune challeng\*"  OR  "immunochalleng\*"  OR  "infect\*"  ) ) NOT ( load OR human OR people ) )*  
@@
 Both Scopus and WoS allow exporting the found records (with limitations on the max.
-number per export, **remember to tick the boxes for exporting abstracts!**) - there are multiple file formats you can use, but .ris and .bib are most commonly used ones  - you will need to figure out which ones work for your reference manager software.
+number per export, **remember to tick the boxes for exporting abstracts!**) - there are multiple file formats you can use, but .ris and .bib are most commonly used ones- you will need to figure out which ones work for your reference manager software.
 The next step would be downloading all found records from your "final" search into reference manager of your choice.
 Combine the references from two databases into a single project/folder and removing duplicates (most reference managers have such function, but it never works perfectly - do a visual check after sorting by titles!) before proceeding to screening stage.
+An R package *revtools* (https://cran.rstudio.com/web/packages/revtools/ and http://revtools.net/) can act as a reference manager and also has a good de-dupiction function.
 
@@
 We prepared a file (*Scopus10.ris*) with 10 references sampled from the set found using search string (these are sets of references from *( ( "terminal investment"  OR  "reproductive effort" )  AND  ( "immune challeng\*"  OR  "immunochalleng\*" ) )  *) in Scopus (same references came up in the equivalent search in WoS).
+
+**Note:** Sometimes the biblio-files with the same extension differ a bit in structure and reference managers (and Rayyan) may have problem opening them.
+Getting your files into exactly right format can sometimes be frustrating.
 
@@
 There are some full-text papers that passed the initial screening in the directory - assess Nielsen2012 and Ardia2005 using our full-text screening decision tree.
+Hint: dont read the whole paper - just try to skim it and look for the key information.
 
@@
 
-*Can you think of any other good questions for this section?   *
+**QUESTION 18:**  
+Above you can see a PRISMA (flow) diagram. The other important document(s) associated with PRISMA is a checklist (http://prisma-statement.org/documents/PRISMA%202009%20checklist.pdf). 
+PRISMA checklist is a document that helps making sure that you reported the key information how a systematic review (including meta-analysis) was done.    
+Have a look at the PRISMA checklist. Find and read the items that are related to what we have covered in this exercise. What are the numbers (in the # column) of these items?
+
+**Note:** this PRISMA was designed for meedical systematic reviews, and some items are not very relevant or useful. Also, it is due to be updated.
+
+**ANSWER 18:**    
+Items: 3 to 9, 17
+
+*We hope you enjoyed this exercise! Remeber, we just roughly covered the basics and there is much more to learn!*

### 2019-01-24 Papers_search.Rmd
question answers moved to "solutions" file

@@
 
-**ANSWER 0:**  
-The most general question would be "What is the evidence for terminal investment happening?"    
-But that is not a good question, because it is way too general.
-
 **QUESTION 1:**  
 What are at least 2 ways of improving this initial question to make it more amenable for a meta-analysis?
-
-**ANSWER 1:**  
- 1.
-Note that the listed examples involve challenge to immune system - this is because experimental manipulation can give clearest answer to whether terminal investment can be happening.
-So, the more precise  question would be "whether immune challenge can result in terminal investment"
- 2.
-The examples given involve animals (birds), but plants and organisms from other kingdoms can also be used to answer such questions.
-However, they have quite different immune systems and life histories, so it might be a good idea to restrict the question to animals, or even to a specific animal taxon (birds or insects).
-Note that in plant research different terminology can be used, e.g.
-overcompensation (there is a meta-analysis just published in Ecology on overcompensation in response to herbivory: https://www.ncbi.nlm.nih.gov/pubmed/30554427 , and a few older ones).
 
@@
  e) All of the above.
-
-**ANSWER 2:**  
-e) All of the above.   
-Doing some scoping searches and some more reading will help to get to know theoretical background, make predictions and get a feel of what sort of empirical papers are available, and order of magnitude of the number of available papers.
-These searches can reveal existing meta-analyses on the topic, key narrative reviews and empirical papers, all of these will help refining the question and deciding on the inclusion criteria for meta-analysis.
-We don't have time to do this all during this exercise, so we just do a quick scoping search for existing meta-analyses to familiarize you with Google Scholar:
 
@@
 
-**ANSWER 3:**   
-  * P = Population                = animal species   
-  * I = Intervention              = immune challenge   
-  * C = Comparison/Control group  = unchallenged group of animals, otherwise in the same state   
-  * O = Outcome                   = reproduction-related traits   
-(further restrictions for all of these can be added here or later at the criteria stage - we will add them later)
-
 **RESOURCES:**   
@@
 
-**ANSWER 4:**   
-Probably yes, unless you can come up with a reason for not doing so.
-
 **QUESTION 5:**  
@@
 
-**ANSWER 5:**   
-Most meta-analyses only include studies published in English by default.
-However, if possible it is recommend to include studies also in other languages (usually not many are found, unless you also perform your searches in other languages).
 **Note:** publication status can be also used as a criterium - are you happy to include some types of grey literature (e.g. conference proceedings, theses)?
@@
 
-**ANSWER 6:**   
-Since we investigate a specific intervention, we should restrict study type to experimental (we should explicitly exclude simulation or experimental studies kind of matching our question).
-
 **QUESTION 7:**  
 Taxa: should we include any type of taxa?
-**ANSWER 7:**   
-That is a difficult one.
-Animal kingdom is vast and diverse, and some species are especially odd (e.g.
-humans).
-Does it make sense to include single-cell animals?
-Does it make sense to include humans?
-Probably not.
-How about multicellular animals, except humans?
-If many studies across multiple taxa exist, we could focus on the most frequently studied taxa or biologically most interesting.
-Alternatively just vertebrates, except humans?
-Another thing worth being specific about is that the animals should be wild-type (non-mutant, not disease models) and ideally not subject to any other significant manipulations that could affect their response (e.g.
-drug treatment; note there is a thin fuzzy line what these manipulations are).
-There might also be reasons to exclude standard laboratory species (mice, rat, yeast, fruit fly) or maybe focus just on them?
-But, for now, we will stick to any multicellular animals, except humans.
 
@@
 Manipulation type: should we include any immune challenge?
-**ANSWER 8:**   
-There are several established types of experimental immune challenge (blood cells (e.g.
-SRBC), protein (e.g.
-PHA), dead or live pathogens or parasites), so its worth including these and similar ones.
-Non-experimental immune challenge (e.g.
-natural infestation levels)  should not be included.
-If something odd comes up later it can come under "exclusion criteria", with justification.
 
@@
 Control group: should we accept any type of control group?
-**ANSWER 9:**   
-Control group (unmanipulated) should consist of similar individuals, as far as reasonable and not be subject to any other manipulations.
 
@@
 Outcome type: should we accept any reproduction-related measurements and what that means?
-**ANSWER 10:**    
-What is reproduction-related trait will depend on the species studied (may require additional reading or relying on the assertions of the authors).
-Different taxa will have different suites of reproduction related-traits, but these can be grouped into broader categories (e.g.
-mate attraction includes vocal signals, visual ornaments, pheromone production etc.; reproductive fitness will include number and quality of offspring) and different specific measurements will be available for different taxa and in different studies - this can be quite messy.
-If we do not have many studies on the topic and very specific question (e.g.
-"is the number of offspring affected?"") it is OK to be inclusive and later on try to figure out whether some traits are affected more than others.
 
@@
 Data type: should we accept any type of data for meta-analysis?   
-**ANSWER 11:**   
-It depends on the effect size we are planning to use.
-In our project it can be Cohen's d (Hedges g) or lnRR, which are typically calculated from mean, SD or SE and N (sample sizes).
-However, they can be also inferred from some of the test statistics, or even proportions, if means and other descriptive statistics are not available.
-Other considerations include what to do if some data is missing or unclear.
 <br>
@@
 How many stages there are during literature screening process?
-**ANSWER 12:**   
-The screening of literature is usually done in two stages:   
- 1. abstract +  title + keywords  
- 2. full text + supplementary information
 
@@
 How many decision trees do we need, and why, for the literature screening process?
-**ANSWER 13:**    
-We need two decision trees! (one for each stage).  
-This is because abstracts, titles and keywords do not contain enough information to conclusively answer all of the inclusion criteria questions - you should be excluding every paper for which you cannot answer "yes"" to all the questions.
-Thus this first stage of screening hes to be more tentative, and we would use a more permissive and simpler decision tree.
-A good tree starts from the questions that are easiest to answer (if we had any year or language restrictions we would put these on the top).
-Note that during abstract screening we exclude the records where definite "no" can be inferred from available information, but we allow for some uncertainty for the "yes" answers by allowing "probably" as an acceptable answer for inclusion, so we can check for more info in the full text.
 
-Example decision tree for abstract screening:
+An example decision tree for abstract screening:
 
@@
 
-**ANSWER 14:**  
-Most likely this will not be reported in the abstract, or only the measurements that were found to be significant will be mentioned This question needs to be answered during the full-text screening phase.
-
 Full-text screening is more strict and we want to have answers to all our inclusion criteria.
@@
 
-Example decision tree for full-text screening:
+An example decision tree for full-text screening:
 
@@
 (Tip: its not about the Language or publication year)   
-
-**ANSWER 15:**    
-The question is "Is the full text of this study available for our assessment?"   
-It is often not explicitly stated, but without the full text it is impossible to extract all the necessary information about the study and the data. 
-It is a good practice to keep this in mind and use as one of your potential inclusion/exclusion critera.   
 <br>
@@
 
-**QUESTION 15:** 
+**QUESTION 16:** 
 How many papers you excluded? How many conflicts you got if you collaboratively screened (and if so what was causing these conflicts, can the decision tree be improved?)
-
-**ANSWER 15:**   
-We think 4 should have been excluded: Simmons2009 and Clancy 2013 as human studies, Gershman2010 and Gonzales-Tokman 2013 as not likely to have any reproduction-related measurements.
 
@@
 
-**QUESTION 16:**  
+**QUESTION 17:**  
 What was your decision about Nielsen2012 and Ardia2005 papers?   
-
-**ANSWER 16:**    
-Nielsen2012 - include, Ardia2005 - exclude due to the lack of good control group.  
 <br>
@@
 
-**QUESTION 17:**   
+**QUESTION 18:**   
 How many papers in total were excluded during the full-text screening?
-
-**ANSWER 17:**    
-15+6 = 21
 
@@
 
-**QUESTION 18:**  
+**QUESTION 19:**  
 Above you can see a PRISMA (flow) diagram. The other important document(s) associated with PRISMA is a checklist (http://prisma-statement.org/documents/PRISMA%202009%20checklist.pdf). 
@@
 **Note:** this PRISMA was designed for meedical systematic reviews, and some items are not very relevant or useful. Also, it is due to be updated.   
-
-**ANSWER 18:**    
-Items: 3 to 9, 17
+<br>
 

### 2019-01-25 Papers_search_solutions.Rmd
Search and searching exercise solutions files

---
title: "Papers_search_solutions"
author: "ML"
date: "1/23/2019"
output: html_document
---

```{r setup, cache = F, echo=FALSE, results = FALSE}
knitr::opts_chunk$set(error = TRUE) #allow some execution errors for demonstration purposes
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, warning = FALSE, collapse = TRUE, comment = "#>")
sessionInfo()
```

# EXERCISES - SEARCHING AND SCREENING PAPERS FOR META-ANALYSIS  - SOLUTIONS

Prepared by Malgorzata (Losia) Lagisz, BEES, UNSW, AU, losialagisz@gmail.com.

<br>

********************************************************************************  
### Introduction

**ANSWER 0:**  
The most general question would be "What is the evidence for terminal investment happening?"    
But that is not a good question, because it is way too general.

**ANSWER 1:**  
 1. Note that the listed examples involve challenge to immune system - this is because experimental manipulation can give clearest answer to whether terminal investment can be happening. So, the more precise  question would be "whether immune challenge can result in terminal investment"
 2. The examples given involve animals (birds), but plants and organisms from other kingdoms can also be used to answer such questions. However, they have quite different immune systems and life histories, so it might be a good idea to restrict the question to animals, or even to a specific animal taxon (birds or insects). Note that in plant research different terminology can be used, e.g. overcompensation (there is a meta-analysis just published in Ecology on overcompensation in response to herbivory: https://www.ncbi.nlm.nih.gov/pubmed/30554427 , and a few older ones).

**ANSWER 2:**  
e) All of the above.   
Doing some scoping searches and some more reading will help to get to know theoretical background, make predictions and get a feel of what sort of empirical papers are available, and order of magnitude of the number of available papers. These searches can reveal existing meta-analyses on the topic, key narrative reviews and empirical papers, all of these will help refining the question and deciding on the inclusion criteria for meta-analysis. We don't have time to do this all during this exercise, so we just do a quick scoping search for existing meta-analyses to familiarize you with Google Scholar:

**ANSWER 3:**   
  * P = Population                = animal species   
  * I = Intervention              = immune challenge   
  * C = Comparison/Control group  = unchallenged group of animals, otherwise in the same state   
  * O = Outcome                   = reproduction-related traits   
(further restrictions for all of these can be added here or later at the criteria stage - we will add them later)

**ANSWER 4:**   
Probably yes, unless you can come up with a reason for not doing so.

**ANSWER 5:**   
Most meta-analyses only include studies published in English by default. However, if possible it is recommend to include studies also in other languages (usually not many are found, unless you also perform your searches in other languages).

**ANSWER 6:**   
Since we investigate a specific intervention, we should restrict study type to experimental (we should explicitly exclude simulation or experimental studies kind of matching our question).

**ANSWER 7:**   
That is a difficult one. Animal kingdom is vast and diverse, and some species are especially odd (e.g. humans). Does it make sense to include single-cell animals? Does it make sense to include humans? Probably not.
How about multicellular animals, except humans? If many studies across multiple taxa exist, we could focus on the most frequently studied taxa or biologically most interesting. Alternatively just vertebrates, except humans? Another thing worth being specific about is that the animals should be wild-type (non-mutant, not disease models) and ideally not subject to any other significant manipulations that could affect their response (e.g. drug treatment; note there is a thin fuzzy line what these manipulations are). There might also be reasons to exclude standard laboratory species (mice, rat, yeast, fruit fly) or maybe focus just on them? But, for now, we will stick to any multicellular animals, except humans.

**ANSWER 8:**   
There are several established types of experimental immune challenge (blood cells (e.g. SRBC), protein (e.g. PHA), dead or live pathogens or parasites), so its worth including these and similar ones. Non-experimental immune challenge (e.g. natural infestation levels)  should not be included. If something odd comes up later it can come under "exclusion criteria", with justification.

**ANSWER 9:**   
Control group (unmanipulated) should consist of similar individuals, as far as reasonable and not be subject to any other manipulations.

**ANSWER 10:**    
What is reproduction-related trait will depend on the species studied (may require additional reading or relying on the assertions of the authors). Different taxa will have different suites of reproduction related-traits, but these can be grouped into broader categories (e.g. mate attraction includes vocal signals, visual ornaments, pheromone production etc.; reproductive fitness will include number and quality of offspring) and different specific measurements will be available for different taxa and in different studies - this can be quite messy. If we do not have many studies on the topic and very specific question (e.g. "is the number of offspring affected?"") it is OK to be inclusive and later on try to figure out whether some traits are affected more than others.

**ANSWER 11:**   
It depends on the effect size we are planning to use. In our project it can be Cohen's d (Hedges g) or lnRR, which are typically calculated from mean, SD or SE and N (sample sizes). However, they can be also inferred from some of the test statistics, or even proportions, if means and other descriptive statistics are not available. Other considerations include what to do if some data is missing or unclear.

**ANSWER 12:**   
The screening of literature is usually done in two stages:   
 1. abstract +  title + keywords  
 2. full text + supplementary information

**ANSWER 13:**    
We need two decision trees! (one for each stage).  
This is because abstracts, titles and keywords do not contain enough information to conclusively answer all of the inclusion criteria questions - you should be excluding every paper for which you cannot answer "yes"" to all the questions. Thus this first stage of screening hes to be more tentative, and we would use a more permissive and simpler decision tree. A good tree starts from the questions that are easiest to answer (if we had any year or language restrictions we would put these on the top). Note that during abstract screening we exclude the records where definite "no" can be inferred from available information, but we allow for some uncertainty for the "yes" answers by allowing "probably" as an acceptable answer for inclusion, so we can check for more info in the full text.

**ANSWER 14:**  
Most likely this will not be reported in the abstract, or only the measurements that were found to be significant will be mentioned This question needs to be answered during the full-text screening phase.

**ANSWER 15:**    
The question is "Is the full text of this study available for our assessment?"   
It is often not explicitly stated, but without the full text it is impossible to extract all the necessary information about the study and the data. 
It is a good practice to keep this in mind and use as one of your potential inclusion/exclusion critera.

**ANSWER 16:**   
We think 4 should have been excluded: Simmons2009 and Clancy 2013 as human studies, Gershman2010 and Gonzales-Tokman 2013 as not likely to have any reproduction-related measurements.

**ANSWER 17:**    
Nielsen2012 - include, Ardia2005 - exclude due to the lack of good control group.

**ANSWER 18:**    
15 + 6 = 21   (from the PRISMA diagram)

**ANSWER 19:**    
Items: 3 to 9, 17

<br>

*We hope it all makes sense - if not sure about some answers, please talk to us*

### 2019-01-28 Data_extractions.Rmd
added more tasks and questions (now 19)

--- 2019-01-10 Data_extractions.Rmd
@@
 author: "ML"
-date: "10 January 2019"
+date: "25 January 2019"
 output: html_document
@@
 
-```{r setup, cache = F, echo=FALSE}
+```{r setup, cache = F, echo=FALSE, results = FALSE}
 knitr::opts_chunk$set(error = TRUE) #allow some execution errors for demonstration purposes
 knitr::opts_chunk$set(eval = TRUE, echo = TRUE, warning = FALSE)
+sessionInfo()
 ```
 
-# EXERCISES - PHYLOGENETIC TREES FOR META-ANALYSIS
+# EXERCISE - DATA EXTRACTION AND ORGANISATION
 
-Prepared by Malgorzata (Losia) Lagisz, BEES, UNSW, AU, losialagisz@yahoo.com.
+Prepared by Malgorzata (Losia) Lagisz, BEES, UNSW, AU, losialagisz@gmail.com.
 
-Introduction  
--------
+**Setup:** Work in pairs or small groups. R/Rstudio and Internet connection needed.
 
-**TO DO**:
+### Purpose of the exercise:   
+In this exercise you get some practice and tips on:   
+ * Organising your data set  
+ * Extracting and coding   
+ * Converting different types of information   
+ * Documentation and reporting
 
-Prepare a set of 3 DAH papers to be extracted using *metaDigitise* package
-See: https://github.com/daniel1noble/metaDigitise (link to the paper?)
+<br>
+ ## Introduction
 
-USE ES_calculator to calculate ES?
+Imagine, you quickly got well and instead of reproducing decided to carry on the Terminal Investment Meta-analysis Project.   
+After a few weeks you fished searching and screening literature for the relevant papers.
+You now have a collection of around 30 included full-text papers (it could be 300 as well), stored neatly as pdf files and supplementary information, in whatever format available.
+Its time to start extracting the data (ideally, you should have planned this in detail while writing up your protocol/preregistration, but in this exercise we assume you did not).
 
-Start from installing (as required) and loading R packages.
+<br>
+********************************************************************************
 
-```{r prepare, message = FALSE}
-#install.packages("devtools","fulltxt","ape","metafor")
+###  Organising your data set
 
+Before you extract actual data for the meta-analysis you need some sort of system to manage data extractions and store the data (we assume by now you already have a system for organising and storing your references and papers, so we don't cover this).
+For data extractions, many people just stick to Excel spreadsheets, or any other spreadsheets or extraction forms that can be converted into spreadsheets (e.g.
+Google Forms).
+Some researchers even build elaborated relational databases (there is a high initial investment cost but many benefits later on).
+Whatever you chose, you need a platform where you can freely add and edit columns and be able to store different types of data, so rigid systems purpose-built for medical trial data are not recommended.
+In this exercise we will stick to the simplest, but flexible platform - Excel spreadsheet.
+
+**QUESTION 1:**  
+What would be a minimum set of columns needed in an Excel sheet for a meta-analysis?   
+(Remember, you need a separate column for each data/information type)
+
+**QUESTION 2:**   
+What other columns would be useful? Give examples of at least 5 other columns that could be added?
+
+**QUESTION 3:**   
+What is the purpose of extracting additional variables?
+
+**QUESTION 4:**   
+Whats wrong with this table representing a fragment of an extraction spreadsheet?
+
++---------------+--------------------+---------------+-------------+-------------+-------------+   
+| StudyID       | Species            | mean 1        | SD 1        | N 1         | P           |   
++===============+====================+===============+=============+=============+=============+   
+| Weil2006      | Phodopus sungorus  | 45.4          | 6.4      .  | 10          | 0.023       |   
++---------------+--------------------+---------------+-------------+-------------+-------------+   
+| Sadd2006      | mealworm beetle    | NA            |not reported | 8 - 9       | < 0.011     |   
++---------------+--------------------+---------------+-------------+-------------+-------------+
+
+**QUESTION 5:**    
+What is the use of the CohortID column in the extraction table below (just a fragment of a larger example extraction sheet)?
+
++----------+----------+------------------+-------+-------+-------------+------------+-----------+
+| StudyID  | CohortID |Species           | sex   | age   | mean_control| SD_control | N_control |
++==========+==========+==================+=======+=======+=============+============+===========+
+| One2019  | One2019F |Phodopus sungorus | F     | 60    | 3.6         | 1.3        | 10        |
++----------+----------+------------------+-------+-------+-------------+------------+-----------+
+| One2009  | One2019M |Phodopus sungorus | M     | 60    | 12.1        | 3.1        | 12        |
++----------+----------+------------------+-------+-------+-------------+------------+-----------+
+| One2019  | One2019F |Phodopus sungorus | F     | 120   | 8.3         | 2.2        | 10        |
++----------+----------+------------------+-------+-------+-------------+------------+-----------+
+| One2009  | One2019M |Phodopus sungorus | M     | 120   | 14.5        | 4.4        | 12        |
++----------+----------+------------------+-------+-------+-------------+------------+-----------+
+
+**QUESTION 6:**    
+What is the use of the CohortID column in the extraction table below (just a fragment of a larger example extraction sheet)?
+
++----------+----------+------------------+-------+-------+-------------+------------+-----------+
+| StudyID  | CohortID |Species           | trait | age   | mean_control| SD_control | N_control |
++==========+==========+==================+=======+=======+=============+============+===========+
+| Two2019  | Two2019a1|Phodopus sungorus | calls | 60    | 3.6         | 1.3        | 10        |
++----------+----------+------------------+-------+-------+-------------+------------+-----------+
+| Two2009  | Two2019a1|Phodopus sungorus | mass  | 60    | 12.1        | 3.1        | 12        |
++----------+----------+------------------+-------+-------+-------------+------------+-----------+
+| Two2019  | Two2019a2|Phodopus sungorus | calls | 120   | 8.3         | 2.2        | 10        | 
++----------+----------+------------------+-------+-------+-------------+------------+-----------+
+| Two2009  | Two2019a2|Phodopus sungorus | mass  | 120   | 14.5        | 4.4        | 12        |
++----------+----------+------------------+-------+-------+-------------+------------+-----------+
+
+**QUESTION 7:**    
+What is the use of the CohortID column in the extraction table below (just a fragment of a larger example extraction sheet)?
+
++----------+------------+------------------+-------+------+-------------+------------+-----------+
+| StudyID  | CohortID   |Species           | trait | age  | mean_control| SD_control | N_control |
++==========+============+==================+=======+======+=============+============+===========+
+| Three2019| Three2019s1|Phodopus sungorus | calls | 60   | 3.6         | 1.3        | 10        |
++----------+------------+------------------+-------+------+-------------+------------+-----------+
+| Three2019| Three2019s1|Phodopus sungorus | mass  | 60   | 12.1        | 3.1        | 12        |
++----------+------------+------------------+-------+------+-------------+------------+-----------+
+| Three2019| Three2019s2|Phodopus campbelli| calls | 60   | 8.3         | 2.2        | 10        |
++----------+------------+------------------+-------+------+-------------+------------+-----------+
+| Three2019| Three2019s2|Phodopus campbelli| mass  | 60   | 14.5        | 4.4        | 12        |
++----------+------------+------------------+-------+------+-------------+------------+-----------+
+
+**QUESTION 8:**    
+What is the use of the Control_grID column in the extraction table below (just a fragment of a larger example extraction sheet)?
+
++----------+----------+-------------+----------+-------------+----------+------------+---------+
+| StudyID  | CohortID | Control_grID| treatment| mean_control| N_control| mean_treatm| N_tratm |
++==========+==========+=============+==========+=============+==========+============+=========+
+| Four2019 | Four2019a| Four2019c1  | virus    | 4.8         | 20       | 21.1       | 14      |
++----------+----------+-------------+----------+-------------+----------+------------+---------+
+| Four2009 | Four2019b| Four2019c1  | bacteria | 4.8         | 20       | 11.3       | 16      |
++----------+----------+-------------+----------+-------------+----------+------------+---------+
+| Four2009 | Four2009a| Four2019c2  | virus    | 6.3         | 19       | 33.3       | 10      |
++----------+----------+-------------+----------+-------------+----------+------------+---------+
+| Four2009 | Four2009b| Four2019c2  | bacteria | 6.3         | 19       | 14.4       | 12      |
++----------+----------+-------------+----------+-------------+----------+------------+---------+
+
+**QUESTION 9:**   
+What is meta-data in meta-analysis and what it is used for?
+
+**Note:** In your actuall extraction sheet/form, try to group related columns together.
+You can also color-code them in Excel or put on the same tab/section in a visual database interface, if used.
+Your extraction form (spreadsheet) should be tested on several papers (pilot extractions) to make sure its easy to use and captures necessary information.
+
+<br>
+********************************************************************************
+
+###  Extracting and coding
+
+**QUESTION 10:**   
+Is one person sufficient to extract the data?
+
+**QUESTION 11:**   
+In what order would you extract your included studies? Random? Chronological? Any other?
+
+**QUESTION 12:**   
+When numerical data is reported in the text or tables, it is easy to get exact values for your extraction columns.
+However, often data is reported in the figures?
+Can we extract such data?
+Is it always possible?
+
+**TASK 1**:  
+We will practice data extraction on few of the Terminal investment papers. We will use R package *metaDigitise*. 
+Detailed package description and tutorial are available at: https://cran.r-project.org/web/packages/metaDigitise/vignettes/metaDigitise.html
+https://github.com/daniel1noble/metaDigitise and this publication:
+https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.13118
+
+Here is the package installation code:
+
+```{r prepare metadigitise, eval=FALSE, message = FALSE}
+install.packages("devtools")
+devtools::install_github("daniel1noble/metaDigitise")
+#install.packages("metaDigitise")
 ```
+
+```{r load packages, eval=TRUE, message = FALSE}
+library(metaDigitise)
+library(tidyverse)
+```
+
+Lets try to extract data from:
+ 1. Figure 2a in the paper by Weil et al. 2006: Immune challenge retards seasonal reproductive regression in rodents: evidence for terminal investment (Weil2006.pdf)
+ 2.
+Figure 2a in the paper by Sadd et al.
+2006: Modulation of sexual signalling by immune challenged male mealworm beetles (Tenebrio molitor, L.): evidence for terminal investment and dishonesty (Sadd2006.pdf)
+
+You will need to open the pdf files and take the snapshots of the relevant figures.
+Make sure you capture the axes and axes labels.
+Also, enlarge these figures to fill the screen before taking the snapshot to get higher resolution image.
+
+Name the files as following (this makes it easy to keep track of the figures being digitised):     
+001_Weil_2006_Fig2a.png    
+and    
+002_Sadd_2006_Fig2a.png
+
+We need to place them in a new folder - we called it metaDigitise_figures (read more about organising your extraction directories in the "Setting up directory structures " section of https://cran.r-project.org/web/packages/metaDigitise/vignettes/metaDigitise.html).
+
+To extract our two figures we need to provide the directory path name of the folder with the images to the metaDigise() function, for example:
+
+```{r path metadigitise, eval=FALSE, message = FALSE}
+data <- metaDigitise(dir = "~/Dropbox/Local_git_repos/MA_teaching/metaDigitise_figures/") #you will need to adjust this path for your folder
+```
+
+Tell metaDigitise() what you would like to do by selecting option  (Process new images - enter "1").
+
+The next question is whether the plots are different or same - this is about the plot type and since we have a box plot and bar plot with means and SE, we can treat them as the same plot types, so enter "s" and for the next question "m".
+
+First plot should be loaded at this stage. We do not need to rotate the image, so enter "c".
+
+Now we can store the information about y-variable type - enter "testes volume (mm3)".
+
+You get asked for the data needed to calibrate y-axis. Click on 200 and 900, then enter these values as prompted and complete calibration.
+
+Its good to enter sample sizes at this point.
+Since there is no such info in the plot or under the plot, search the full text for the sample sizes (hint: look in "Statistical anlaysis").
+Enter identifier for the control group ("saline") and the sample size.
+
+Since we have multiple data points for each group of animals, we have to decide whether extract all of them or just one (or some subset) - your decision should be biologically-informed and ideally a priori.
+For this exercise we will extract just the last one (at 6 weeks).
+Click on the outer edge of the error bar of the last black point, and then click the middle of the black dot.
+If it went well you can add the treatment group "LPS" in a similar manner.
+If something went wrong you can delete and repeat the extraction.
+
+Once you extract data for the treatment group, select "Finish plot" (enter "f"), select "se" and continue "y" to move to the next image.
+
+Now you should be able to extract by yourself plot 002_Sadd_2006_Fig2a.png! (hint: extract "treatment control" and "treatment" groups).
+
+Once you get "Congratulations" message, you can look at the extracted data. metaDigitise() stores and sorts all the summary data for each of the figures in the "data"" object:
+
+```{r see extractions metadigitise, eval=FALSE, message = FALSE}
+data
+```
+
+You can add more figures in the directory and add more data to this data table.
+Or you can save it and load back later to continue adding more data or merge with other data frames (see "Re-importing previously digitised data and accessing raw data" in the metaDigitise tutorial at https://cran.r-project.org/web/packages/metaDigitise/vignettes/metaDigitise.html).
+
+```{r save metadigitise, eval=FALSE, message = FALSE}
+write.csv(data, file = "filename.csv")
+```
+
+As you see, metaDigitise guides you gently through the all steps of data extraction from figures.
+It also nicely organises and stores the data for you.
+You should invest some extra time into learning this handy tool.
+If you have some time left after this exercise go through the full tutorial at https://cran.r-project.org/web/packages/metaDigitise/vignettes/metaDigitise.html
+
+<br>
+********************************************************************************
+
+###  Converting different types of information
+
+Since now we have a few data points extracted we could calculate effect sizes from them. We first need to reshape our data frame.
+
+**QUESTION 13:**   
+Why we need to change our data frame (and how)?
+
+**TASK 2**:
+Since we have only 2 data points per file, its easy to reshape the data frame in R.
+
+```{r reshape data, eval=FALSE, message = FALSE}
+data$group <- c("c","t") #assign rows to control or treatment codes (in this case its easy, but be careful)
+tdata <- filter(data, group == "t") #extract treatment data rows
+cdata <- select(filter(data, group == "c"), filename, group_id, mean, sd, n, se) #extract control data rows
+data_new <- full_join(cdata, tdata, by = "filename") #join data
+data_new <- select (data_new, -c(plot_type, error_type)) #remove unnecessary coumns
+data_new
+```
+
+Now the columns appended with .x contain control groups data values, and these with .y are for the treatment groups (you can rename the columns to make it easier to see).
+
+**TASK 3**:  
+We will now use *escalc* function from *metafor* package to calculate effect sizes.
+The function allows to calculate various effect sizes or outcome measures (and the corresponding sampling variances) that are commonly used in meta-analyses, including two effect sizes are relevant to our data and question:
+ * "SMD" stands for standarised mean difference, i.e. Hedges' g (potentially also "SMDH" for heteroscedastic variances).   
+ * "ROM" for the log transformed ratio of means (Hedges et al., 1999; Lajeunesse, 2011), also named lnRR.   
+We calculate them as follows:
+
+```{r calculate SMD , eval=FALSE, message = FALSE}
+SMD <- metafor::escalc(measure = "SMD", data=data_new, m1i=mean.y, m2i = mean.x, sd1i=sd.y, sd2i= sd.x, n1i = n.y, n2i=n.x, append = FALSE) #note we have treatemnt group as 1 and control as 2, so we can get possitive effect size for larger investment in treatment group
+SMD
+```
+
+```{r calculate ROM , eval=FALSE, message = FALSE}
+ROM <- metafor::escalc(measure = "ROM", data=data_new, m1i=mean.y, m2i = mean.x, sd1i=sd.y, sd2i= sd.x, n1i = n.y, n2i=n.x, append = FALSE)
+ROM
+```
+
+**QUESTION 14:**   
+What should you do next?
+
+<br>
+
+With these two papers we just extracted, we had a relatively easy job.
+There were descriptive statistics values reported in the plots.
+If we cannot get means and standard errors (or SD) from the text, tables and figures (or already-calculated effect sizes, which is still rare), we can look for test (inferential) statistics.
+Test statistics that are usable include t-values form t-tests, F-values, Chi-square values, finally, p-values.
+These values from simple statistical tests (like independent or paired t-tests) are easily convertible into SMD (d, g) or r (correlation coefficient).
+These conversions and some others can be calculated in R if you know the formulas (including conversion from SMD to correlation coefficient), but for the exercise purpose we will do this web-based Practical Meta-Analysis Effect Size Calculator:
+https://www.campbellcollaboration.org/research-resources/research-for-resources/effect-size-calculator.html
+
+```{r webEScalcualtor, echo=FALSE, fig.cap=" ", out.width='100%'}
+knitr::include_graphics("./images/webEScalcualtor.png")
+```
+
+Open the above webpage. Main menu has an expandable list of types of effect sizes, and statistical tests you can use to calculate these effect sizes (only d, r, OR and RR).
+
+**TASK 4**:   
+Weil 2006 reports in the results: "At the end of the experiment saline-injected animals had smaller testes (F 1,27 = 4.6875, p<0.05, figure 2b)" and in the methods it is noted these are calculated using ANOVA.
+Use online ES calculator sheet selecting "Standarized Mean difference (d)"" and "F-test, 2 groups" for to calculate d from one-way ANOVA (you will need F-value and sample sizes of treatment and control groups).
+
+You should get d for the testes  mass that is close to what we calculated for tests volume from Fig2a.
+Note that this web calculator provides confidence intervals (CI) and variance for the calculated effect sizes.
+
+**TASK 5**:  
+Using web-based ES calculator, estimate a few other effect sizes using test (inferential) statistics reported in Sadd2006.
+
+<br>
+**QUESTION 15:**   
+If both descriptive statistics (mean, SE or SD) and test (inferential) statistics are reported, which one should we extract from a paper?
+
+**QUESTION 16:**    
+Is the sign of effect size important? Imagine we use d, which represents mean_treatment - mean_control group trait values. How we would interpret d = 0.5 and how we would interpret d = -0.5?
+
+**QUESTION 17:**    
+If we have different traits in the data set (lets say number of eggs laid and probability of next abandonment) what is the problem with the  signs of effect sizes for these traits?
+What should we do about this?
+
+<br>
+********************************************************************************
+
+### Documentation and reporting
+
+**QUESTION 18:**   
+As you hopefully remember, PRISMA checklist is a document that helps making sure that you reported the key information how a systematic review (including meta-analysis) was done.    
+Have a look again at the PRISMA checklist (http://prisma-statement.org/documents/PRISMA%202009%20checklist.pdf).
+Find and read the items that are related to what we have covered in this exercise.
+What are the numbers (in the # column) of these items?
+
+**Note:** this PRISMA was designed for medical systematic reviews, and some items are not very relevant or useful. Also, it is due to be updated.   
+<br>
+
+**QUESTION 19:**   
+Although not mentioned in the original PRISMA checklist, it is important to share the collected data, meta-data and analysis code? Think of east at least 3 reasons for doing so.
+
+**ANSWER 19:**   
+Generally speaking, it is for transparency, reproducibility and giving back to the community.
+
+<br>
+
+*We hope you enjoyed this exercise! Remeber, we just roughly covered the basics and there is much more to learn!*

### 2019-01-28 Data_extractions_solutions.Rmd
answers to the exercise questions moved to a separate file

---
title: "Data_extractions_solutions"
author: "ML"
date: "28/01/2019"
output: html_document
---

```{r setup, cache = F, echo=FALSE, results = FALSE}
knitr::opts_chunk$set(error = TRUE) #allow some execution errors for demonstration purposes
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, warning = FALSE)
sessionInfo()
```

# EXERCISE - DATA EXTRACTION AND ORGANISATION - SOLUTIONS

Prepared by Malgorzata (Losia) Lagisz, BEES, UNSW, AU, losialagisz@gmail.com.

<br>
********************************************************************************

###  Organising your data set

**ANSWER 1:**   
Study ID (String of characters), effect size and its variance (if we want to run a simple weighted meta-analytic model; both numeric).

**ANSWER 2:**   
 1. More detailed information on each study: author, title, journal, publication year, e.g.:

+---------------+---------------+-------------+-------------+-------------+   
| StudyID       | Authors       | Title       | Journal     | Year        |   
+===============+===============+=============+=============+=============+   
| Weil2006      |  Weil, Martin | Immune ...  | Biol Lett   | 2006        |   
+---------------+---------------+-------------+-------------+-------------+   
| Sadd2006      | Sadd, Holman, | Modulation..| J Comp      | 2006        |   
+---------------+---------------+-------------+-------------+-------------+

 2. Information on the study system or species (e.g. species name, taxonomic group, biological details relevant to the studied question).

 3. Most papers will not report effect sizes but descriptive and test (inferential) statistics that need to be converted into effect sizes, we need to extract data from which effect sizes can be calculated. For Terminal Investment project, since we want to compare challenged and unchallenged groups, we will need to collect means and standard errors, sample sizes, or test statistic that can be converted into effect sizes. Since each data type needs its own column, that means several columns already, e.g.: mean_contr, SE_contr, SD_contr, N_contr, mean_treat, SE_treat, SD_treat, N_treat, test_stat_type, test_stat_value, df.

 4. Information on where in the original paper the above data was extracted from, e.g. Figure/Table number, page, raw data; if multiple comparisons were reported in the original paper, which exact groups were compared.

5. Information on study methodology - e.g., where and when manipulations were performed and measurements/data collected.

6. Any comments and additional information related to the extracted data and study. You can have a comment/details column associated with every extracted data column and an overall comment column.

**ANSWER 3:**   
Potentially, there can be a lot of information that could and should be extracted from the original papers and sometimes supplemented by information from other sources (databases, publications). Some of the information can be used as effect size moderators in the statistical models. Other information can be used to characterise the data set (e.g., is it dominated by particular taxon or experimental methodology?).

**ANSWER 4:**    
 1. Non-informative column names (which group should be nr 1 ?)   
 2. Second study has non-numerical values in the columns that should be numerical (NA is good for missing values, but should be used consistently)    
 3. Species column is a mix of Latin and common names (should be all Latin)

**ANSWER 5:**    
CohortID is used to indicate that the same cohorts/group of animals, of different sex (F, M), were measured repeatedly at different ages (60 and 120). These are not independent data points, and this is captured by CohortID.

**ANSWER 6:**    
CohortID is used to indicate that the same cohorts/group of animals were measured repeatedly for different traits (calls, mass) at different ages (60 and 120). These are not independent data points, and this is captured by CohortID.

**ANSWER 7:**    
CohortID is used to indicate that the same cohorts/group of animals representing two species were measured once for different traits (calls, mass). We can also use combination of StudyID and Species columns to take this into account during statistical analyses.

**ANSWER 8:**    
Control_grID is used to indicate that the same control group was used as a comparator to two different cohorts/group that were subject to different types of (could also be levels) of experimental treatment (virus, bacteria). It is clear from identical values of means and sample sizes for the control group but different ones for the treatment groups. The resulting effect sizes are not independent, but in a different way than if the same animals were measured for different traits or more than once for the same trait (rows 3 and 4). And this is captured by combination of CohortID and Control_grID and could be taken into account in statistical models.

**ANSWER 9:**    
Meta-data is a detailed description of all extracted variables/columns and all other variables/columns derived from them. It is a part of the documentation of your meta-analysis and it is necessary when you or somebody else checks the data or tries to re-analyse or re-use it. It is a good practice to start creating such documentation when you start extractions and coding and update it as you make changes and progress to analyses.

<br>
********************************************************************************

###  Extracting and coding

**ANSWER 10:**    
Ideally, two people would be extracting the data in parallel 9independently, in-duplicate), to minimise errors. However it is often not practical. In such case one person could be extracting and another person checking the extractions (there are always mistakes!).

**ANSWER 11:**    
It is often easier and quicker to group similar studies together, Especially grouping the studies by study/system and research group helps find overlapping studies (e.g. where the same experiment was reported in separate publications, but focusing on different aspects) or collect more details on how studies were done. Also leaving complicated and odd studies for later on might be a good idea.

**ANSWER 12:**    
There are several pieces of software that helps digitising various types of plots and extracting the data (you can find a list here: https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.13118), so it is often possible.  However, sometimes plot resolution can be too low, or data points / lines are overlapping and obscuring some of the values. In such case, you can try to contact the authors for the actual numbers or raw data used to make the plot.

<br>
********************************************************************************

###  Converting different types of information

**ANSWER 13:**   
Since we compare pairs of measurements for control and treatment groups, it is convenient to have these in the same row in the data frame.

**ANSWER 14:**   
Would be good to merge these values back to the data frame.

```{r merge ES, eval=FALSE, message = FALSE}
data_new <- bind_cols(data_new, SMD, ROM) #bind data
data_new
```

**ANSWER 15:**  
We usually give priority to descriptive statistics (mean, SE or SD), which usually represent raw data, unadjusted for other variables and also can be converted to different units or used to calculate alternative types of effect sizes.

**ANSWER 16:**   
The sign represents direction of the effect on a given trait. d = 0.5 means that the trait value in the treatment group is moderately larger than the trait value in the control group, d = -0.5 would mean that the trait value in the treatment group is moderately smaller than the trait value in the control group.

**ANSWER 17:**   
Larger numbers of eggs indicate increased reproductive effort (and thus reproductive investment), while increased probability to abandon the nest indicates reduced reproductive effort. This would translate to negative effect size sign if treatment group is less likely to abandon the nest. To fix this we need to reverse the sign of the effect sizes for traits where increased trait values represent LESS reproductive investment, so we have consistent biological interpretation over the whole data set.

<br>
********************************************************************************

### Documentation and reporting

**ANSWER 18:**   
Items 10, 11 and 18

**ANSWER 19:**   
Generally speaking, it is for transparency, reproducibility and giving back to the community.

<br>

*We hope it all makes sense - if not sure about some answers, please talk to us*

### 2019-01-28 Data_extractions.Rmd
minor formatting edits and typos fixed

@@
 
-Lets try to extract data from:
- 1. Figure 2a in the paper by Weil et al. 2006: Immune challenge retards seasonal reproductive regression in rodents: evidence for terminal investment (Weil2006.pdf)
+Let's try to extract data from:   
+ 1. Figure 2a in the paper by Weil et al. 2006: Immune challenge retards seasonal reproductive regression in rodents: evidence for terminal investment (**Weil2006.pdf**)   
  2.
 Figure 2a in the paper by Sadd et al.
-2006: Modulation of sexual signalling by immune challenged male mealworm beetles (Tenebrio molitor, L.): evidence for terminal investment and dishonesty (Sadd2006.pdf)
+2006: Modulation of sexual signalling by immune challenged male mealworm beetles (Tenebrio molitor, L.): evidence for terminal investment and dishonesty (**Sadd2006.pdf**)
 
@@
 
-Name the files as following (this makes it easy to keep track of the figures being digitised):     
-001_Weil_2006_Fig2a.png    
+Name the snapshot files as following (this makes it easy to keep track of the figures being digitised):     
+**001_Weil_2006_Fig2a.png**    
 and    
-002_Sadd_2006_Fig2a.png
+**002_Sadd_2006_Fig2a.png**
 
-We need to place them in a new folder - we called it metaDigitise_figures (read more about organising your extraction directories in the "Setting up directory structures " section of https://cran.r-project.org/web/packages/metaDigitise/vignettes/metaDigitise.html).
+You have to place them in a new folder - let's call it **metaDigitise_figures** (read more about organising your extraction directories in the "Setting up directory structures " section of https://cran.r-project.org/web/packages/metaDigitise/vignettes/metaDigitise.html).
 
-To extract our two figures we need to provide the directory path name of the folder with the images to the metaDigise() function, for example:
+To extract our two figures we need to provide the directory path name of the folder with the images to the *metaDigise()* function, for example:
 
@@
 **QUESTION 17:**    
-If we have different traits in the data set (lets say number of eggs laid and probability of next abandonment) what is the problem with the  signs of effect sizes for these traits?
+If we have different traits in the data set (lets say number of eggs laid and probability of nest abandonment) what is the problem with the signs of effect sizes for these traits?
 What should we do about this?

### 2019-01-28 Trees_making.Rmd
added schematic trees examples and tree grafting

@@
 
+**Setup:** Work can be done individually, in pairs or small groups. R/Rstudio and Internet connection needed.
+
 Introduction
@@
 1. We can use a tree with or without branch lengths (divergence measures).
-2. The tree has to be binary (strictly bifurcating = no polytomies, i.e. only two branches are coming from each node).  
+
+```{r tree exmaple1, eval=TRUE, echo=FALSE, fig.width=6, fig.height=4, fig.cap=" "}
+library(ape)
+par(mfcol=c(1,2))
+plot(read.tree(text = "(((Species4, Species3), Species2), Species1);"), main = "no branch lengths")
+plot(read.tree(text = "(((Species4:0.1, Species3:0.1):0.3, Species2:0.4):1.9, Species1:2.3);"), main = "with branch lengths", use.edge.length = TRUE)
+```
+
+2. The tree has to be binary (strictly bifurcating = no polytomies, i.e. only two branches are coming from each node). Here is a simple example:
+
+```{r tree exmaple2, eval=TRUE, echo=FALSE, fig.width=6, fig.height=4, fig.cap=" "}
+par(mfcol=c(1,2))
+plot(read.tree(text = "(((Species4, Species3), Species2), Species1);"), main = "binary")
+plot(read.tree(text = "((Species4, Species3, Species2), Species1);"), main = "non-binary")
+```
+
 3. The tree needs to be rooted (has a node representing the most recent common ancestor of all the species in the tree).
+
+```{r tree exmaple3, eval=TRUE, echo=FALSE, fig.width=6, fig.height=4, fig.cap=" "}
+par(mfcol=c(1,2))
+plot(read.tree(text = "(((Species4, Species3), Species2), Species1);"), main = "rooted")
+plot(read.tree(text = "((Species4, Species3), Species2, Species1);"), main = "unrooted")
+```
+
 4. The tree tip labels (representing species) must match **exactly** with the species names in the data set (otherwise MA models will not run with them).
+
+```{r tree exmaple4, eval=TRUE, echo=FALSE, fig.width=6, fig.height=4, fig.cap=" "}
+par(mfcol=c(1,1))
+plot(read.tree(text = "(((Species4 != SPECIES4, Species3 == Species3), Species2 == Species2), Species1 == Species1);"), main = "rooted")
+```
+
 5. Trees can be also made and used for higher taxonomic levels than species, e.g. genus-level (not covered in this tutorial, but general principles are the same).
+
+```{r tree exmaple5, eval=TRUE, echo=FALSE, fig.width=6, fig.height=3.2, fig.cap=" "}
+par(mfcol=c(1,1))
+plot(read.tree(text = "((GenusC, GenusB), GenusA);"), main = "genus-level tree")
+```
 
@@
 At the webpage, use an input box at the top of the page to search each species name or upload list of the species names from a file.
-This allows to generate a scaffold for a tree.
+This allows to generate a scaffold for a tree (i.e.
+the pattern of basal branching for the main clades).
  - If few major clades are present, assemble together sub-trees created from bigger trees, if these exist for these clades.    
@@
 Again, your final tree, must contain exactly the same list of species names as your meta-analytic data set.
+
+Run the below code chanks IF you have not installed these packages earlier (it may take 15-20 minutes to install all).
 
@@
 
-Extract and clean the list of species included. Note that genus and species name are stored in separate columns.
+Extract and clean the list of species included.
+Note that genus and species name are stored in separate columns.
+Also, smae species name can appear several times in the data set, but since we need a species list, we need each name to apper only once (i.e.
+unique names), and capitalisation to be consistent (letter case matters when matching strings).
 
@@
 
+**Note:** populus deltoides
+
 ```{r get Curtis1998 tree}
@@
 So, we run into a problem with one of the Populus names - need to fix this by using an alternative name.
+Populus euramericana, is a naturally occurring hybrid of Populus deltoides, so we can replace it with this name for the search (given its colesely related and we dont have this species on our species list).
 
 ```{r fix Curtis1998 species_Curtis1998 list}
-populus <- tnrs_match_names(names = ("populus deltoides"))
+populus <- tnrs_match_names(names = ("populus canadensis"))
 populus #looks ok
-species_Curtis1998 <- gsub("Populus euramericana", "Populus deltoides", species_Curtis1998) #replace populus euramericana with populus deltoides
+species_Curtis1998 <- gsub("Populus euramericana", "Populus deltoides", species_Curtis1998) #replace populus euramericana with populus canadensis
 ```
@@
 Acers_tree <- search_treebase("418", by="id.tree", exact_match = TRUE) #returns a single tree in a list
-Acers_tree <- Acers_tree[[1]] #extract tree from the list
+Acers_tree <- Acers_tree[[1]] #extract the tree from the list
 ```
@@
 
-```{r plot pruned Acers_tree, fig.width=10, fig.height=4, echo=TRUE, message=FALSE}
+```{r plot pruned Acers_tree, fig.width=10, fig.height=3.2, echo=TRUE, message=FALSE}
 plot(Acers4sp_tree, cex=.8, label.offset =.1, no.margin = TRUE)
@@
 setdiff(Acer4sp, as.character(Acers4sp_tree$tip.label)) #no mismatches
-is.binary.tree(tree_random) #TRUE
+is.binary.tree(Acers4sp_tree) #TRUE
+```
+
+We can graft this tree on the Curtis1998 tree to non-randomly resolve polytomy for genus Acer.   
+**Note:** when grafting you lose branch lengths on the original tree, if present.
+
+We first need to prune given branch to remove all but one species affected by polytomy. Then we can replace this species with the resolved sub-tree (Acers4sp_tree):
+
+```{r graft Acers_tree}
+tree_pruned <- drop.tip(tree, c("Acer saccharum", "Acer rubrum", "Acer saccharinum")) #prune 3 species
+tree_grafted <- bind.tree(tree_pruned, Acers4sp_tree, where=which(tree_pruned$tip.label=="Acer pensylvanicum")) #graft 3 species to the remaining Acer species
+```
+
+```{r plot tree_grafted Acers, fig.width=10, fig.height=8, echo=TRUE, message=FALSE}
+plot(tree_grafted, cex=.8, label.offset =.1, no.margin = TRUE)
 ```
@@
 
-Use *treebase* to create a tree for the three Quercus species from dat.Curtis1998.
+Use *treebase* to create a tree for the three Quercus species from dat.Curtis1998. You can try to graft them on the Curtis 1998 tree.
 

### 2019-01-28 Trees_making.Rmd
fixed tree saving and reading

@@
 # is.ultrametric(tree) #the tree has no branch lengths, so cant be checked
+write.tree(tree, file="tree_curtis1998.tre") #save the tree
+# tree <- read.tree(file="tree_curtis1998.tre") #if you need to read it
 ```
@@
 write.tree(tree_random, file="tree_curtis1998_random.tre") #save the tree
-# read.tree(file="tree_curtis1998_random.tre") #if you need to read it
+# tree_random <- read.tree(file="tree_curtis1998_random.tre") #if you need to read it
 ```
@@
 
-We first need to prune given branch to remove all but one species affected by polytomy. Then we can replace this species with the resolved sub-tree (Acers4sp_tree):
+We first need to prune given branch from the main tree to remove all but one species affected by polytomy. Then we can replace this species with the resolved sub-tree (Acers4sp_tree):
 

### 2019-01-28 Trees_making_solutions.Rmd
added tree grafting solution

@@
 Quercus3sp_tree <- drop.tip(Quercuss_tree, Quercuss_tree$tip.label[-match(Quercus3sp, Quercuss_tree$tip.label)])
-plot(Quercus3sp_tree, cex=.8, label.offset =.1, no.margin = TRUE)
+#plot(Quercus3sp_tree, cex=.8, label.offset =.1, no.margin = TRUE)
 
@@
 Quercus3sp_tree$tip.label <- gsub("_"," ", Quercus3sp_tree$tip.label) #get rid of the underscores
+```
+
+```{r plot Quercus3sp_tree, fig.width=10, fig.height=3, echo=TRUE, message=FALSE}
 plot(Quercus3sp_tree, cex=.8, label.offset =.1, no.margin = TRUE)
@@
 
+We first need to prune given branch from the main tree to remove all but one species affected by polytomy. Then we can replace this species with the resolved sub-tree (Quercus3sp_tree):
+
+```{r graft Quercus_tree}
+tree <- read.tree(file="tree_curtis1998.tre") #run if you need to read it in
+tree_pruned <- drop.tip(tree, c("Quercus prinus", "Quercus alba")) #prune 3 species
+tree_grafted <- bind.tree(tree_pruned, Quercus3sp_tree, where=which(tree_pruned$tip.label=="Quercus_rubra")) #graft 3 species to the remaining Acer species
+```
+
+```{r plot tree_grafted Acers, fig.width=10, fig.height=8, echo=TRUE, message=FALSE}
+plot(tree_grafted, cex=.8, label.offset =.1, no.margin = TRUE)
+```
+
 Done!

### 2019-01-28 Trees_making.Rmd
tidying up and minor fixes

@@
 
-# write.tree(tree, file="tree_curtis1998.tre") #save the tree in Newick format, if needed
+write.tree(tree, file="tree_curtis1998.tre") #save the tree in Newick format, if needed
 # read.tree(file="tree_curtis1998.tre") #if you need to read it

### 2019-02-02 Data_extractions.Rmd
Incorporated comments from Alfredo and Nick

@@
 Imagine, you quickly got well and instead of reproducing decided to carry on the Terminal Investment Meta-analysis Project.   
-After a few weeks you fished searching and screening literature for the relevant papers.
+After a few weeks you finished searching and screening literature for the relevant papers.
 You now have a collection of around 30 included full-text papers (it could be 300 as well), stored neatly as pdf files and supplementary information, in whatever format available.
-Its time to start extracting the data (ideally, you should have planned this in detail while writing up your protocol/preregistration, but in this exercise we assume you did not).
+It is time to start extracting the data (ideally, you should have planned this in detail while writing up your protocol/preregistration, but in this exercise we assume you did not).
 
@@
 **QUESTION 4:**   
-Whats wrong with this table representing a fragment of an extraction spreadsheet?
+What is wrong with this table representing a fragment of an extraction spreadsheet?
 
@@
 **QUESTION 9:**   
-What is meta-data in meta-analysis and what it is used for?
+Why we should record meta-data when extracting fdata for meta-analysis?
 
-**Note:** In your actuall extraction sheet/form, try to group related columns together.
+**Note:** In your actual extraction sheet/form, try to group related columns together.
 You can also color-code them in Excel or put on the same tab/section in a visual database interface, if used.
@@
 We will practice data extraction on few of the Terminal investment papers. We will use R package *metaDigitise*.   
-Detailed package description and tutorial are available at: https://cran.r-project.org/web/packages/metaDigitise/vignettes/metaDigitise.html
+Detailed package description and tutorial are available at: https://cran.r-project.org/web/packages/metaDigitise/vignettes/metaDigitise.html or
+
 https://github.com/daniel1noble/metaDigitise and this publication:  
@@
 
-Here is the package installation code:
+Here is the package installation code (run if you have not installed it earlier):
 
@@
 
+Load the packages we will need:   
 ```{r load packages, eval=TRUE, message = FALSE}
@@
 library(tidyverse)
+library(compute.es)
 ```
@@
 
-You will need to open the pdf files and take the snapshots of the relevant figures.
+You will need to open the pdf files and take the snapshots of the relevant figures (for MacOS the easiest is via Command+Shift+4 and then select the screen fragment; for Windows see https://www.windowscentral.com/how-take-windows-10-screenshot-7-best-techniques).
 Make sure you capture the axes and axes labels.
@@
 We will now use *escalc* function from *metafor* package to calculate effect sizes.
-The function allows to calculate various effect sizes or outcome measures (and the corresponding sampling variances) that are commonly used in meta-analyses, including two effect sizes are relevant to our data and question:
+The function allows to calculate various effect sizes or outcome measures (and the corresponding sampling variances) that are commonly used in meta-analyses, including two effect sizes that are relevant to our data and question:
  * "SMD" stands for standarised mean difference, i.e. Hedges' g (potentially also "SMDH" for heteroscedastic variances).   
@@
 
-Open the above webpage. Main menu has an expandable list of types of effect sizes, and statistical tests you can use to calculate these effect sizes (only d, r, OR and RR).
+Open the above webpage. Main menu has an expandable list of types of effect sizes, and statistical tests you can use to calculate these effect sizes (only *d*, *r*, *OR* and *RR*).
 
@@
 Weil 2006 reports in the results: "At the end of the experiment saline-injected animals had smaller testes (F 1,27 = 4.6875, p<0.05, figure 2b)" and in the methods it is noted these are calculated using ANOVA.
-Use online ES calculator sheet selecting "Standarized Mean difference (d)"" and "F-test, 2 groups" for to calculate d from one-way ANOVA (you will need F-value and sample sizes of treatment and control groups).
+Use online ES calculator sheet selecting "Standarized Mean difference (d)" and "F-test, 2 groups" to calculate *d* from one-way ANOVA (you will need F-value and sample sizes of treatment and control groups).
 
-You should get d for the testes  mass that is close to what we calculated for tests volume from Fig2a.
+You should get *d* for the testes  mass that is close to what we calculated for tests volume from Fig2a.
 Note that this web calculator provides confidence intervals (CI) and variance for the calculated effect sizes.
@@
 **TASK 5**:   
-Using web-based ES calculator, estimate a few other effect sizes using test (inferential) statistics reported in Sadd2006.
+Using web-based effect size calculator, estimate a few other effect sizes using test (inferential) statistics reported in Sadd2006.
+
+In practice, we would preer to do it in R, rather than in aan online up or any other non-scriptable software that cannot socument our calculations.
+Luckily, the R package *compute.es* allows not only compute various effect sizes from descriptive statistics, but also from inferential statistics.
+r function inputs and names are as follows:
+ * ANCOVA F-test: a.fes()
+ * ANCOVA means: a.mes()
+ * ANCOVA means (pooled sd): a.mes2()
+ * ANCOVA p-value: a.pes()
+ * ANCOVA t-test: a.tes()
+ * Chi-squared (1 df): chies()
+ * Correlation: res()
+ * d-statistic: des()
+ * Failure group (binary): failes()
+ * F-test: fes()
+ * Log odds ratio: lores()
+ * Means: mes()
+ * Means (pooled sd): mes2()
+ * Proportions (binary): propes()
+ * p-value: pes()
+ * t-test: tes()
+To learn more see the package vignette or https://www.rdocumentation.org/packages/compute.es/versions/0.2-4/topics/compute.es-package
+
+As an example, we will use F value from **TASK 4**, note that functions from the *compute.es* packages automatically calcualte all available effect sizes for the given imput and if you run it on a data frame it will generate a new data.frame with several effect size estimates.
+Try:
+
+```{r use compute.es, eval=TRUE, echo = TRUE, message = FALSE}
+#single calculation:
+a.fes(f = 4.6875, n.1 = 28, n.2 = 10, R = 0, q = 0) #where we enter F-value, sample sizes of treatment and control group, covariate correlation and number of covariates (we have none in ANOVA))
+
+#for vectors of values (lieke in a data frame):
+results_ES <- a.fes(f = c(4.6875, 4.6875), n.1 = c(28, 10), n.2 = c(10, 10), R = c(0, 0), q = c(0, 0))
+str(results_ES)
+```
 
@@
 **QUESTION 16:**    
-Is the sign of effect size important? Imagine we use d, which represents mean_treatment - mean_control group trait values. How we would interpret d = 0.5 and how we would interpret d = -0.5?
+Is the sign of effect size important? Imagine we use *d*, which represents mean_treatment - mean_control group trait values. How we would interpret *d* = 0.5 and how we would interpret *d* = -0.5?
 
@@
 **QUESTION 19:**   
-Although not mentioned in the original PRISMA checklist, it is important to share the collected data, meta-data and analysis code? Think of east at least 3 reasons for doing so.
-
-**ANSWER 19:**   
-Generally speaking, it is for transparency, reproducibility and giving back to the community.
+Although not mentioned in the original PRISMA checklist, it is important to share the collected data, meta-data and analysis code? Think of at least 3 reasons for doing so.
 

### 2019-02-02 Data_extractions_solutions.Rmd
Incorporated comments from Alfredo and Nick

@@
  3. Species column is a mix of Latin and common names (should be all Latin)   
+ 4. There is an extra dot in " 6.4      .""
 
@@
 These are not independent data points, and this is captured by CohortID.
+For different study question this could be an unique identifier of various study sites or populations that need tp be distinguished within and between studies.
 
@@
 Meta-data is a detailed description of all extracted variables/columns and all other variables/columns derived from them.
-It is a part of the documentation of your meta-analysis and it is necessary when you or somebody else checks the data or tries to re-analyse or re-use it.
+More specifically it describes how the data was coded (e.g.
+the measurement units, how conversions were made, any decisions and assumptions that had to be made in general for a given column of data"; e.g.
+whether sex was coded as "F" and "M" and "B", and what these values mean, jhow missing values were coded, and anything else that would allow you or somebody to extract and code the same data).
+Meta-data is an essential part of the documentation of your meta-analysis and it is extremely useful when you or somebody else checks the data or tries to re-analyse or re-use it.
 It is a good practice to start creating such documentation when you start extractions and coding and update it as you make changes and progress to analyses.
@@
 **ANSWER 10:**    
-Ideally, two people would be extracting the data in parallel 9independently, in-duplicate), to minimise errors.
+Ideally, two people would be extracting the data in parallel (independently, in-duplicate), to minimise errors.
 However it is often not practical.
@@
 **ANSWER 12:**    
-There are several pieces of software that helps digitising various types of plots and extracting the data (you can find a list here: https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.13118), so it is often possible.
+There are several pieces of software that helps digitising various types of plots and extracting the data (you can find a list here: https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.13118, so it is often possible.
 However, sometimes plot resolution can be too low, or data points / lines are overlapping and obscuring some of the values.

### 2019-02-02 Papers_search.Rmd
added setup comment
incorporated comments from Rose and Yong

@@
 
+**Setup:** Work in pairs or small groups. R/Rstudio and Internet connection needed.
+
 ### Purpose of the exercise:   
@@
 Imagine, you woke up in the morning feeling rather sick.
-You decided to work from home to get some rest a get better more quickly.
-You don't not feel energetic enough to crunch numbers or write, so you thought you may catch up on scientific reading instead.
+You decided to work from home to get some rest.
+You are not feeling energetic enough to crunch numbers or write, so you decide to catch up on scientific reading instead.
 You pick up the book sitting on top of the "to-read"" pile and open it at a random place.
@@
 ```
+**Note**: LPS is a lipopolysaccharide that is constituent of the outer membranes of some bacteria and it is commonly used to trigger immune reaction without causing greater harm in animals.   
+<br>
 
-You find it intriguing and start thinking this might be something (a topic) you could do a meta-analysis on.
+You find this fragment intriguing and start thinking this might be something (a topic) you could do a meta-analysis on.
 *(Ok, this is a hypothetical example of coming up with the topic for a meta-analysis, but it sometimes happens - you read or hear some interesting questions and it sparks desire to put some numbers on some pattern...)*
@@
 
-In this exercise, we will use a few basic tools to illustrate how to perform the first few steps of meta-analysis: refining the question, searching and screening literature (these are the same steps as in Systematic Reviews, so the tools are the same and can be usually described as systematic literature/evidence reviewing tools).
-Note that some R packages (and many other online/software tools) are available (and more always coming) that can help with some of the tasks during this exercise, but there is no single "perfect" tool for the whole process usable for all disciplines and types of questions (although some attempts have been made).
+In this exercise, we will use a few basic tools to illustrate how to perform the first few steps of a meta-analysis: refining the question, searching the literature, and screening the found studies (these are the same steps as in Systematic Reviews, so the tools are the same and can be usually described as systematic literature/evidence reviewing tools).
+Note that some R packages (and many other online/software tools) are available (and more are bieng developed) that can help with some of the tasks during this exercise, but there is no single "perfect" tool for the whole process usable for all disciplines and types of questions (although some attempts have been made).
 We simply do not have enough time to introduce more tools and the main purpose of this exercise is to familiarize you with the basic principles/issues that apply when you use any tools.
@@
 
-Topic is usually broad and vague area of interest, for a meta-analysis we need a much more concrete question.
+The topic is usually broad and vague area of interest, for a meta-analysis we need a much more concrete question.
 Generally, the more focused the question, the easier and quicker it is to perform the meta-analysis, but the answer becomes less general and there might be less evidence available.
@@
 
-Other things you should know about Google Scholar - its search interface is rather simplistic and restrictive, it shows only up to 1,000 results for any particular search query, ranking algorithms are obscured, its hard to export the references, the literature coverage may fluctuate (its based on web scraping) and the searches may not be reproducible.
+Other things you should know about Google Scholar - its search interface is rather simplistic and restrictive, it shows only up to 1,000 results for any particular search query, ranking algorithms are obscured, its hard to export the references, the literature coverage may fluctuate (it is based on web scraping) and the searches may not be reproducible.
 Because of this it is not really recommended for the main searches in systematic reviews.
@@
 **RESOURCES:**   
-Other software useful for exploratory (scoping) ad additional searches (check them up later, also note that they may have similar disadvantages to these of Google Scholar):
+Other software useful for exploratory (scoping) and additional searches (check them out later, and note that they may have similar disadvantages to these of Google Scholar):
  * Semantic Scholar: https://www.semanticscholar.org/
@@
 *"terminal investment" reproduction*   
-This time you should get around 1900 records.
+This time you should get around 1,900 records (you will be allowed to access only the first 1,000).
 Note that Google Scholar does so called Automated Query Expansion for terms that are not set to be exact phrases (in quotes) - it uses alternative forms of these terms.
@@
 
-Formulate meta-analytic question "whether immune challenge can result in terminal investment" using PICO terms (actually this fits really well because we can use experimental data).
+Formulate our meta-analytic question "whether immune challenge can result in terminal investment" using PICO terms (actually this fits really well because we can use experimental data).
 
 **RESOURCES:**   
-An extensive list of alternative systematic review formulation question frameworks here:  https://www.networks.nhs.uk/nhs-networks/nwas-library-and-information-service/documents/alternative-question-structures-for-different-types-of-systematic-review    )
+An extensive list of alternative systematic review formulation question frameworks can be found here:  https://www.networks.nhs.uk/nhs-networks/nwas-library-and-information-service/documents/alternative-question-structures-for-different-types-of-systematic-review    )
 
@@
 
-**Note:** publication status can be also used as a criterium - are you happy to include some types of grey literature (e.g. conference proceedings, theses)?
+**Note:** publication status can be also used as a criterion - are you happy to include some types of grey literature (e.g. conference proceedings, theses)?
 
 **QUESTION 6:**  
-Study type: should we include any type of study?
+Study type: should we include any type of study? (e.g. correlational vs experimental, if experimental what type of experiment?)
 
@@
 Data type: should we accept any type of data for meta-analysis?
+(e.g.
+descriptive statistics, and which ones, or also inferential model statsistics, and which ones?
+This will be linked to the effect size you are planning to use!)
+
 <br>
@@
 ********************************************************************************
-### Making decision trees
+### Making decision trees for the literature screening
 
-Good decision trees make screening the literature faster and more reliable.  
-Briefly, if the initial (most general) questions, representing screening criteria, are not fulfilled, the study gets excluded and there is no need to evaluate the remaining criteria, so we save some time.
+**Note**: this part could be introduced later with the screening, but its closely linked to the inclusion criteria, so it comes here.
+
+Decision trees are visual representations of our inclusion criteria and decisions we have to make when screening literature.
+Importantly, good decision trees can make screening the literature faster and more reliable.
+Briefly, if the initial (most general) questions representing screening criteria are not fulfilled, the study gets excluded and there is no need to evaluate the remaining criteria, so we save some time.
 
@@
 What question is missing on the top of the full-text decision tree?   
-(Tip: its not about the Language or publication year)   
+(Tip: what happens between abstract and full text screening?)  
 <br>
@@
 The reason is that we want to have as many relevant data points as possible to increase the power of statistical analyses.
-The other reason is that we make effort to capture obscure studies that potentially are more likely to report non-significant results, and this way we actively try to reduce publication bias in our data set.
+The other reason is that we make effort to capture obscure studies that potentially are more likely to report non-significant results, and ttherefore help reduce publication bias in our data set.
 Search strategy should include using at least two wide-coverage/interdisciplinary databases (we recommend Scopus and Web of Science), and a few additional sources of literature such as forward and backward reference searching (looking at cited and citing papers), publication lists of the key authors or labs, searching for unpublished studies and data sets (we will not cover these here).
@@
 Web of Science (WoS): http://apps.webofknowledge.com   
-**Note:** WoS coverage depends of institutional subscription package purchased, so results of searches can vary among institutions.
+**Note:** WoS coverage depends on the institutional subscription package purchased, so results of searches can vary among institutions.
 Also searching just core collection vs.
@@
 **TASK 5:**   
-Using Scopus **Advanced Search** (you can read more about the syntax here: https://dev.elsevier.com/tips/ScopusSearchTips.htm) run this search strings:  
+Using Scopus **Advanced Search** (you can read more about the syntax here: https://dev.elsevier.com/tips/ScopusSearchTips.htm) run these search strings:  
  1. *TITLE-ABS-KEY ( "terminal investment"  AND  "immune challenge" )*    
@@
 
-Ideally, we should think and test some more terms for inclusion, after doing some scoping searches, reading reviews and sentinel relevant empirical papers.
+Ideally, we should think and test some more terms for inclusion, after doing some scoping searches, reading reviews and seminal relevant empirical papers.
+Also, it is important to validate the search by checking how many papers it finds that we know contain useable data.
+You should also do some pilot abstract screening on papers exported from just one database, to check whether there’s some exclusion words you did not think about, and which could be included in your search string as exclusion terms.
+
 We stop for now and assume that the string above is our final search string.
-Note that Scopus Advanced Search stores all your searches ans search strings in a neat table at the bottom of the search window - its easy to copy this out and include in the documentation on how you developed your search strategy (same applies to WoS database; you need to register in Scopus and WoS to save searches and alerts across multiple search sessions).
+Note that Scopus Advanced Search stores all your searches and search strings in a neat table at the bottom of the search window - its easy to copy this out and include in the documentation on how you developed your search strategy (same applies to WoS database; you need to register in Scopus and WoS to save searches and alerts across multiple search sessions).
 The \* symbol stands for wildcard and indicates "any symbols" since these databases do not do Automatic Query Expansion.
@@
 249 hits - more than from Scopus).
-WoS stores your search history and also which sub-data bases were included in the search: Indexes=SCI-EXPANDED, SSCI, A&HCI, CPCI-S, CPCI-SSH, BKCI-S, BKCI-SSH, ESCI, CCR-EXPANDED, IC Timespan=All years" (you can restrict these in Advanced Search)
+WoS stores your search history and also which sub-data bases were included in the search: Indexes=SCI-EXPANDED, SSCI, A&HCI, CPCI-S, CPCI-SSH, BKCI-S, BKCI-SSH, ESCI, CCR-EXPANDED, IC Timespan=All years" (you can restrict these in the Advanced Search).
+
+```{r WoS_adv_search, echo=FALSE, fig.cap=" ", out.width='100%'}
+knitr::include_graphics("./images/WoS_adv_search.png")
+```
 
@@
 number per export, **remember to tick the boxes for exporting abstracts!**) - there are multiple file formats you can use, but .ris and .bib are most commonly used ones- you will need to figure out which ones work for your reference manager software.
-The next step would be downloading all found records from your "final" search into reference manager of your choice.
-Combine the references from two databases into a single project/folder and removing duplicates (most reference managers have such function, but it never works perfectly - do a visual check after sorting by titles!) before proceeding to screening stage.
+The next step would be downloading all found records from your "final" search into the reference manager of your choice.
+Combine the references from two databases into a single project/folder and remove duplicates (most reference managers have this function, but it never works perfectly - do a visual check after sorting by titles!) before proceeding to the reference screening stage.
 An R package *revtools* (https://cran.rstudio.com/web/packages/revtools/ and http://revtools.net/) can act as a reference manager and also has a good de-dupiction function.
 
-**Note:** It is also a good time to write up your meta-analysis protocol and preregistering your meta-analysis (e.g. on OSF)!
+**Note:** This is also a good time to write up your meta-analysis protocol and register your meta-analysis (e.g. on OSF)!
 <br>
@@
 
-**Note:* it is possible to deduplication algorithms in these are not very good, so its advisable to deduplicate and check your records in the reference manager before uploading to the screening software.
+**Note:** it is possible that the deduplication algorithms in these are not very good, so it is advisable to deduplicate and check your records in the reference manager before uploading to the screening software.
 
 **TASK 7:**  
-We prepared a file (*Scopus10.ris*) with 10 references sampled from the set found using search string (these are sets of references from *( ( "terminal investment"  OR  "reproductive effort" )  AND  ( "immune challeng\*"  OR  "immunochalleng\*" ) )  *) in Scopus (same references came up in the equivalent search in WoS).
+We prepared a file (*Scopus10.ris*) with 10 references sampled from the set found using the search string in Scopus database (used *( ( "terminal investment"  OR  "reproductive effort" )  AND  ( "immune challeng\*"  OR  "immunochalleng\*" ) )  *) (same references came up in the equivalent search in WoS).
 
-**Note:** Sometimes the biblio-files with the same extension differ a bit in structure and reference managers (and Rayyan) may have problem opening them.
-Getting your files into exactly right format can sometimes be frustrating.
+**Note:** Sometimes the biblio-files with the same extension differ a bit in structure and reference managers, and Rayyan may have problem opening them.
+Getting your files into exactly the right format can sometimes be frustrating.
 
-We will use these ten references to practice title and abstract screening in Rayyan. Screening can be also done in Abstracker, but we found Rayyan bit more reliable and more user-friendly.   
+We will use these ten references to practice title and abstract screening in Rayyan.
+Screening can be also done in Abstracker, but we find Rayyan a bit more reliable and more user-friendly (could be slow with updating the display).
+
 If you are working in pairs or small groups, one person should create a new project in Rayyan (see below) and invite another person/people to the project using their email address.
@@
 Click on the "Uploaded References [scopus10.ris]" link in this panel to see the references in the right panel.
-Generally clicking on the links within panels on the left will filter the records in the right panels to include on highlighted sets (filtering, to get rid of the filter click on it again to remove highlighting).
+Generally clicking on the links within panels on the left will filter the records in the right panels to include only highlighted sets (to get rid of the filter click on it again to remove highlighting).
 In the left panels there are several sub-panels.  
@@
  6.
-"Keywords for exclude" - " set to: review, reviewed, model, human, people, women, men (these terms will be highlighted red in the titles and abstracts to help them find more easily; Rayyan tries to populate this list by itself - but its algorithm is medical-centered so you should adjust this list for your projects).
+"Keywords for exclude" - " set to: review, reviewed, model, human, people, women, men (these terms will be highlighted red in the titles and abstracts to help them find more easily; again, Rayyan tries to populate this list by itself).
  7.
-Click on "All reviews" button in the top right corner to go to the project lists, then open your=project again by choosing it from the list.
+Click on "All reviews" button in the top right corner to go to the project lists, then open your project again by choosing it from the list.
 A view with a pie chart (your screening progress) and a few buttons will appear.
@@
  9.
-Using pour abstract screening decision tree decide whether include this paper for the next stage of screening and press "Include" or "Exclude" button accordingly.
+Using our abstract screening decision tree decide whether include this paper for the next stage of screening and press the "Include" or "Exclude" button accordingly.
 And do the assessment for the remaining records.
@@
 You could also enter reasons for exclusion, e.g.
-"human" (we normally don't do this as there are often hundreds or thousands excluded records per project at this stage).
+"human" (we normally don't do this as there are often hundreds or thousands of excluded records per project at this stage).
 10.
 Go to "All reviews" and check if other people finished their screening, if applicable.
-After that you can lift the "blind" mode, export all the records with decisions into a .csv files.
+After that you can lift the "blind" mode (when colleborators cannot see each others decisions).
 For collaborative screens Rayyan will now provide filters to display all consensus included, consensus excluded and records with conflicting decisions.
 If you have any "conflicts" have a look again and discuss why you couldn't agree on a given record , come to agreement and make final decision.
-Export the records with final consensus decisions, for documentation.
+Export the records with final consensus decisions, for documentation, into a .csv file.
+You can filter the records into “Included” and just export those, also you can export them as a .bib or .ris file into a reference manager where you will collect full texts for the enxt stage of the screening.
 
 **QUESTION 16:** 
-How many papers you excluded? How many conflicts you got if you collaboratively screened (and if so what was causing these conflicts, can the decision tree be improved?)
+How many papers did you exclude? How many conflicts did you have, if you collaboratively screened and, if so, what was causing these conflicts? Can the decision tree be improved and how?
 
 **TASK 8:**  
-Its time to do the full-text screening for the papers that passed title and abstract screening.
-There are some full-text papers that passed the initial screening in the directory - assess Nielsen2012 and Ardia2005 using our full-text screening decision tree.
-Hint: dont read the whole paper - just try to skim it and look for the key information.
+It is time to do the full-text screening for the papers that passed title and abstract screening.
+There are some full-text papers that passed the initial screening in the directory - assess **Nielsen2012** and **Ardia2005** using our full-text screening decision tree.
+Hint: do not read the whole paper - just try to skim it and look for the key information.
 
 **QUESTION 17:**  
-What was your decision about Nielsen2012 and Ardia2005 papers?   
+What was your decision about **Nielsen2012** and **Ardia2005** papers?   
 <br>
@@
 
-It is very important to keep detailed record of the whole searching, screening and data extraction process. Searching and screening are usually represented as some variant of PRISMA diagram.
-With our 144 records from Scopus, 249 from WoS (these two databases are our "main search"), and imaginary around 400 records from other searches we might have done if we had more time, we would probably end up with something like this one:
+It is very important to keep detailed record of the whole searching, screening and data extraction process. Searching and screening are usually represented as some variant of a PRISMA diagram.
+With our 144 records from Scopus, 249 from WoS (these two databases are our "main search"), and imaginary around 400 records from other searches of other sources of literature we might have done if we had more time, we would probably end up with something like this one:
 
@@
 text(elpos[26, 1] - 0.07, elpos[26, 2] + 0.02, "15")
-text(elpos[19, 1] + 0.12, elpos[19, 2] + 0.12, "approx.380")
+text(elpos[19, 1] + 0.12, elpos[19, 2] + 0.12, "approx.385")
 text(elpos[26, 1] + 0.14, elpos[26, 2] + 0.02, "6")
@@
 
-**Note:** When you exclude records in the full-text screen keep record for each individual paper on the main reason for exclusion.
+**Note:** When you exclude records in the full-text screen keep a record for each individual paper on the main reason for exclusion.
 Later present this information as a supplementary table for your meta-analyses.
@@
 Above you can see a PRISMA (flow) diagram. The other important document(s) associated with PRISMA is a checklist (http://prisma-statement.org/documents/PRISMA%202009%20checklist.pdf). 
-PRISMA checklist is a document that helps making sure that you reported the key information how a systematic review (including meta-analysis) was done.    
-Have a look at the PRISMA checklist. Find and read the items that are related to what we have covered in this exercise. What are the numbers (in the # column) of these items?
+The PRISMA checklist is a document that helps ensure you reported the key information on how a systematic review (including meta-analysis) was done.    
+Have a look at the PRISMA checklist. Find and read the items that are related to what we have covered in this exercise. What are their item numbers (in the # column)?
 
-**Note:** this PRISMA was designed for meedical systematic reviews, and some items are not very relevant or useful. Also, it is due to be updated.   
+**Note:** this PRISMA was designed for medical systematic reviews, and some items are not very relevant or useful. Also, it is due to be updated.   
 <br>
 
-*We hope you enjoyed this exercise! Remeber, we just roughly covered the basics and there is much more to learn!*
+*We hope you enjoyed this exercise! Remember, we just roughly covered the basics and there is much more to learn!*

### 2019-02-06 Data_extractions.Rmd
incorporated comments from Rose and Yong

@@
 
-Imagine, you quickly got well and instead of reproducing decided to carry on the Terminal Investment Meta-analysis Project.   
+Imagine, you quickly got well and decided to carry on the Terminal Investment Meta-analysis Project.   
 After a few weeks you finished searching and screening literature for the relevant papers.
-You now have a collection of around 30 included full-text papers (it could be 300 as well), stored neatly as pdf files and supplementary information, in whatever format available.
+You now have a collection of around 30 included full-text papers (relatively manageable; some projects have 300 or more), stored neatly as pdf files and supplementary information, in whatever format available.
 It is time to start extracting the data (ideally, you should have planned this in detail while writing up your protocol/preregistration, but in this exercise we assume you did not).
@@
 Google Forms).
-Some researchers even build elaborated relational databases (there is a high initial investment cost but many benefits later on).
+Some researchers even build elaborated relational databases (which recoup the high initial investment cost with many benefits later on).
 Whatever you chose, you need a platform where you can freely add and edit columns and be able to store different types of data, so rigid systems purpose-built for medical trial data are not recommended.
-In this exercise we will stick to the simplest, but flexible platform - Excel spreadsheet.
+In this exercise we will stick to the simplest, but flexible platform - an Excel spreadsheet.
 
@@
 **QUESTION 9:**   
-Why we should record meta-data when extracting fdata for meta-analysis?
+Why should we record meta-data when extracting data for a meta-analysis?
 
 **Note:** In your actual extraction sheet/form, try to group related columns together.
-You can also color-code them in Excel or put on the same tab/section in a visual database interface, if used.
-Your extraction form (spreadsheet) should be tested on several papers (pilot extractions) to make sure its easy to use and captures necessary information.
+You can also color-code them in Excel or put them on the same tab/section in a visual database interface, if used.
+Your extraction form (spreadsheet) should be tested on several papers (pilot extractions) to make sure it is easy to use and captures necessary information.
 
@@
 When numerical data is reported in the text or tables, it is easy to get exact values for your extraction columns.
-However, often data is reported in the figures?
+However, often data is reported in the figures.
 Can we extract such data?
@@
 **TASK 1**:  
-We will practice data extraction on few of the Terminal investment papers. We will use R package *metaDigitise*.   
+We will practice data extraction on few of the Terminal investment papers. We will use the R package *metaDigitise*.
+
 Detailed package description and tutorial are available at: https://cran.r-project.org/web/packages/metaDigitise/vignettes/metaDigitise.html or
@@
 Figure 2a in the paper by Sadd et al.
-2006: Modulation of sexual signalling by immune challenged male mealworm beetles (Tenebrio molitor, L.): evidence for terminal investment and dishonesty (**Sadd2006.pdf**)
+2006: Modulation of sexual signalling by immune challenged male mealworm beetles (*Tenebrio molitor*, L.): evidence for terminal investment and dishonesty (**Sadd2006.pdf**)
 
-You will need to open the pdf files and take the snapshots of the relevant figures (for MacOS the easiest is via Command+Shift+4 and then select the screen fragment; for Windows see https://www.windowscentral.com/how-take-windows-10-screenshot-7-best-techniques).
+You will need to open the pdf files and take the snapshots of the relevant figures (for MacOS the keyboard shortcut is easiest: press Command+Shift+4 and then select the screen fragment; for Windows see https://www.windowscentral.com/how-take-windows-10-screenshot-7-best-techniques).
 Make sure you capture the axes and axes labels.
-Also, enlarge these figures to fill the screen before taking the snapshot to get higher resolution image.
+Also, enlarge these figures to fill the screen before taking the snapshot to get a higher resolution image.
 
@@
 
-Tell metaDigitise() what you would like to do by selecting option  (Process new images - enter "1").
+Tell *metaDigitise()* you would like to process new images by selecting option "1".
 
@@
 
-First plot should be loaded at this stage. We do not need to rotate the image, so enter "c".
+The first plot (**001_Weil_2006_Fig2a.png**) should be loaded at this stage. We do not need to rotate the image, so enter "c".
 
@@
 
-You get asked for the data needed to calibrate y-axis. Click on 200 and 900, then enter these values as prompted and complete calibration.
+First you are asked to calibrate the y-axis.
+To do this as precisely as possible, enlarge the plot viewer before clicking on the lowest and highest marked values (in this case, 200 and 900).
+Enter these values as prompted.
+The program will then show you your calibration and ask if you want to re-calibrate; if it looks accurate enough, click “n” to proceed, otherwise click “y” to try again.
 
@@
 Since there is no such info in the plot or under the plot, search the full text for the sample sizes (hint: look in "Statistical anlaysis").
-Enter identifier for the control group ("saline") and the sample size.
+Ideally, to do it efficiently you would already have extracted this and other available information into your spreadsheet, then saved the figures, and then come back at the end to process all the figures at once.
+This way you have everything at hand when you do data extraction from the figures.
 
-Since we have multiple data points for each group of animals, we have to decide whether extract all of them or just one (or some subset) - your decision should be biologically-informed and ideally a priori.
+Once you have the information on the group names and sample sizes, enter an identifier for the control group ("saline") and sample size for this group.
+
+Since we have multiple data points for each group of animals, we have to decide whether to extract all of them or just one (or some subset) - your decision should be biologically-informed and ideally a priori.
 For this exercise we will extract just the last one (at 6 weeks).
@@
 
-Now you should be able to extract by yourself plot 002_Sadd_2006_Fig2a.png! (hint: extract "treatment control" and "treatment" groups).
+Now you should be able to extractdata from **plot 002_Sadd_2006_Fig2a.png**! (hint: extract "treatment control" and "treatment" groups).
 
-Once you get "Congratulations" message, you can look at the extracted data. metaDigitise() stores and sorts all the summary data for each of the figures in the "data"" object:
+Once you get a "Congratulations" message, you can look at the extracted data. *metaDigitise()* stores and sorts all the summary data for each of the figures in the "data"" object:
 
@@
 
-You can add more figures in the directory and add more data to this data table.
-Or you can save it and load back later to continue adding more data or merge with other data frames (see "Re-importing previously digitised data and accessing raw data" in the metaDigitise tutorial at https://cran.r-project.org/web/packages/metaDigitise/vignettes/metaDigitise.html).
+You can add more figures in the directory and add more data to this data table later, or save it and load back later to continue adding more data or merge with other data frames (see "Re-importing previously digitised data and accessing raw data" in the *metaDigitise* tutorial at https://cran.r-project.org/web/packages/metaDigitise/vignettes/metaDigitise.html).
 
@@
 
-As you see, metaDigitise guides you gently through the all steps of data extraction from figures.
+As you see, *metaDigitise* guides you gently through the all steps of data extraction from figures.
 It also nicely organises and stores the data for you.
@@
 **QUESTION 13:**   
-Why we need to change our data frame (and how)?
+Why do we need to change our data frame (and how)?
 
 **TASK 2**:
-Since we have only 2 data points per file, its easy to reshape the data frame in R.
+Since we have only 2 data points per file, it is easy to reshape the data frame in R.
 
@@
 
-Now the columns appended with .x contain control groups data values, and these with .y are for the treatment groups (you can rename the columns to make it easier to see).
+Now the columns appended with .x contain control groups data values, and these with .y are for the treatment groups (you can rename the columns to make it easier to see, e.g.
+using the *rename* function https://dplyr.tidyverse.org/reference/select.html).
 
 **TASK 3**:  
-We will now use *escalc* function from *metafor* package to calculate effect sizes.
-The function allows to calculate various effect sizes or outcome measures (and the corresponding sampling variances) that are commonly used in meta-analyses, including two effect sizes that are relevant to our data and question:
- * "SMD" stands for standarised mean difference, i.e. Hedges' g (potentially also "SMDH" for heteroscedastic variances).   
+We will now use the *escalc* function from *metafor* package to calculate effect sizes.
+This function allows to calculate various effect sizes or outcome measures (and the corresponding sampling variances) that are commonly used in meta-analyses, including two effect sizes that are relevant to our data and question:
+ * "SMD" stands for standarised mean difference, i.e. Hedges' *g* (potentially also "SMDH" for heteroscedastic variances).   
  * "ROM" for the log transformed ratio of means (Hedges et al., 1999; Lajeunesse, 2011), also named lnRR.   
@@
 If we cannot get means and standard errors (or SD) from the text, tables and figures (or already-calculated effect sizes, which is still rare), we can look for test (inferential) statistics.
-Test statistics that are usable include t-values form t-tests, F-values, Chi-square values, finally, p-values.
+Test statistics that are usable include t-values form t-tests, F-values, Chi-square values, and p-values.
 These values from simple statistical tests (like independent or paired t-tests) are easily convertible into SMD (d, g) or r (correlation coefficient).
-These conversions and some others can be calculated in R if you know the formulas (including conversion from SMD to correlation coefficient), but for the exercise purpose we will do this web-based Practical Meta-Analysis Effect Size Calculator:
+These conversions and some others can be calculated in R if you know the formulas (including conversion from SMD to correlation coefficient), but for the exercise purpose we will use this web-based Practical Meta-Analysis Effect Size Calculator:
 https://www.campbellcollaboration.org/research-resources/research-for-resources/effect-size-calculator.html
@@
 
-Open the above webpage. Main menu has an expandable list of types of effect sizes, and statistical tests you can use to calculate these effect sizes (only *d*, *r*, *OR* and *RR*).
+Open the above webpage. The main menu has an expandable list of types of effect sizes, and statistical tests you can use to calculate these effect sizes (only *d*, *r*, *OR* and *RR*).
 
@@
 Weil 2006 reports in the results: "At the end of the experiment saline-injected animals had smaller testes (F 1,27 = 4.6875, p<0.05, figure 2b)" and in the methods it is noted these are calculated using ANOVA.
-Use online ES calculator sheet selecting "Standarized Mean difference (d)" and "F-test, 2 groups" to calculate *d* from one-way ANOVA (you will need F-value and sample sizes of treatment and control groups).
+Use the web-based effect size calculator sheet selecting "Standarized Mean difference (d)" and "F-test, 2 groups" to calculate *d* from one-way ANOVA (you will need F-value and sample sizes of treatment and control groups).
 
@@
 **TASK 5**:   
-Using web-based effect size calculator, estimate a few other effect sizes using test (inferential) statistics reported in Sadd2006.
+Using the web-based effect size calculator, estimate a few other effect sizes using test (inferential) statistics reported in Sadd2006.
 
-In practice, we would preer to do it in R, rather than in aan online up or any other non-scriptable software that cannot socument our calculations.
-Luckily, the R package *compute.es* allows not only compute various effect sizes from descriptive statistics, but also from inferential statistics.
-r function inputs and names are as follows:
- * ANCOVA F-test: a.fes()
- * ANCOVA means: a.mes()
- * ANCOVA means (pooled sd): a.mes2()
- * ANCOVA p-value: a.pes()
- * ANCOVA t-test: a.tes()
- * Chi-squared (1 df): chies()
- * Correlation: res()
- * d-statistic: des()
- * Failure group (binary): failes()
- * F-test: fes()
- * Log odds ratio: lores()
- * Means: mes()
- * Means (pooled sd): mes2()
- * Proportions (binary): propes()
- * p-value: pes()
- * t-test: tes()
+In practice, we would preer to do this in R, rather than in an online app or any other non-scriptable software that cannot socument our calculations.
+Luckily, the R package *compute.es* allows us to not only compute various effect sizes from descriptive statistics, but also from inferential statistics.
+
+The *compute.es* function inputs and names are as follows:   
+ * ANCOVA F-test: "a.fes()""   
+ * ANCOVA means: "a.mes()"   
+ * ANCOVA means (pooled sd): "a.mes2()"   
+ * ANCOVA p-value: "a.pes()"    
+ * ANCOVA t-test: "a.tes()"   
+ * Chi-squared (1 df): "chies()"   
+ * Correlation: "res()"    
+ * d-statistic: "des()"    
+ * Failure group (binary): "failes()"    
+ * F-test: "fes()"   
+ * Log odds ratio: "lores()"    
+ * Means: "mes()"    
+ * Means (pooled sd): "mes2()"   
+ * Proportions (binary): "propes()"    
+ * p-value: "pes()""    
+ * t-test: "tes()""    
 To learn more see the package vignette or https://www.rdocumentation.org/packages/compute.es/versions/0.2-4/topics/compute.es-package
 
-As an example, we will use F value from **TASK 4**, note that functions from the *compute.es* packages automatically calcualte all available effect sizes for the given imput and if you run it on a data frame it will generate a new data.frame with several effect size estimates.
+As an example, we will use the F-value from **TASK 4**.
+Note that functions from the *compute.es* packages automatically calculate all available effect sizes for the given input and if you run it on a data frame it will generate a new data.frame with several effect size estimates.
 Try:
@@
 
-#for vectors of values (lieke in a data frame):
+#for vectors of values (like in a data frame):
 results_ES <- a.fes(f = c(4.6875, 4.6875), n.1 = c(28, 10), n.2 = c(10, 10), R = c(0, 0), q = c(0, 0))
@@
 **QUESTION 16:**    
-Is the sign of effect size important? Imagine we use *d*, which represents mean_treatment - mean_control group trait values. How we would interpret *d* = 0.5 and how we would interpret *d* = -0.5?
+Is the sign of the effect size important? Imagine we use *d*, which represents mean_treatment - mean_control group trait values. How we would interpret *d* = 0.5 and how we would interpret *d* = -0.5?
 
 **QUESTION 17:**    
-If we have different traits in the data set (lets say number of eggs laid and probability of nest abandonment) what is the problem with the signs of effect sizes for these traits?
+If we have different traits in the data set (let's say number of eggs laid and probability of nest abandonment) what is the problem with the signs of effect sizes for these traits?
 What should we do about this?
@@
 **QUESTION 18:**   
-As you hopefully remember, PRISMA checklist is a document that helps making sure that you reported the key information how a systematic review (including meta-analysis) was done.    
+As you hopefully remember, the PRISMA checklist is a document that helps ensure you report the key information on how a systematic review (including meta-analysis) was done.
+
 Have a look again at the PRISMA checklist (http://prisma-statement.org/documents/PRISMA%202009%20checklist.pdf).
@@
 
-**Note:** this PRISMA was designed for medical systematic reviews, and some items are not very relevant or useful. Also, it is due to be updated.   
+**Note:** worth repeating again, this PRISMA was designed for medical systematic reviews, and some items are not very relevant or useful. Also, it is due to be updated.   
 <br>
@@
 **QUESTION 19:**   
-Although not mentioned in the original PRISMA checklist, it is important to share the collected data, meta-data and analysis code? Think of at least 3 reasons for doing so.
+Although not mentioned in the original PRISMA checklist, it is important to share the collected data, meta-data and analysis code. Think of at least 3 reasons for doing so.
 
@@
 
-*We hope you enjoyed this exercise! Remeber, we just roughly covered the basics and there is much more to learn!*
+*We hope you enjoyed this exercise! Remember, we just roughly covered the basics and there is much more to learn!*

### 2019-02-06 Data_extractions.Rmd
typo fixed

@@
 
-You will need to open the pdf files and take the snapshots of the relevant figures (for MacOS the keyboard shortcut is easiest: press Command+Shift+4 and then select the screen fragment; for Windows see https://www.windowscentral.com/how-take-windows-10-screenshot-7-best-techniques).
+You will need to open the pdf files and take the snapshots of the relevant figures (for Mac OS the keyboard shortcut is easiest: press Command+Shift+4 and then select the screen fragment; for Windows see https://www.windowscentral.com/how-take-windows-10-screenshot-7-best-techniques).
 Make sure you capture the axes and axes labels.
@@
 Its good to enter sample sizes at this point.
-Since there is no such info in the plot or under the plot, search the full text for the sample sizes (hint: look in "Statistical anlaysis").
+Since there is no such info in the plot or under the plot, search the full text for the sample sizes (hint: look in "Statistical analysis").
 Ideally, to do it efficiently you would already have extracted this and other available information into your spreadsheet, then saved the figures, and then come back at the end to process all the figures at once.
@@
 
-Now you should be able to extractdata from **plot 002_Sadd_2006_Fig2a.png**! (hint: extract "treatment control" and "treatment" groups).
+Now you should be able to extract data from **plot 002_Sadd_2006_Fig2a.png**! (hint: extract "treatment control" and "treatment" groups).
 
@@
 
-In practice, we would preer to do this in R, rather than in an online app or any other non-scriptable software that cannot socument our calculations.
+In practice, we would prefer to do this in R, rather than in an online app or any other non-scriptable software that cannot document our calculations.
 Luckily, the R package *compute.es* allows us to not only compute various effect sizes from descriptive statistics, but also from inferential statistics.

### 2019-02-06 Trees_making.Rmd
Spelling corrections

@@
 A few things to note:  
-1. We can use a tree with or without branch lengths (divergence measures).
+1. We can use a tree with or without branch lengths (divergence measures) although we do need branch lengths for analysis (often making certain assumptions).
 
@@
 a super-tree).
-Currently, the easiest way to build a custom phylogenetic tree usable for meta-analysis is by using an R package called *rotl*, which accesses a synthetic super-tree from Open Tree of Life database (https://opentree.wikispaces.com/).
+Currently, the easiest way to build a custom phylogenetic tree usable for meta-analysis is by using an R package called *rotl*, which accesses a synthetic super-tree from Open Tree of Life database (https://	opentreeoflife.org).
 We will use this approach in our examples below.
@@
 
-Run the below code chanks IF you have not installed these packages earlier (it may take 15-20 minutes to install all).
+Run the below code chunks IF you have not installed these packages earlier (it may take 15-20 minutes to install all).
 
@@
 We will try to create a simple tree using 10 commonly used taxa (lab model species) from across the tree of life.  
-Rotl is an R package (https://peerj.com/preprints/1471/) allowing us to access synthetic phylogenetic tree available at Open Tree of Life database (https://opentree.wikispaces.com/).
+Rotl is an R package (https://peerj.com/preprints/1471/) allowing us to access synthetic phylogenetic tree available at Open Tree of Life database (https://	opentreeoflife.org/).
 
@@
 Note that the species names have the Open Tree identifiers appended to them and *E.coli* was replaced with "mrcaott616ott617", which is non-taxon node within Bacteria.
-Given we dont have any other bacteria in the data set, and the position on the tree is correct, that's fine, we just need to rename the tree tip back to "Escherichia coli".
+Given we don't have any other bacteria in the data set, and the position on the tree is correct, that's fine, we just need to rename the tree tip back to "Escherichia coli".
 
@@
 Note that genus and species name are stored in separate columns.
-Also, smae species name can appear several times in the data set, but since we need a species list, we need each name to apper only once (i.e.
+Also, same species name can appear several times in the data set, but since we need a species list, we need each name to appear only once (i.e.
 unique names), and capitalisation to be consistent (letter case matters when matching strings).
@@
 So, we run into a problem with one of the Populus names - need to fix this by using an alternative name.
-Populus euramericana, is a naturally occurring hybrid of Populus deltoides, so we can replace it with this name for the search (given its colesely related and we dont have this species on our species list).
+Populus euramericana, is a naturally occurring hybrid of Populus deltoides, so we can replace it with this name for the search (given its closely related and we don't have this species on our species list).
 

### 2019-02-06 Papers_search.Rmd
typos fixed

@@
 In this exercise, we will use a few basic tools to illustrate how to perform the first few steps of a meta-analysis: refining the question, searching the literature, and screening the found studies (these are the same steps as in Systematic Reviews, so the tools are the same and can be usually described as systematic literature/evidence reviewing tools).
-Note that some R packages (and many other online/software tools) are available (and more are bieng developed) that can help with some of the tasks during this exercise, but there is no single "perfect" tool for the whole process usable for all disciplines and types of questions (although some attempts have been made).
+Note that some R packages (and many other online/software tools) are available (and more are being developed) that can help with some of the tasks during this exercise, but there is no single "perfect" tool for the whole process usable for all disciplines and types of questions (although some attempts have been made).
 We simply do not have enough time to introduce more tools and the main purpose of this exercise is to familiarize you with the basic principles/issues that apply when you use any tools.
@@
 **QUESTION 0:**  
-What is your initial feeling about the meta-analytical question that can be derived from the text fragment cited above?
+What is your initial feeling about the meta-analytic question that can be derived from the text fragment cited above?
 
@@
 (e.g.
-descriptive statistics, and which ones, or also inferential model statsistics, and which ones?
+descriptive statistics, and which ones, or also inferential statistics, and which ones?
 This will be linked to the effect size you are planning to use!)
@@
 However, in practice, if a study looks very relevant but some information is ambiguous or missing, you may decide to contact the authors to get some extra information or data.
-in such case, you can provisionary include such papers and make final decision once you get, or not, the replies.
+in such case, you can provisionally include such papers and make final decision once you get, or not, the replies.
 
@@
 The reason is that we want to have as many relevant data points as possible to increase the power of statistical analyses.
-The other reason is that we make effort to capture obscure studies that potentially are more likely to report non-significant results, and ttherefore help reduce publication bias in our data set.
+The other reason is that we make effort to capture obscure studies that potentially are more likely to report non-significant results, and therefore help reduce publication bias in our data set.
 Search strategy should include using at least two wide-coverage/interdisciplinary databases (we recommend Scopus and Web of Science), and a few additional sources of literature such as forward and backward reference searching (looking at cited and citing papers), publication lists of the key authors or labs, searching for unpublished studies and data sets (we will not cover these here).
@@
 Go to "All reviews" and check if other people finished their screening, if applicable.
-After that you can lift the "blind" mode (when colleborators cannot see each others decisions).
+After that you can lift the "blind" mode (when collaborators cannot see each others decisions).
 For collaborative screens Rayyan will now provide filters to display all consensus included, consensus excluded and records with conflicting decisions.
@@
 Export the records with final consensus decisions, for documentation, into a .csv file.
-You can filter the records into “Included” and just export those, also you can export them as a .bib or .ris file into a reference manager where you will collect full texts for the enxt stage of the screening.
+You can filter the records into “Included” and just export those, also you can export them as a .bib or .ris file into a reference manager where you will collect full texts for the nest stage of the screening.
 

### 2019-02-06 Papers_search_solutions.Rmd
typos fixed

@@
 In our project it can be Cohen's d (Hedges g) or lnRR, which are typically calculated from mean, SD or SE and N (sample sizes).
-However, they can be also inferred from some of the test statistics, or even proportions, if means and other descriptive statistics are not available.
+However, Cohen's d (Hedges g) can be also inferred from some of the test statistics, or even proportions, if means and other descriptive statistics are not available.
+But one cannot get lnRR from inhferential statistics.
 Other considerations include what to do if some data is missing or unclear.
@@
 It is often not explicitly stated, but without the full text it is impossible to extract all the necessary information about the study and the data. 
-It is a good practice to keep this in mind and use as one of your potential inclusion/exclusion critera.
+It is a good practice to keep this in mind and use as one of your potential inclusion/exclusion criteria.
 
 **ANSWER 16:**   
-We think 4 should have been excluded: Simmons2009 and Clancy 2013 as human studies, Gershman2010 and Gonzales-Tokman 2013 as not likely to have any reproduction-related measurements.
+We think 4 should have been excluded: Simmons2009 and Clancy2013 as human studies, Gershman2010 and Gonzales-Tokman2013 as not likely to have any reproduction-related measurements.
 

### 2019-02-07 R/Data_extractions_solutions.Rmd
reorganised

--- 2019-02-02 Data_extractions_solutions.Rmd
@@
 **ANSWER 4:**    
- 1. Non-informative column names (which group should be nr 1 ?)   
+ 1. Non-informative column names (which group should be N 1 ?)  and column names should be one string (word) 
  2. Second study has non-numerical values in the columns that should be numerical (NA is good for missing values, but should be used consistently)    
@@
 These are not independent data points, and this is captured by CohortID.
-For different study question this could be an unique identifier of various study sites or populations that need tp be distinguished within and between studies.
+For different study question this could be an unique identifier of various study sites or populations that need to be distinguished within and between studies.
 
@@
 the measurement units, how conversions were made, any decisions and assumptions that had to be made in general for a given column of data"; e.g.
-whether sex was coded as "F" and "M" and "B", and what these values mean, jhow missing values were coded, and anything else that would allow you or somebody to extract and code the same data).
+whether sex was coded as "F" and "M" and "B", and what these values mean, how missing values were coded, and anything else that would allow you or somebody to extract and code the same data).
 Meta-data is an essential part of the documentation of your meta-analysis and it is extremely useful when you or somebody else checks the data or tries to re-analyse or re-use it.

### 2019-02-07 R/Trees_making_solutions.Rmd
reorganised

--- 2019-01-28 Trees_making_solutions.Rmd
@@
 #if Treebase does not work use: Quercuss_tree <- read.nexus(file = "T113123.nex") #read Nexus tree file
-
+intersect(as.character(Quercuss_tree$tip.label), Quercus3sp) # all 3 are present and matching
 # plot(Quercuss_tree, cex=.8, label.offset =.1, no.margin = TRUE) #a big tree of all sorts of trees - needs good trimming!
@@
 #check the overlap between species name from the data fila nad the tree file
-intersect(as.character(Quercus3sp_tree$tip.label), Quercus3sp) # all matching, but we changed the species name in teh list, so it will no longer match the original dataset
+intersect(as.character(Quercus3sp_tree$tip.label), Quercus3sp) # all matching, but we changed the species name in the list, so it will no longer match the original dataset
 Quercus3sp_tree$tip.label <- gsub("montana", "prinus", Quercus3sp_tree$tip.label) #replace the tip label back with our species name
@@
 tree <- read.tree(file="tree_curtis1998.tre") #run if you need to read it in
-tree_pruned <- drop.tip(tree, c("Quercus prinus", "Quercus alba")) #prune 3 species
+tree_pruned <- drop.tip(tree, c("Quercus_prinus", "Quercus_alba")) #prune 3 species
 tree_grafted <- bind.tree(tree_pruned, Quercus3sp_tree, where=which(tree_pruned$tip.label=="Quercus_rubra")) #graft 3 species to the remaining Acer species

### 2019-02-07 R/Papers_search.Rmd
updated new subdirectories for the used files

--- 2019-02-06 Papers_search.Rmd
@@
 ```{r TI_book_fragment, echo=FALSE, fig.cap=" ", out.width='50%'}
-knitr::include_graphics("./images/TI_book_fragment.png")
+knitr::include_graphics("../images/TI_book_fragment.png")
 ```
@@
 These question refinements are leading us gently into deciding which studies to use in a meta-analysis.
-However, before we move onto inclusion criteria, it is wise to gather more intelligence (Unless you are already an expert on the topic).
+However, before we move onto inclusion criteria, it is wise to gather more intelligence (unless you are already an expert on the topic).
 
@@
 ```{r WoS_adv_search, echo=FALSE, fig.cap=" ", out.width='100%'}
-knitr::include_graphics("./images/WoS_adv_search.png")
+knitr::include_graphics("../images/WoS_adv_search.png")
 ```
@@
  3. Enter project title, e.g. "Terminal Investment"" and description, e.g. "test"."   
- 4. Select files -> upload *Scopus10.ris* and press Continue.     
+ 4. Select files -> upload *Scopus10.ris* from the *data* sub-folder and press Continue.     
 It may take several minutes for larger files and slower connections, but eventually the uploaded references should appear in the left panel under "Search methods" (you can upload more files with references via "Add new" link).
@@
  5.
-In the "Keywords for include"" you can add (via "Add new" link) following terms (or at least a few of these): challenge, challenged, injected, injection, infected, infection, PHA, SRBC, vaccine (these terms will be highlighted green in the titles and abstracts to help them find more easily; Rayyan tries to populate this list by itself - but its algorithm is medical-centered so you should adjust this list for your projects).
+In the "Keywords for include" you can add (via "Add new" link) following terms (or at least a few of these): challenge, challenged, injected, injection, infected, infection, PHA, SRBC, vaccine (these terms will be highlighted green in the titles and abstracts to help them find more easily; Rayyan tries to populate this list by itself - but its algorithm is medical-centered so you should adjust this list for your projects).
  6.
@@
 It is time to do the full-text screening for the papers that passed title and abstract screening.
-There are some full-text papers that passed the initial screening in the directory - assess **Nielsen2012** and **Ardia2005** using our full-text screening decision tree.
-Hint: do not read the whole paper - just try to skim it and look for the key information.
+There are some full-text papers that passed the initial screening in the directory - assess **Nielsen2012** and **Ardia2005** (they are in "pdf" sub-folder) using our full-text screening decision tree.
+(Tip: do not read the whole paper - just try to skim it and look for the key information.)
 

### 2019-02-07 R/Papers_search_solutions.Rmd
added some bold and italic formatting, line breaks

--- 2019-02-06 Papers_search_solutions.Rmd
@@
 It depends on the effect size we are planning to use.
-In our project it can be Cohen's d (Hedges g) or lnRR, which are typically calculated from mean, SD or SE and N (sample sizes).
-However, Cohen's d (Hedges g) can be also inferred from some of the test statistics, or even proportions, if means and other descriptive statistics are not available.
-But one cannot get lnRR from inhferential statistics.
+In our project it can be Cohen's *d* (Hedges *g*) or *lnRR*, which are typically calculated from mean, SD or SE and N (sample sizes).
+However, Cohen's *d* (Hedges *g*) can be also inferred from some of the test statistics, or even proportions, if means and other descriptive statistics are not available.
+But one cannot get *lnRR* from inhferential statistics.
 Other considerations include what to do if some data is missing or unclear.
@@
 **ANSWER 16:**   
-We think 4 should have been excluded: Simmons2009 and Clancy2013 as human studies, Gershman2010 and Gonzales-Tokman2013 as not likely to have any reproduction-related measurements.
+We think 4 should have been excluded:    
+**Simmons2009** and **Clancy2013** as human studies, but    
+**Gershman2010** and **Gonzales-Tokman2013** as not likely to have any reproduction-related measurements.
 
 **ANSWER 17:**    
-Nielsen2012 - include, Ardia2005 - exclude due to the lack of good control group.
+**Nielsen2012** - include, **Ardia2005** - exclude due to the lack of good control group.
 

### 2019-02-07 R/Data_extractions.Rmd
updated file sub-directories and bold/italic formatting

--- 2019-02-06 Data_extractions.Rmd
@@
 
-Let's try to extract data from:   
+Let's try to extract data from (all pdf files are located in the "pdf" sub-folder):   
  1. Figure 2a in the paper by Weil et al. 2006: Immune challenge retards seasonal reproductive regression in rodents: evidence for terminal investment (**Weil2006.pdf**)   
@@
 ```{r save metadigitise, eval=FALSE, message = FALSE}
-write.csv(data, file = "filename.csv")
+write.csv(data, file = "filename.csv") #adjust the file name and path as needed
 ```
@@
 Test statistics that are usable include t-values form t-tests, F-values, Chi-square values, and p-values.
-These values from simple statistical tests (like independent or paired t-tests) are easily convertible into SMD (d, g) or r (correlation coefficient).
+These values from simple statistical tests (like independent or paired t-tests) are easily convertible into SMD (*d*, *g*) or r (correlation coefficient).
 These conversions and some others can be calculated in R if you know the formulas (including conversion from SMD to correlation coefficient), but for the exercise purpose we will use this web-based Practical Meta-Analysis Effect Size Calculator:
@@
 ```{r webEScalcualtor, echo=FALSE, fig.cap=" ", out.width='100%'}
-knitr::include_graphics("./images/webEScalcualtor.png")
+knitr::include_graphics("../images/webEScalcualtor.png")
 ```
@@
 **TASK 5**:   
-Using the web-based effect size calculator, estimate a few other effect sizes using test (inferential) statistics reported in Sadd2006.
+Using the web-based effect size calculator, estimate a few other effect sizes using test (inferential) statistics reported in **Sadd2006.pdf**.
 
@@
  * t-test: "tes()""
+
 To learn more see the package vignette or https://www.rdocumentation.org/packages/compute.es/versions/0.2-4/topics/compute.es-package

### 2019-02-07 R/Data_extractions_solutions.Rmd
minor text formatting fixes

@@
 **ANSWER 11:**    
-It is often easier and quicker to group similar studies together, Especially grouping the studies by study/system and research group helps find overlapping studies (e.g.
+It is often easier and quicker to group similar studies together.
+Especially, grouping the studies by study/system and research group helps find overlapping studies (e.g.
 where the same experiment was reported in separate publications, but focusing on different aspects) or collect more details on how studies were done.
@@
 The sign represents direction of the effect on a given trait.
-d = 0.5 means that the trait value in the treatment group is moderately larger than the trait value in the control group, d = -0.5 would mean that the trait value in the treatment group is moderately smaller than the trait value in the control group.
+*d* = 0.5 means that the trait value in the treatment group is moderately larger than the trait value in the control group, *d* = -0.5 would mean that the trait value in the treatment group is moderately smaller than the trait value in the control group.
 

### 2019-02-07 R/Trees_making.Rmd
updated file subdirectories and minor formatting fixes

--- 2019-02-06 Trees_making.Rmd
@@
 ```{r list birds load} 
-birds <- unique(read.csv("bird_list.csv")$species_name) #load our list of bird species and make the names unique
-birds_stree <- read.tree("Ericson.tre") #load bird supertree
+birds <- unique(read.csv("../data/bird_list.csv")$species_name) #load our list of bird species and make the names unique
+birds_stree <- read.tree("../data/Ericson.tre") #load bird supertree
 birds_stree #9993 tips = species
@@
 
-write.tree(tree, file="tree_curtis1998.tre") #save the tree in Newick format, if needed
-# read.tree(file="tree_curtis1998.tre") #if you need to read it
+# write.tree(tree, file="../data/tree_curtis1998.tre") #save the tree in Newick format, if needed
+# read.tree(file="../data/tree_curtis1998.tre") #if you need to read it
 ```
@@
 # is.ultrametric(tree) #the tree has no branch lengths, so cant be checked
-write.tree(tree, file="tree_curtis1998.tre") #save the tree
-# tree <- read.tree(file="tree_curtis1998.tre") #if you need to read it
+#write.tree(tree, file="../data/tree_curtis1998.tre") #save the tree
+# tree <- read.tree(file="../data/tree_curtis1998.tre") #if you need to read it
 ```
@@
 is.binary.tree(tree_random) #TRUE
-write.tree(tree_random, file="tree_curtis1998_random.tre") #save the tree
-# tree_random <- read.tree(file="tree_curtis1998_random.tre") #if you need to read it
+#write.tree(tree_random, file="../data/tree_curtis1998_random.tre") #save the tree
+# tree_random <- read.tree(file="../data/tree_curtis1998_random.tre") #if you need to read it
 ```
@@
 ```{r dwonload one tree from file, eval = FALSE}
-Acers_tree <- read.nexus(file = "T418.nex") #read Nexus tree file
+Acers_tree <- read.nexus(file = "../data/T418.nex") #read Nexus tree file
 ```

### 2019-02-08 R/Data_extractions.Rmd
fixed brackets and package installation code, re-knitted

@@
 #install.packages("metaDigitise")
+#install.packages("tidyverse") #in case you have not installed this earlier
+#install.packages("compute.es") #in case you have not installed this earlier
 ```
@@
 **TASK 4**:   
-Weil 2006 reports in the results: "At the end of the experiment saline-injected animals had smaller testes (F 1,27 = 4.6875, p<0.05, figure 2b)" and in the methods it is noted these are calculated using ANOVA.
+Weil (2006) reports in the results: "At the end of the experiment saline-injected animals had smaller testes (F 1,27 = 4.6875, p<0.05, figure 2b)" and in the methods it is noted these are calculated using ANOVA.
 Use the web-based effect size calculator sheet selecting "Standarized Mean difference (d)" and "F-test, 2 groups" to calculate *d* from one-way ANOVA (you will need F-value and sample sizes of treatment and control groups).

### 2021-08-09 R/grey_lit.Rmd
add grey lit part

---
title: "grey_lit"
author: "ML"
date: "1/24/2019"
output: html_document
---

```{r setup, cache = F, echo=FALSE, results = FALSE}
knitr::opts_chunk$set(error = TRUE) #allow some execution errors for demonstration purposes
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, warning = FALSE)
sessionInfo()
```

# A FEW WORDS ABOUT GREY LITERATURE

Prepared by Malgorzata (Losia) Lagisz, BEES, UNSW, AU, losialagisz@gmail.com

********************************************************************************

### What "grey literature" is?

The officel definiton from the Cochrane Handbook for Systematic Reviews of Interventions is:   
*"...literature that is not formally published in sources such as books or journal articles."*

So, effectively, everything else that has been written could be classfied as grey literature (including this text). 
Obviously, for a meta-analysis we are only interested in research-related literature, and more specfically literature related to a given meta-analystic topic. 
But, this still includes a lot: graduate dissertations, conference abstracts and proceedings, research reports, blogs, news stories...  (for the full list see http://www.greynet.org/greysourceindex/documenttypes.html).

Notably, peer-review is not a defining feature of grey literature - some of the literature types mentioned above can be peer-reviewed, for example MsC and PhD theses, some of the reports.

*Note:* most grey literature-related guidelines are written for medical sciences, thsi one is tailored for ecology and evolution, so I have skipped less relevant items...

********************************************************************************

### What "grey literature" is NOT?

 * It is not the same as books and academic journal papers that are published in languages other than English.   
 * It is not equivalent to the papers published in obscure journals, not indexed in the main databases.   
 * It is not equivalent to old jouranl papers and books (pre-digital era) that are less likely to have proper electronic records and be indexed online.

********************************************************************************

### Why we need to include grey literature?

We want to include data from grey literature in our meta-analyses in order to reduce publication bias... (studies with small or non significant effects being more liely to remain unpublished).  We cannot completely get rid of publication bias, since we can never be sure that we actually collected all available data that is relevant, but we can do our best to include as much unpublished data as possible in a reasonable timeframe.

********************************************************************************

### What to search for?

As just noted above, we need to do it in reasonable time (weeks, months). So the best strategy is to focus on the sources of grey literature that have the highest probability of reporting actual data we are after, also it needs to be reported with sufficient level of detail. The best and quite reliable source of such data is from graduate dissertations. Doctoral and master theses are more likely than actual papers based on them (if ever published) to provide detailed descriptions of the methods and results, including non-significant ones.  We should also broaden the definition of grey literature to include unpublished data (including meta-data), since we this is our effective target. It also implies additional ways for searching.

********************************************************************************

### How to search for grey literture?

Databases of mixed-type grey-literature:   
 * OpenGrey (Grey Literature in Europe): http://www.opengrey.eu/   
 * Trove (Australian; apart of theses it hes other types of grey literature - you have to search for theses from the Books -> Australian content category!) : https://trove.nla.gov.au/   
 * Informit: https://search.informit.com.au/

Databases (including databases aggregation services) of graduate dissertations:   
 * ProQuest Dissertation & Theses Global (PQDT Global): http://www.proquest.com/products-services/cpi-set-c.html  
 * WorldCatDissertations: http://www.worldcat.org   
 * BSCO Open Dissertations: https://www.ebsco.com/products/research-databases/ebsco-open-dissertations

Databases (including databases aggregation services) of datasets:  
 * Google Data Search: https://toolbox.google.com/datasetsearch   
 * Elsevier Data Search: https://datasearch.elsevier.com/#/  
 * Mendeley Data: https://data.mendeley.com/

Full list of web-based resources in grey literature is available here: http://www.greynet.org/greysourceindex.html.

Note that many of the above resources only use a simple basic search query, so run your searches using kjust the key phrases for your topic.

Using genrealistic search engines:  
 * Google / Google Scholar (note, it includes grey literature and can search in full texts). Our searches are usually Google Personalized Searching since 2009 - our search resulst are influenced by stored search histories and influenced by geographical locations. Try using country-specific version of Google. You can also limit the search to pages from a given country (under the Web tab) or Google Advance Search with Region Limit (other helpful limit: file type PDF).   
 * DuckDuckGo: https://duckduckgo.com . Search results are not personalized.   
 * Startpage: https://startpage.com . Searches via Google, but with privacy settings not allowing to store pervious search histories.

 Last, but not least, contact the key research labs and people working on the given topic and ask thatm for any unpublished data thay may have (or published papers that you jave missed in your earlier searches).
